### 2016-03-04 ms.Rmd
Move over to R markdown to allow embeded cod in ms. Will mess up the OSX/Unix stuff

---
title: "The `zoon` R package for reproducible and shareable species distribution modelling"
author: "Nick Golding, Tom August, Tim C.D. Lucas, Emiel Van Loon & Greg McInerny"
date: "4 March 2016"
output: 
  pdf_document:
    fig_caption: yes
---

### Abstract

1. The diverse array of software and methodological research available for species distribution modelling (SDM) hinders independent evaluation of new methods and their dissemination to SDM users.

2. The `zoon` R package encodes SDM analyses as a simple, but fully reproducible workflow of five steps: obtaining occurrence data, obtaining covariate data, pre-processing these data, fitting a model, and generating outputs.

3. Each of these steps is carried out by one or more community-contributed software modules maintained in a version-controlled online repository and built upon existing SDM code form R-packages and the research community.

4. `zoon` workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.

5. We demonstrate `zoon` by recreating SDM analyses from two published research articles as zoon workflows, which readers can interrogate and extend.

### Introduction

<!-- The Problem -->
Reproducibility crisis in species distribution modelling (SDM).
Difficult for new SDM users to access latest methods.
Difficult for anyone to compare methods effectively. 
Equally difficult for methods developers to disseminate their advances.

<!-- The Solution: ZOOOOOON!  -->
The `zoon` R package encodes SDM analyses as a simple workflow of five steps: obtaining occurrence data, obtaining covariate data, pre-processing these data, fitting a model, and generating outputs.
Each of these steps is carried out by one or more community-contributed software modules. 
This framework facilitates rapid dissemination of novel SDM methodologies by lowering the bar to creating research software.
`zoon` workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.

<!-- How It Works -->
### Implementing a `zoon` workflow

Module types (and examples) in a diagram (Figure 1).

```{r, echo = FALSE, fig.cap = "The modular SDM structure encoded by a zoon workflow. A) Flow diagram representing the module types. B) The required inputs and outputs for each module types (full details given in the `zoon` vignette 'Building a module'). C) Chaining and listing modules of the same type. D) An example workflow, corresponding to Example 1 below."}

# set up plotting window
plot.new()
par(mar = rep(0, 4))
plot.window(0:1, 0:1)

# grey rectangle (no background in vector pdf?)
rect(-1, -1, 2, 2, col = grey(0.9), border = NA)

# text over the top
text(0.5, 0.5, 'PLACEHOLDER',
     col = grey(0.3),
     cex = 3,
     xpd = NA)
```

Structure of workflow objects:

* code (call and modules used)
* output of each module; data, results and intermediate steps
* recording the session info and package and module versions

Extra stuff workflows do (handle cross validation, run parallel methods/data comparisons).

Things you can do to workflows:

* visualise the structure
* execute whole thing from scratch (grabs new data from web)
* execute from part way through

Include a figure visualising the sturcutre of the workflow object and how it can be adopted in the ChangeWorkflow function (corresponding to an example below?).

### Example Applications

We demonstrate the `zoon` R package by recreating two SDM analyses from published research articles and extending them.
Workflow objects created by these analyses can be accessed at [http://figshare.com/articles/zoon_applications_paper_workflows](http://figshare.com/articles/zoon_applications_paper_workflows).
We encourage readers to download, interrogate and alter these workflows for themselves.
Full code and metadata for all of the modules used in the examples below, can be found at [https://github.com/zoonproject/modules/R](https://github.com/zoonproject/modules/R)

#### Re-running a fairly simple SDM, tweaking some different stuff and looking at the results

Explanation & citation

One of the modules we had to write:

```{r, eval=FALSE}
AwesomeModule <- function(.df) {
    something
    return (something else)
}
```

What the workflow looks like

```{r, eval=FALSE}
carruthers_et_al <- workflow(
    occurrence = ...,
    covariate = ...,
    process = ...,
    model = ...,
    output = ...)
```

Some tweak we did to it:

```{r, eval=FALSE}
carruthers_et_al_interactive <- ReRunWorkflow(
    carruthers_et_al,
    output = InteractiveMap)
```

![The interactive map module overlay the raw data and predicted distribution, allowing users to interactivly explore thier results.](./figs/interactive_map.png)

Some plottable results of this analysis

#### Repeating a MEE methods comparison

Proposed paper:

>Merow & Silander (2014) A comparison of Maxlike and Maxent for modelling species distributions. *Methods in Ecology and Evolution*

This is a method that saw a lot of discussion and for which more experiments could be useful.

Possible experiments:
* evaluating with different criteria (e.g. deviance or pseudo-r-squared which measure calibration capacity)
* fitting the same models with GBIF data and evaluating against the BBS PA data,
* fitting/evaluating on spatially-stratified holdout data
* fitting the models with other species

(choose only one of these)

One of the modules we had to write (simpler than the MaxLike one!):

```{r, eval=FALSE}
CarolinaWrenPO <- function() {

  # load maxlike and the dataset into this environment
  zoon::GetPackage('maxlike')
  data(carw, envir = environment())
  occ <- na.omit(carw.data$pa.data)

  # keep only presences
  occ <- occ[occ$y == 1, ]

  # build the occurrence dataset and return
  data.frame(longitude = occ$Lon,
             latitude = occ$Lat,
             value = occ$y,
             type = 'presence'),
             fold = 1)

}
```

What the workflow looks like:

```{r, eval=FALSE}
merow_and_silander <- workflow(
    occurrence = CarolinaWrenPO,
    covariate = ...,
    process = ...,
    model = list(MaxEnt, MaxLike),
    output = ...)
```

So we ran it again with disc-based spatial stratification:

```{r, eval=FALSE}
merow_and_silander_spatial <- ReRunWorkflow(
    merow_and_silander,
    process = Chain(..., PartitionDisc))
```

### Future developments

Tutorials on how to create workflows and modules, as well as full technical details for module developers, are provided as vignettes distributed with `zoon`.

While the `zoon` R package can be used in isolation, future work will develop an online platform to facilitate exploring contributed modules and workflows.
This platform will also provide an online space to discuss, and openly evaluate, proposed best practices in SDM.
The zoon R package therefore represents a step towards a more reproducible ecosystem of SDM software.

### 2016-03-04 code/workflow_schematic.R

@@ -1,26 +1,0 @@
-# make figure 1, stick it in ./figs
-
-# Create a figs dir if it doesnt exist
-dir.create('figs')
-
-# set up device
-pdf('figs/workflow_schematic.pdf',
-    width = 6,
-    height = 2)
-
-# set up plotting window
-plot.new()
-par(mar = rep(0, 4))
-plot.window(0:1, 0:1)
-
-# grey rectangle (no background in vector pdf?)
-rect(-1, -1, 2, 2, col = grey(0.9), border = NA)
-
-# text over the top
-text(0.5, 0.5, 'PLACEHOLDER',
-     col = grey(0.3),
-     cex = 3,
-     xpd = NA)
-
-# close the plotting device
-dev.off()

### 2016-03-18 ms.Rmd
dumping Greg's bits in

@@
 
+<!-- from Greg: -->
+In 2008, Beale et al.
+(2008) suggested that “we are currently unable to build useful distribution models for many species”.
+Using null models, Beale et al.
+(2008) found that chance associations between climate and species’ distributions could account for the modelled relationships which they then suggested undermined the rationale for SDM.
+Amongst the replies to this article was a re-analysis by Araujou et al.
+(2009) that disputed this finding.
+Beale et al.
+(2008) put the alternative results to differences in distribution data, climate data and spatial domain, but emphasised the use of different null model distributions.
+Whilst some code was provided in the original publication by Beale et al.
+(2008), the following papers did not.
+The resolution of this very important discussion was hindered by the lack of data and code sharing.
+
+The second example comes from the study by Elith et al.
+(2006), that was exceptional in its scope by comparing “16 modelling methods over 226 species from 6 regions of the world”.
+This paper provided benchmarks that both figured highly in many researchers selection of method (and software).
+However, as the data and code is not available this much needed set of benchmarks cannot be updated by the community.
+For example, when a new modelling method is produced (e.g.
+GRAF, Golding…) authors cannot compare the outcomes to transparent set of community contributed modelling benchmarks and standards.
+In this case, the lack of data and code sharing hinder the legacy of one of the key SDM studies to date.
+
+Thirdly, Soininen and Luoto (2014) “assembled over 4900 published AUC (area under the curve of a receiver operating characteristic plot) values” to examine the predictability of SDM in relation to species traits.
+Having conducted this labour intensive review they suggested that researchers should “give precise reports of model calibration and evaluation methods” to enable subsequent met analyses.
+Relatedly, the methodological review of Yackulic et al.
+(2012) investigated the use of the MAXENT algorithm in 108 articles, and have a similar concern regarding the “transparency and usefulness” in the reporting of SDM.
+In these studies there is little to no opportunity afforded by the original studies to investigate the sensitivity of the conclusion to choices in model building and data selection.
+These studies point to a far wider reduction in scientific productivity due to reporting of SDM as ‘closed knowledge’.
+
+These three scenarios are illustrative of some of the need for the data and modelling aspects of SDM to become more repeatable, reproducible, accessible, modifiable and updateable.
+Each are crucial examples of the barriers to this science.
+Science has to be this way and written statements of the modelling carried out do not facilitate a productive science. 
+SDM, as a science, is not extensible given current working practices.
+
+##### Past work and other software - the USP of ZOON and requirements
+
+Most of the SDM users replying to a survey in 200X (Ahmed et al.
+2015) were using either MaxEnt or R as the first choice software for their analyses.
+The software market is much diverse, however, including a variety of software developed specifically for SDM (e.g.
+MaxEnt (), the BioMod () and dismo () packages for R, SDMtoolbox (Brown 2014), openModeller (), BioEnsembles (), ModEco ()) and software that has been appropriated for SDM studies, such as statistical software (R (), WinBugs (), OpenBugs (), Python()) and other less general software  (Domain (), Canoco (), MARS()).
+Reproducibility (and the issues discussed above) were not a primary requirement of these software, and whilst some have the capacity for an analysis to be repeated, many do not.
+Software such as BioMod, OpenModeller, BioEnsembles, ModEco were developed to enable users to carry out analyses with multiple models.
+The BIOVEL software (De Giovanni et al.
+2015) is an exception, sharing many of the general principles of ZOON, such as the increased accessibility to science via community contributed resources and increased sustainability by creating an e-infrastructure for this science.
+There are other workflow systems that can be used to wrap around SDM analyses (e.g.
+Kepler, Vis-Trails, Taverna), yet there has not been a large uptake by the community.
+Perhaps this is because of the unfamiliarity or assumed complexity of these tools.
+The SDM package (Naimi & Araujo 2016) has some functionality for users to incorporate new modelling methods into their analyses.
+
 <!-- The Solution: ZOOOOOON!  -->
@@
 `zoon` workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.
+
+ZOON allows users to make use of modules whether they are using a ZOON workflow or not. 
+Users can create new modules for any modelling method and integrate it into another analysis. 
+Unlike R itself, there are far reduced overheads on creating a package...
 

### 2016-03-18 ms.Rmd
Seperate paras into one sentence per line.

@@
 <!-- from Greg: -->
-In 2008, Beale et al.
-(2008) suggested that “we are currently unable to build useful distribution models for many species”.
+In 2008, Beale et al. (2008) suggested that “we are currently unable to build useful distribution models for many species”. 
 Using null models, Beale et al.
 (2008) found that chance associations between climate and species’ distributions could account for the modelled relationships which they then suggested undermined the rationale for SDM.
-Amongst the replies to this article was a re-analysis by Araujou et al.
-(2009) that disputed this finding.
-Beale et al.
-(2008) put the alternative results to differences in distribution data, climate data and spatial domain, but emphasised the use of different null model distributions.
-Whilst some code was provided in the original publication by Beale et al.
-(2008), the following papers did not.
+Amongst the replies to this article was a re-analysis by Araujou et al. (2009) that disputed this finding. 
+Beale et al. (2008) put the alternative results to differences in distribution data, climate data and spatial domain, but emphasised the use of different null model distributions. 
+Whilst some code was provided in the original publication by Beale et al. (2008), the following papers did not. 
 The resolution of this very important discussion was hindered by the lack of data and code sharing.
 
-The second example comes from the study by Elith et al.
-(2006), that was exceptional in its scope by comparing “16 modelling methods over 226 species from 6 regions of the world”.
+The second example comes from the study by Elith et al. (2006), that was exceptional in its scope by comparing “16 modelling methods over 226 species from 6 regions of the world”. 
 This paper provided benchmarks that both figured highly in many researchers selection of method (and software). 
 However, as the data and code is not available this much needed set of benchmarks cannot be updated by the community. 
-For example, when a new modelling method is produced (e.g.
-GRAF, Golding…) authors cannot compare the outcomes to transparent set of community contributed modelling benchmarks and standards.
+For example, when a new modelling method is produced (e.g. GRAF, Golding…) authors cannot compare the outcomes to transparent set of community contributed modelling benchmarks and standards. 
 In this case, the lack of data and code sharing hinder the legacy of one of the key SDM studies to date.
@@
 Thirdly, Soininen and Luoto (2014) “assembled over 4900 published AUC (area under the curve of a receiver operating characteristic plot) values” to examine the predictability of SDM in relation to species traits.
-Having conducted this labour intensive review they suggested that researchers should “give precise reports of model calibration and evaluation methods” to enable subsequent met analyses.
+Having conducted this labour intensive review they suggested that researchers should “give precise reports of model calibration and evaluation methods” to enable subsequent meta-analyses. 
 Relatedly, the methodological review of Yackulic et al.
@@
 
-Most of the SDM users replying to a survey in 200X (Ahmed et al.
-2015) were using either MaxEnt or R as the first choice software for their analyses.
+Most of the SDM users replying to a survey in 2015 (Ahmed et al. 2015) were using either MaxEnt or R as the first choice software for their analyses. 
 The software market is much diverse, however, including a variety of software developed specifically for SDM (e.g.
@@
 2015) is an exception, sharing many of the general principles of ZOON, such as the increased accessibility to science via community contributed resources and increased sustainability by creating an e-infrastructure for this science.
-There are other workflow systems that can be used to wrap around SDM analyses (e.g.
-Kepler, Vis-Trails, Taverna), yet there has not been a large uptake by the community.
+There are other workflow systems that can be used to wrap around SDM analyses (e.g. Kepler, Vis-Trails, Taverna), yet there has not been a large uptake by the community. 
 Perhaps this is because of the unfamiliarity or assumed complexity of these tools. 
@@
 Each of these steps is carried out by one or more community-contributed software modules. 
+Firstly, occurrence data must be read into the software package or downloaded from online repositories.
+Occurrence usually takes the form of presence-absence data or presence-only data, though abundance data is also used.
+Secondly, predictor variables or covariates (typically environmental covariates) are required and the values of these covariates, at the locations of the occurrence data, must be extracted.
+Thirdly, a wide variety of processes are often applied to the occurrence and covariate data.
+These processes include data cleaning, data thinning to account for spatial biases, feature selection using PCA or association tests and the splitting of data into training and test sets or cross validation folds.
+Fourthly, once the data has been suitable manipulated a model is fitted to estimate the relationships between the covariates and occurrence data.
+These models include simple statistical models such as GLMs as well as modern, flexible machine-learning methods such as MaxEnt and boosted regression trees.
+Finally, model outputs are needed.
+The predictive power and goodness of fit must be assessed and the model parameters or response curves must be examined.
+The model is likely to be used to predict species occurrence, either in the vicinity of the occurrence data or elsewhere, or into the past or future.
+
+This structure defines how `zoon` approaches SDM. 
+For each of these steps, the user selects one or more 'modules' and combined them in a call to `workflow`.
+To combine multiple modules of the same type we provide the `Chain` command.
+For occurrence and covariate modules, this command takes multiple modules and simply combines the data acquired by each module.
+Chained process models are run sequentially.
+For example if a user wants to generate background or pseudo absence data and then split the data into crossvalidation folds, modules implementing these two seperate process would be chained in that order.
+Finally, chained output modules are simply all run seperately allowing the user to create multiple maps and summary figures, calcualte performance metrics and create other model outputs in one workflow.
+Model modules cannot be chained.
+
 This framework facilitates rapid dissemination of novel SDM methodologies by lowering the bar to creating research software.
@@
 
-Module types (and examples) in a diagram (Figure 1).
-
+Module types (and examples) in a diagram 
 ```{r, echo = FALSE, fig.cap = "The modular SDM structure encoded by a zoon workflow.
@@
 
-Include a figure visualising the sturcutre of the workflow object and how it can be adopted in the ChangeWorkflow function (corresponding to an example below?).
+Include a figure visualising the structure of the workflow object and how it can be adopted in the ChangeWorkflow function (corresponding to an example below?).
 

### 2016-03-21 ms.Rmd
setting up bibliography & MEE-style citations

@@
 author: "Nick Golding, Tom August, Tim C.D. Lucas, Emiel Van Loon & Greg McInerny"
-date: "4 March 2016"
+date: "21 March 2016"
+bibliography: zoon_app.bib
+csl: mee.csl
 output: 
@@
 <!-- The Problem -->
+
+SDM boom & growth of methods literature [@Barbosa2015]
+
 Reproducibility crisis in species distribution modelling (SDM).
+
 Difficult for new SDM users to access latest methods.
+
 Difficult for anyone to compare methods effectively.
+
 Equally difficult for methods developers to disseminate their advances.
@@
 The zoon R package therefore represents a step towards a more reproducible ecosystem of SDM software.
+
+### References

### 2016-03-21 ms.Rmd
adding citations for Greg's intro text

@@
 <!-- from Greg: -->
-In 2008, Beale et al. (2008) suggested that “we are currently unable to build useful distribution models for many species”. 
-Using null models, Beale et al.
-(2008) found that chance associations between climate and species’ distributions could account for the modelled relationships which they then suggested undermined the rationale for SDM.
-Amongst the replies to this article was a re-analysis by Araujou et al. (2009) that disputed this finding. 
-Beale et al. (2008) put the alternative results to differences in distribution data, climate data and spatial domain, but emphasised the use of different null model distributions. 
-Whilst some code was provided in the original publication by Beale et al. (2008), the following papers did not. 
+In 2008, @Beale2008 suggested that “we are currently unable to build useful distribution models for many species”. 
+Using null models, @Beale2008 found that chance associations between climate and species’ distributions could account for the modelled relationships which they then suggested undermined the rationale for SDM.
+Amongst the replies to this article was a re-analysis by @Araujo2009 that disputed this finding. 
+@Beale2008 put the alternative results to differences in distribution data, climate data and spatial domain, but emphasised the use of different null model distributions. 
+Whilst some code was provided in the original publication by @Beale2008, the following papers did not. 
 The resolution of this very important discussion was hindered by the lack of data and code sharing.
 
-The second example comes from the study by Elith et al. (2006), that was exceptional in its scope by comparing “16 modelling methods over 226 species from 6 regions of the world”. 
+The second example comes from the study by @Elith2006, that was exceptional in its scope by comparing “16 modelling methods over 226 species from 6 regions of the world”. 
 This paper provided benchmarks that both figured highly in many researchers selection of method (and software). 
@@
 
-Thirdly, Soininen and Luoto (2014) “assembled over 4900 published AUC (area under the curve of a receiver operating characteristic plot) values” to examine the predictability of SDM in relation to species traits.
+Thirdly, @Soininen2014 “assembled over 4900 published AUC (area under the curve of a receiver operating characteristic plot) values” to examine the predictability of SDM in relation to species traits.
 Having conducted this labour intensive review they suggested that researchers should “give precise reports of model calibration and evaluation methods” to enable subsequent meta-analyses. 
-Relatedly, the methodological review of Yackulic et al.
-(2012) investigated the use of the MAXENT algorithm in 108 articles, and have a similar concern regarding the “transparency and usefulness” in the reporting of SDM.
+Relatedly, the methodological review of @Yackulic2013 investigated the use of the MAXENT algorithm in 108 articles, and have a similar concern regarding the “transparency and usefulness” in the reporting of SDM.
 In these studies there is little to no opportunity afforded by the original studies to investigate the sensitivity of the conclusion to choices in model building and data selection. 
@@
 Software such as BioMod, OpenModeller, BioEnsembles, ModEco were developed to enable users to carry out analyses with multiple models.
-The BIOVEL software (De Giovanni et al.
-2015) is an exception, sharing many of the general principles of ZOON, such as the increased accessibility to science via community contributed resources and increased sustainability by creating an e-infrastructure for this science.
+The BIOVEL software [@DeGiovanni2015] is an exception, sharing many of the general principles of ZOON, such as the increased accessibility to science via community contributed resources and increased sustainability by creating an e-infrastructure for this science.
 There are other workflow systems that can be used to wrap around SDM analyses (e.g. Kepler, Vis-Trails, Taverna), yet there has not been a large uptake by the community. 
 Perhaps this is because of the unfamiliarity or assumed complexity of these tools. 
-The SDM package (Naimi & Araujo 2016) has some functionality for users to incorporate new modelling methods into their analyses.
+The SDM package [@Naimi2016] has some functionality for users to incorporate new modelling methods into their analyses.
 

### 2016-03-21 ms.Rmd
restructuring intro, adding figure referencing

@@
 
-Difficult for new SDM users to access latest methods.
-
-Difficult for anyone to compare methods effectively.
-
-Equally difficult for methods developers to disseminate their advances.
-
 <!-- from Greg: -->
@@
 
-##### Past work and other software - the USP of ZOON and requirements
+Core, soluble problems:
+
+Difficult for new SDM users to access latest methods.
+
+Difficult for anyone to compare methods effectively.
+
+Equally difficult for methods developers to disseminate their advances.
+
+<!-- Past work towards this goal & where it falls short - the USP of ZOON and requirements -->
 
@@
 MaxEnt (), the BioMod () and dismo () packages for R, SDMtoolbox (Brown 2014), openModeller (), BioEnsembles (), ModEco ()) and software that has been appropriated for SDM studies, such as statistical software (R (), WinBugs (), OpenBugs (), Python()) and other less general software  (Domain (), Canoco (), MARS()).
+
 Reproducibility (and the issues discussed above) were not a primary requirement of these software, and whilst some have the capacity for an analysis to be repeated, many do not. 
 Software such as BioMod, OpenModeller, BioEnsembles, ModEco were developed to enable users to carry out analyses with multiple models.
+
 The BIOVEL software [@DeGiovanni2015] is an exception, sharing many of the general principles of ZOON, such as the increased accessibility to science via community contributed resources and increased sustainability by creating an e-infrastructure for this science.
@@
 Perhaps this is because of the unfamiliarity or assumed complexity of these tools.
-The SDM package [@Naimi2016] has some functionality for users to incorporate new modelling methods into their analyses.
+
+The SDM package [@Naimi2016] implements a number of modern SDM approaches in a simple modular framework.
+This package also has some functionality for users to incorporate new modelling methods into their analyses.
+There is, however still a signifiact technical barrier to those wishing to incorporate their methods, and to share them more widely.
+
+#### The ZOON project
+
+The ZOON project takes a different approach, instead defining only the core architecture of workflows, and interfacing with an open repository of community-contributed software modules.
+This paper introduces version 0.4-22 of the `zoon` R package, which encodes the majority of this functionality.
+First, we describe the modular structure of zoon workflows, illustrate how to download and upload software modules, and demonstrate how completed `zoon` workflows can be shared, evaluated and extended.
+Then we illustrate these concepts, and illustrate the contribution we hope `zoon` will make to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.
 
 <!-- The Solution: ZOOOOOON!  -->
+### The `zoon` R package
+
+Versions 0.4-22 and later of the `zoon` R package can be installed directly from CRAN.
+
+#### Constructing a `zoon` workflow
+
 The `zoon` R package encodes SDM analyses as a simple workflow of five steps: obtaining occurrence data, obtaining covariate data, pre-processing these data, fitting a model, and generating outputs.
 Each of these steps is carried out by one or more community-contributed software modules.
+
+Figure \ref{fig:workflows} illustrates the structure of a `zoon` workflow.
+
+```{r, echo = FALSE, fig.cap = "The modular SDM structure encoded by a zoon workflow.
+A) Flow diagram representing the module types.
+B) The required inputs and outputs for each module types (full details given in the `zoon` vignette 'Building a module').
+C) Chaining and listing modules of the same type.
+D) An example workflow, corresponding to Example 1 below.\\label{fig:workflows}"}
+
+# set up plotting window
+plot.new()
+par(mar = rep(0, 4))
+plot.window(0:1, 0:1)
+
+# grey rectangle (no background in vector pdf?)
+rect(-1, -1, 2, 2, col = grey(0.9), border = NA)
+
+# text over the top
+text(0.5, 0.5, 'PLACEHOLDER',
+     col = grey(0.3),
+     cex = 3,
+     xpd = NA)
+```
+
 Firstly, occurrence data must be read into the software package or downloaded from online repositories.
@@
 
-<!-- How It Works -->
-### Implementing a `zoon` workflow
-
-Module types (and examples) in a diagram 
-```{r, echo = FALSE, fig.cap = "The modular SDM structure encoded by a zoon workflow.
-A) Flow diagram representing the module types.
-B) The required inputs and outputs for each module types (full details given in the `zoon` vignette 'Building a module').
-C) Chaining and listing modules of the same type.
-D) An example workflow, corresponding to Example 1 below."}
-
-# set up plotting window
-plot.new()
-par(mar = rep(0, 4))
-plot.window(0:1, 0:1)
-
-# grey rectangle (no background in vector pdf?)
-rect(-1, -1, 2, 2, col = grey(0.9), border = NA)
-
-# text over the top
-text(0.5, 0.5, 'PLACEHOLDER',
-     col = grey(0.3),
-     cex = 3,
-     xpd = NA)
-```
-
 Structure of workflow objects:
@@
 
-#### Re-running a fairly simple SDM, tweaking some different stuff and looking at the results
+#### Example 1. A fairly simple SDM, tweaking some different stuff and looking at the results
 

### 2016-03-21 ms.Rmd
working on paper structure

@@
 
-### Abstract
+# Abstract
 
@@
 
-### Introduction
+# Introduction
 
@@
 
-Reproducibility crisis in species distribution modelling (SDM).
+... reproducibility crisis in SDM.
 
-<!-- from Greg: -->
+<!-- from Greg, will need trimming down -->
 In 2008, @Beale2008 suggested that “we are currently unable to build useful distribution models for many species”. 
@@
 
-Core, soluble problems:
+### Solutions to these problems
+
+The core, soluble problems are:
 
@@
 <!-- Past work towards this goal & where it falls short - the USP of ZOON and requirements -->
+### Reproducible software projects
 
@@
 
-#### The ZOON project
+### The ZOON project
 
@@
 This paper introduces version 0.4-22 of the `zoon` R package, which encodes the majority of this functionality.
-First, we describe the modular structure of zoon workflows, illustrate how to download and upload software modules, and demonstrate how completed `zoon` workflows can be shared, evaluated and extended.
-Then we illustrate these concepts, and illustrate the contribution we hope `zoon` will make to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.
+First, we describe the modular structure of zoon workflows and demonstrate how completed `zoon` workflows can be shared, evaluated and extended.
+Then we illustrate these concepts, and the contribution we hope `zoon` will make to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.
 
 <!-- The Solution: ZOOOOOON!  -->
-### The `zoon` R package
+# The `zoon` R package
 
@@
 
-#### Constructing a `zoon` workflow
+### Constructing a `zoon` workflow
 
@@
 C) Chaining and listing modules of the same type.
-D) An example workflow, corresponding to Example 1 below.\\label{fig:workflows}"}
+D) The structure of a workflow object.\\label{fig:workflows}"}
+
+# four panel plot, no margins
+par(mar = c(1, 1, 2, 1),
+    mfrow = c(2, 2))
+
+# placeholder plots for each panel
+for (i in 1:4) {
 
@@
   plot.new()
-par(mar = rep(0, 4))
   plot.window(0:1, 0:1)
@@
        col = grey(0.3),
-     cex = 3,
+       cex = -13,
        xpd = NA)
+
+  # add a panel letter
+  mtext(text = LETTERS[i],
+        side = 3,
+        line = 0.5,
+        adj = 0)  
+}
 ```
 
-Firstly, occurrence data must be read into the software package or downloaded from online repositories.
-Occurrence usually takes the form of presence-absence data or presence-only data, though abundance data is also used.
-Secondly, predictor variables or covariates (typically environmental covariates) are required and the values of these covariates, at the locations of the occurrence data, must be extracted.
-Thirdly, a wide variety of processes are often applied to the occurrence and covariate data.
+#### Module types
+
+* *Occurrence* - Usually presence-absence data or presence-only data, though abundance data is also used.
+* *Covariates* - Predictor variables or covariates (typically environmental covariates) are required and the values of these covariates, at the locations of the occurrence data, must be extracted.
+* *Process* - Processes  applied to the occurrence and covariate data.
 These processes include data cleaning, data thinning to account for spatial biases, feature selection using PCA or association tests and the splitting of data into training and test sets or cross validation folds.
-Fourthly, once the data has been suitable manipulated a model is fitted to estimate the relationships between the covariates and occurrence data.
+* *Model* - Once the data has been suitable manipulated a model is fitted to estimate the relationships between the covariates and occurrence data.
 These models include simple statistical models such as GLMs as well as modern, flexible machine-learning methods such as MaxEnt and boosted regression trees.
-Finally, model outputs are needed.
-The predictive power and goodness of fit must be assessed and the model parameters or response curves must be examined.
+* *Output* - The predictive power and goodness of fit must be assessed and the model parameters or response curves must be examined.
 The model is likely to be used to predict species occurrence, either in the vicinity of the occurrence data or elsewhere, or into the past or future.
@@
 Unlike R itself, there are far reduced overheads on creating a package...
+
+### Inspecting, sharing and extending a `zoon` workflow
 
@@
 
-#### Example 1. A fairly simple SDM, tweaking some different stuff and looking at the results
+#### Example 1. Modelling the distribution of ???
 
-Explanation & citation
+??? constructed a species distribution model for ??? using ???.
 
-One of the modules we had to write:
-
-```{r, eval=FALSE}
-AwesomeModule <- function(.df) {
-    something
-    return (something else)
-}
-```
-
-What the workflow looks like
+Such a model can be re-constructed as a `zoon` workflow using modules which are already available in the software repository:
 
@@
 
+The resulting workflow contains all the c
 Some tweak we did to it:
@@
 
-While the `zoon` R package can be used in isolation, future work will develop an online platform to facilitate exploring contributed modules and workflows.
+The `zoon` packages provides solutions to some of the technical barriers to the ultimate goal of more open, productive, reproducible SDM research.
+Achieving this goal will also depend on overcoming the social hurdle of developing a common tool for comparing and quantitatively evaluating SDMs.
+
+This will require achieving a critical mass of `zoon` modules, so that there is an incentive for species distribution modellers to use the software, and a critical mass of users, so that there is an incentive for SDM methods developers to encode their proposed methods as modules in the community repo.
+At this point `zoon` will provide a common environment in which to evaluate SDM methods and enable the SDM research community to make evidence-based decisions about best practice and the goals for development of the field.
+
+In order to facilitate this social aspect, future work will develop an online platform to enable exploring contributed modules and workflows.
 This platform will also provide an online space to discuss, and openly evaluate, proposed best practices in SDM.

### 2016-03-21 ms.Rmd
fleshing out the paper

@@
 The ZOON project takes a different approach, instead defining only the core architecture of workflows, and interfacing with an open repository of community-contributed software modules.
+Users can create new modules for any modelling method and integrate it into another analysis.
+
+Similarly to programming languages such as R and Python, facilitating wider development of software will enable the whole research community to develop the software that it requires, rather than rely on the developers of the software they use to do it for them.
+Unlike software packages for programming languages however, the ZOON project enforces a modular structure that ensures interoperability between contributed software modules.
+This also helps to lower the bar for creating modules (each module is simply an R function with a set of required inputs and outputs) and facilitates automated testing of modules ot ensure software quality .
+
 This paper introduces version 0.4-22 of the `zoon` R package, which encodes the majority of this functionality.
@@
 Each of these steps is carried out by one or more community-contributed software modules.
+For each of these steps, the user selects one or more 'modules' and combined them in a call to `workflow`.
 
@@
 
-This structure defines how `zoon` approaches SDM. 
-For each of these steps, the user selects one or more 'modules' and combined them in a call to `workflow`.
+#### Lists and Chains
+
 To combine multiple modules of the same type we provide the `Chain` command.
@@
 
-This framework facilitates rapid dissemination of novel SDM methodologies by lowering the bar to creating research software.
-`zoon` workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.
-
-ZOON allows users to make use of modules whether they are using a ZOON workflow or not. 
-Users can create new modules for any modelling method and integrate it into another analysis. 
-Unlike R itself, there are far reduced overheads on creating a package...
+Add something on lists!
 
 ### Inspecting, sharing and extending a `zoon` workflow
+
+`zoon` workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.
 
@@
 
+#### Building modules
+
+Modules are simply R function definitions, with a set of required inputs and outputs.
+For example, the`Bioclim` module, which uses the `raster` R package [@raster] to download the widely used bioclim [] covariate layers in the correct format, is defined as:
+
+```{r, eval=FALSE}
+Bioclim <- function (extent = c(-180, 180, -90, 90),
+                     resolution = 10,
+                     layers = 1:19) {
+
+    if(!(resolution %in% c(2.5, 5, 10))){
+      stop('only 2.5, 5 and 10 degree resolutions are supported currently')
+    }
+
+    if (length(layers) < 1 |
+        length(layers) > 19 |
+        !all(layers %in% 1:19)) {
+      stop ('layers must be a vector of integers between 1 and 19 indicating layers to obtain')
+    }
+
+    stopifnot(length(extent) == 4)
+    stopifnot(all(is.numeric(extent)))
+
+    world <- getData('worldclim', var = 'bio', res = resolution)
+    world <- world[[layers]]
+    cropped <- crop(world, extent(extent))
+    return (cropped)
+  }
+```
+
+This function meets the minimal requirements of a *covariate* module (see Figure \ref{fig:workflows} Panel B); returning a `Raster*` object of the correct dimensions.
+
+As long as a valid module function is defined in an R session, it can be used in workflows in combination with modules downloaded from the ZOON module repository.
+Whilst this makes it easy to develop modules locally, the strength of `zoon` comes from the ability to upload modules to the online repository so that others can access them.
+
+To do this, it first is necessary to provide additional metadata including the module tile and description, module type, author name and definitions for any non-mandatory arguments.
+`zoon` provides the function `BuildModule()` to facilitate entering this metadata and checking that it matches the module.
+
+The `zoon` vignette *Building modules* provides a detailed tutorial for building modules of each type.
+
 ### Example Applications
@@
 
-Such a model can be re-constructed as a `zoon` workflow using modules which are already available in the software repository:
+Such a model can be quickly and easily re-constructed as a `zoon` workflow using modules which are already available in the software repository:
 
@@
 
-The resulting workflow contains all the c
-Some tweak we did to it:
+The resulting workflow contains all the code required to re-run the workflow, and the results of executing each module.
+`zoon` provides functionality to quickly update an existing workflow object (even one created by someone else) by switching out modules.
+
+Next, we update the workflow to produce an interactive map enabling anyone to inspect the data and  predictions on a zoomable map, and to inspect the response curves of the fitted model:
 
@@
     carruthers_et_al,
-    output = InteractiveMap)
+    output = Chain(InteractiveMap, ResponseCurve))
 ```
 
-![The interactive map module overlay the raw data and predicted distribution, allowing users to interactivly explore thier results.](./figs/interactive_map.png)
+![The interactive map module overlay the raw data and predicted distribution, allowing users to interactively explore their results.](./figs/interactive_map.png)
 
-Some plottable results of this analysis
+<!-- add response curves output here -->
 
-#### Repeating a MEE methods comparison
+`zoon` users can inspect existing workflows using these and a range of other *output* modules, or create and share their own output modules.
 
-Proposed paper:
+#### Example 2. Evaluating the MaxLike model
 
->Merow & Silander (2014) A comparison of Maxlike and Maxent for modelling species distributions. *Methods in Ecology and Evolution*
+@Royle2012 proposed a model for fitting SDMs to presence-only data, which they suggest is able to estimate the absolute (rather than relative) probability of species presence in each grid cell.
+This model and the associated claims have been the subject of several research articles evaluating the model against both real-world and simulated datasets [@Fitzpatrick2013;@Merow2014;@Hastie2013;@Phillips2013].
 
-This is a method that saw a lot of discussion and for which more experiments could be useful.
+In this example we illustrate how `zoon` can help contribute to discussions such as these by enabling rapid evaluation of a new method against a variety of datasets and test procedures.
 
+For this example, it was necessary to create three new modules: two to load the presence/absence and presence-only Carolina Wren datasets from @Royle2012; and one to implement the MaxLike model.
+
+<!--
 Possible experiments:
@@
 * fitting the models with other species
+(choose only one of these)
+-->
 
-(choose only one of these)
-
-One of the modules we had to write (simpler than the MaxLike one!):
+All of these have now been uploaded to the ZOON modules repository (under the names `CarolinaWrenPO`, `CarolinaWrenPA`, and `MaxLike`), though we diusplay the code required for the module `CarolinaWrenPO` here as an illustration of the minimal overhead required in writing `zoon` software modules:
 
@@
 
-  # load maxlike and the dataset into this environment
+  # load maxlike package (installing if it isn't in the library)
   zoon::GetPackage('maxlike')
+
+  # load the dataset from the maxlike package
   data(carw, envir = environment())
@@
 
-  # keep only presences
+  # keep only presence records
   occ <- occ[occ$y == 1, ]
@@
 ```
+
+This function meets the requirements of an *occurrence* module; providing as output a dataframe with the required columns and datatypes.
+
+We can run automated checks on this module
 

### 2016-03-21 ms.Rmd
adding armadillo example

@@
 
+`zoon` provides functionality to quickly update an existing workflow object (even one created by someone else) by switching out modules.
+
 Include a figure visualising the structure of the workflow object and how it can be adopted in the ChangeWorkflow function (corresponding to an example below?).
@@
 
-#### Example 1. Modelling the distribution of ???
+#### Example 1. Modelling the potential distribution of nine-banded armadillo in the USA
 
-??? constructed a species distribution model for ??? using ???.
+@Feng2015 constructed a MaxEnt species distribution model for nine-banded armadillo in the USA using presence-only data on the species' current distribution, and the bioclim [] set of environmental correlates.
+This model was then used to predict areas in the Americas which may be suitable for the species to become established.
 
-Such a model can be quickly and easily re-constructed as a `zoon` workflow using modules which are already available in the software repository:
+Such a model can be quickly and easily re-constructed as a `zoon` workflow using modules available in the ZOON module repository.
+@Feng2015 used a combination of occurrence data from GBIF, and additional occurrence data manually collected from the published literature.
+Unfortunately the latter data have not been made publically available, so here we use only data from GBIF.
+If the the additional data had been made available it would be straightforward to incorporate them, for example using the `LocalOccurrenceData` module.
+
+<!--
+covariates Feng & Papes used:
+1    annual mean temperature
+2    mean diurnal range
+3    isothermality
+4    temperature seasonality
+6    minimum temperature of the coldest month
+9    mean temperature of the driest quarter
+10   mean temperature of the warmest quarter
+12   annual precipitation
+15   precipitation seasonality
+-->
 
 ```{r, eval=FALSE}
-carruthers_et_al <- workflow(
-    occurrence = ...,
-    covariate = ...,
-    process = ...,
-    model = ...,
-    output = ...)
+Feng_Papes <- workflow(
+    occurrence = SpOcc('Dasypus novemcinctus',
+                       extent = c(-130, -20, -60, 60)),
+    covariate = Bioclim(extent = c(-130, -20, -60, 60),
+                        layers = c(1:4, 6, 9, 10, 12, 15)),
+    process = Chain(TidyGBIF, # some sort of data cleaning?
+                    BufferBackground, # background w/in 2 degrees of occurrences
+                    MESSMask),  # mask out areas beyond extremes in train data
+    model = MaxEnt,
+    output = PrintMap)  # They plotted at 5% omission error level
 ```
 
-The resulting workflow contains all the code required to re-run the workflow, and the results of executing each module.
-`zoon` provides functionality to quickly update an existing workflow object (even one created by someone else) by switching out modules.
+The resulting workflow contains all the code required to re-run the workflow, the input data and the results of executing each module.
+The object `Feng_Papes` could therefore be saved as a binary file and shared as a reproducible representation of this research.
 
@@
 ```{r, eval=FALSE}
-carruthers_et_al_interactive <- ReRunWorkflow(
-    carruthers_et_al,
-    output = Chain(InteractiveMap, ResponseCurve))
+Feng_Papes_interactive <- ReRunWorkflow(
+    Feng_Papes,
+    output = Chain(PrintMap, InteractiveMap, ResponseCurve))
 ```
@@
 
-<!-- add response curves output here -->
+<!-- add the original and the response curves to this figure too -->
 
-`zoon` users can inspect existing workflows using these and a range of other *output* modules, or create and share their own output modules.
-
-#### Example 2. Evaluating the MaxLike model
+#### Example 2. Evaluating MaxLike
 

### 2016-03-21 ms.Rmd
sorting out figure placement ad todying a couple od references

@@
     fig_caption: yes
+    includes:  
+      in_header: latex_preamble.tex
 ---
@@
 
-Next, we update the workflow to produce an interactive map enabling anyone to inspect the data and  predictions on a zoomable map, and to inspect the response curves of the fitted model:
+Next, we update the workflow to produce an interactive map enabling anyone to inspect the data and  predictions on a zoomable map, and to inspect the response curves of the fitted model.
+These outputs are shown in Figure \ref{fig:interactive}.
 
@@
 
-![The interactive map module overlay the raw data and predicted distribution, allowing users to interactively explore their results.](./figs/interactive_map.png)
+![The interactive map module overlay the raw data and predicted distribution, allowing users to interactively explore their results.\label{fig:interactive}](./figs/interactive_map.png)
 

### 2016-03-21 compile.sh
removing files no longer needed

@@ -1,12 +1,0 @@
-#!/bin/bash
-
-# run code and compile the ms to pdf
-
-# make directories for the figrues and output, if they don't already exist
-mkdir -p figs output
-
-# run the R code to create the main figure
-Rscript code/workflow_schematic.R
-
-# compile the manuscript using pandoc & latex
-pandoc -s --template=tex_template/default.latex -H tex_template/header.tex -V papersize=a4paper ms.md -o output/zoon_applications_paper.pdf

### 2016-03-21 ms.Rmd
adding a first attempt at a programmtic workflow illustration

@@
 
-```{r, echo = FALSE, fig.cap = "The modular SDM structure encoded by a zoon workflow.
+```{r, echo = FALSE, fig.height = 9, fig.cap = "The modular SDM structure encoded by a zoon workflow.
 A) Flow diagram representing the module types.
@@
 
-# four panel plot, no margins
-par(mar = c(1, 1, 2, 1),
-    mfrow = c(2, 2))
+# four panel plot
+par(mfrow = c(2, 2))
 
+#plot the saved image in the first panel
+library(raster)
+r <- brick('figs/workflow_basic.png')
+par(mar = c(1, 1, 2, 1))
+plotRGB(r, maxpixels = Inf)
+
+# add a panel letter
+# this is an awful hack and completely contingent on the screenshot size :/
+text(40, 980, labels = LETTERS[1],
+      xpd = NA,
+     cex = 1.2)
+
+par(mar = c(1, 1, 2, 1))
 # placeholder plots for each panel
-for (i in 1:4) {
+for (i in 2:4) {
 

### 2016-03-21 ms.Rmd
Add paragraphs on lists.

@@
 
-Add something on lists!
+Many SDM analyses require running similar workflow in parallel. 
+In applied settings this may be running the same analysis for multiple species.
+In methodological work, it is important to compare new methods or models to benchmarks.
+To divide a `zoon` workflow into multiple parallel analyses the `list` function is used.
+`list` can be used to split the analysis at any point: listing multiple occurrence modules together will run the workflow on different occurrence datasets while listing multiple model modules will fit different models to a shared, identical dataset but then apply output modules to each model separately.
+
+The complete workflow, with multiple modules defined with the `Chain` function and parallel analyses being defined with the `list` function can be visualised by calling `plot` with the workflow object as it's argument.
+This displays the names of all modules and the broader structure of the workflow as well as displaying which modules are not available in the online repository (Figure ).
 

### 2016-03-21 ms.Rmd
Add info on rerunworkflow, changeworkflow and crossvalidation.

@@
 
+#### Crossvalidation
+
+`workflow` sensibly handles cross-validation.
+Process modules are used to split the data into a training and test set or into multiple folds so that each fold in turn will be held back and used as the test for the predictive ability of any fitted models.
+The same test/train splits will be used for each analysis if the multiple models or methods are being used so that model performance measures are comparable.
+Furthermore, the test/train splits are saved along with the workflow so that other researchers adding to a workflow can make fair comparisons. 
+After models are fitted to these subsets of the data, they are also fitted to the entire dataset as this will provide the best models for predictions from further new data.
+
 ### Inspecting, sharing and extending a `zoon` workflow
-
-`zoon` workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.
 
@@
 
-Extra stuff workflows do (handle cross validation, run parallel methods/data comparisons).
+`zoon` workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.
+A zoon workflow returns an object of class `zoonWorkflow`.
+This object contains all the information needed to reproduce the analysis.
+This includes the original call, a record of the packages and modules used and their versions when the analysis was run as well as all the data and fitted models.
+
+A workflow is reproducible, reusable and alterable by way of the functions `RerunWorkflow` and `ChangeWorkflow`.
+Using `RerunWorkflow` a workflow object can be run from scratch or from part way though the analysis to avoid computationally slow steps.
+`ChangeWorkflow` takes as it's arguments a workflow object and new selections for any or all of the workflow steps.
+The workflow is then rerun from the first altered step.
 

### 2016-03-23 ms.Rmd
Update Feng_Papes workflow with new modules.

@@
 ```{r, eval=FALSE}
-Feng_Papes <- workflow(
-    occurrence = SpOcc('Dasypus novemcinctus',
+Feng_Papes <- workflow(occurrence = SpOcc('Dasypus novemcinctus',
                                           extent = c(-130, -20, -60, 60)),
@@
                                             layers = c(1:4, 6, 9, 10, 12, 15)),
-    process = Chain(TidyGBIF, # some sort of data cleaning?
-                    BufferBackground, # background w/in 2 degrees of occurrences
-                    MESSMask),  # mask out areas beyond extremes in train data
+                          process = Chain(Clean, # Remove rows w/ impossible, unlikely or missing lat lon.
+                                          Background(n = 1000, bias = 200), #background w/in 200km of occurrences
+                                          # MESSMask),  # Still to be implemented. mask out areas beyond extremes in train data
+                                          Crossvalidate(k = 5)),
                             model = MaxEnt,
-    output = PrintMap)  # They plotted at 5% omission error level
+                           output = Chain(PerformanceMeasures, PrintMap))
+
 ```

### 2016-03-23 ms.Rmd
Add chunk to load zoon

@@
 
-```{r, eval=FALSE}
+```{r, eval = FALSE}
 Bioclim <- function (extent = c(-180, 180, -90, 90),
@@
 
-```{r, eval=FALSE}
+<!---
+Notes:
+
+-->
+
+```{r, eval = TRUE, echo = FALSE, cache = FALSE}
+# Keep cache = FALSE. Not supposed to cache chunks with library()
+#  I assume we want this not echoed.
+library(zoon)
+
+```
+
+```{r, eval = TRUE}
 Feng_Papes <- workflow(occurrence = SpOcc('Dasypus novemcinctus',
@@
 
-```{r, eval=FALSE}
+```{r, eval = FALSE}
 Feng_Papes_interactive <- ReRunWorkflow(
@@
 
-```{r, eval=FALSE}
+```{r, eval = FALSE}
 CarolinaWrenPO <- function() {
@@
 
-```{r, eval=FALSE}
+```{r, eval = FALSE}
 merow_and_silander <- workflow(
@@
 
-```{r, eval=FALSE}
+```{r, eval = FALSE}
 merow_and_silander_spatial <- ReRunWorkflow(

### 2016-03-23 ms.Rmd
Add chunk names.

@@
 
-```{r, echo = FALSE, fig.height = 9, fig.cap = "The modular SDM structure encoded by a zoon workflow.
+```{r plotzoonoverview, echo = FALSE, fig.height = 9, fig.cap = "The modular SDM structure encoded by a zoon workflow.
 A) Flow diagram representing the module types.
@@
 
-```{r, eval = FALSE}
+```{r bioclimmodule, eval = FALSE}
 Bioclim <- function (extent = c(-180, 180, -90, 90),
@@
 
-```{r, eval = TRUE, echo = FALSE, cache = FALSE}
+```{r loadzoon, eval = TRUE, echo = FALSE, cache = FALSE}
 # Keep cache = FALSE. Not supposed to cache chunks with library()
@@
 
-```{r, eval = TRUE}
+```{r fengworkflow, eval = TRUE}
 Feng_Papes <- workflow(occurrence = SpOcc('Dasypus novemcinctus',
@@
 
-```{r, eval = FALSE}
-Feng_Papes_interactive <- ReRunWorkflow(
+```{r fengChangeWorkflow, eval = FALSE}
+Feng_Papes_interactive <- ChangeWorkflow(
     Feng_Papes,
@@
 
-```{r, eval = FALSE}
+```{r carolinamodule, eval = FALSE}
 CarolinaWrenPO <- function() {
@@
 
-```{r, eval = FALSE}
+```{r merowworkflow, eval = FALSE}
 merow_and_silander <- workflow(
@@
 
-```{r, eval = FALSE}
-merow_and_silander_spatial <- ReRunWorkflow(
+```{r mewoChangeWorkflow, eval = FALSE}
+merow_and_silander_spatial <- ChangeWorkflow(
     merow_and_silander,

### 2016-03-28 ms.Rmd
Escape asterisk that was being counted as a list.

@@
 5. We demonstrate `zoon` by recreating SDM analyses from two published research articles as zoon workflows, which readers can interrogate and extend.
+
+```{r knitrOpts, echo = FALSE, cache = FALSE, eval = TRUE}
+opts_chunk$set(fig.path = 'figs/')
+
+```
 
@@
 
-This function meets the minimal requirements of a *covariate* module (see Figure \ref{fig:workflows} Panel B); returning a `Raster*` object of the correct dimensions.
+This function meets the minimal requirements of a *covariate* module (see Figure \ref{fig:workflows} Panel B); returning a `Raster\*` object of the correct dimensions.
 
@@
 
-```{r fengworkflow, eval = TRUE}
+```{r fengworkflow, eval = TRUE, results = 'hide', message = FALSE}
 Feng_Papes <- workflow(occurrence = SpOcc('Dasypus novemcinctus',

### 2016-03-29 ms.Rmd
Change fengChangeWorkflow to eval = TRUE and reknit. ResponseCurve ouputs 10 or so figures. Need to decide what we want to present and how.

@@
                           process = Chain(Clean, # Remove rows w/ impossible, unlikely or missing lat lon.
+                                          MESSMask,  # Still to be implemented. mask out areas beyond extremes in train data
                                           Background(n = 1000, bias = 200), #background w/in 200km of occurrences
-                                          # MESSMask),  # Still to be implemented. mask out areas beyond extremes in train data
                                           Crossvalidate(k = 5)),
@@
 
-```{r fengChangeWorkflow, eval = FALSE}
+```{r fengChangeWorkflow, eval = TRUE}
 Feng_Papes_interactive <- ChangeWorkflow(
     Feng_Papes,
-    output = Chain(PrintMap, InteractiveMap, ResponseCurve))
+    output = Chain(PerformanceMeasures, PrintMap, InteractiveMap, ResponseCurve))
 ```

### 2016-03-30 ms.Rmd
Suppress message output and hide responsecurve figures.

@@
 ```{r knitrOpts, echo = FALSE, cache = FALSE, eval = TRUE}
-opts_chunk$set(fig.path = 'figs/')
+opts_chunk$set(fig.path = 'figs/', message = FALSE)
 
@@
 
-```{r fengChangeWorkflow, eval = TRUE}
+```{r fengChangeWorkflow, eval = TRUE, fig.show = 'hide'}
 Feng_Papes_interactive <- ChangeWorkflow(

### 2016-03-30 ms.Rmd
Make knitr create pdf and png files for each output image. The pdf output should default to using the pdfs, while the md output will use the pngs.
Clean up code a bit.

@@
 ```{r knitrOpts, echo = FALSE, cache = FALSE, eval = TRUE}
-opts_chunk$set(fig.path = 'figs/', message = FALSE)
+opts_chunk$set(fig.path = 'figs/', message = FALSE, dev = c('pdf', 'png'))
 
@@
 ```{r fengworkflow, eval = TRUE, results = 'hide', message = FALSE}
-Feng_Papes <- workflow(occurrence = SpOcc('Dasypus novemcinctus',
+Feng_Papes <- 
+  workflow(occurrence = SpOcc('Dasypus novemcinctus',
                               extent = c(-130, -20, -60, 60)),
@@
                                 layers = c(1:4, 6, 9, 10, 12, 15)),
-                          process = Chain(Clean, # Remove rows w/ impossible, unlikely or missing lat lon.
-                                          MESSMask,  # Still to be implemented. mask out areas beyond extremes in train data
-                                          Background(n = 1000, bias = 200), #background w/in 200km of occurrences
+              process = Chain(Clean, 
+                              MESSMask,  
+                              Background(n = 1000, bias = 200), 
                               Crossvalidate(k = 5)),
@@
     Feng_Papes,
-    output = Chain(PerformanceMeasures, PrintMap, InteractiveMap, ResponseCurve))
+    output = Chain(PerformanceMeasures, 
+                   PrintMap, 
+                   InteractiveMap, 
+                   ResponseCurve))
 ```

### 2016-03-30 ms.Rmd
Wrap long error messages. But might mess up syntax highlighting.
More code tidying.

@@
     if(!(resolution %in% c(2.5, 5, 10))){
-      stop('only 2.5, 5 and 10 degree resolutions are supported currently')
+      stop('only 2.5, 5 and 10 degree resolutions 
+            are supported currently')
     }
@@
         !all(layers %in% 1:19)) {
-      stop ('layers must be a vector of integers between 1 and 19 indicating layers to obtain')
+      stop ('layers must be a vector of integers between 
+             1 and 19 indicating layers to obtain')
     }
@@
 ```{r fengChangeWorkflow, eval = TRUE, fig.show = 'hide'}
-Feng_Papes_interactive <- ChangeWorkflow(
-    Feng_Papes,
+Feng_Papes_interactive <- 
+  ChangeWorkflow(workflow = Feng_Papes,
                    output = Chain(PerformanceMeasures, 

### 2016-03-30 ms.Rmd
Add bioclim reference.

@@
 Modules are simply R function definitions, with a set of required inputs and outputs.
-For example, the`Bioclim` module, which uses the `raster` R package [@raster] to download the widely used bioclim [] covariate layers in the correct format, is defined as:
+For example, the`Bioclim` module, which uses the `raster` R package [@raster] to download the widely used bioclim [@Hijmans2005] covariate layers in the correct format, is defined as:
 
@@
 
-@Feng2015 constructed a MaxEnt species distribution model for nine-banded armadillo in the USA using presence-only data on the species' current distribution, and the bioclim [] set of environmental correlates.
+@Feng2015 constructed a MaxEnt species distribution model for nine-banded armadillo in the USA using presence-only data on the species' current distribution, and the bioclim [@Hijmans2005] set of environmental correlates.
 This model was then used to predict areas in the Americas which may be suitable for the species to become established.
@@
 
-```{r fengworkflow, eval = TRUE, results = 'hide', message = FALSE}
+```{r fengworkCaptions, eval = TRUE, echo = FALSE}
+fengworkcap <- '
+  Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence (darker blue showing higher probability of occurrence).
+  White areas are masked due to being marine or due the MESS mask [@Elith2010]
+'
+
+```
+
+```{r fengworkflow, eval = TRUE, results = 'hide', message = FALSE, fig.cap = fengworkcap}
 Feng_Papes <- 

### 2016-03-30 ms.Rmd
Make knitr cache output.

@@
 ```{r knitrOpts, echo = FALSE, cache = FALSE, eval = TRUE}
-opts_chunk$set(fig.path = 'figs/', message = FALSE, dev = c('pdf', 'png'))
+opts_chunk$set(fig.path = 'figs/', message = FALSE, dev = c('pdf', 'png'), cache = TRUE)
 
@@
 # this is an awful hack and completely contingent on the screenshot size :/
-text(40, 980, labels = LETTERS[1],
+text(40, 940, labels = LETTERS[1],
       xpd = NA,

### 2016-03-31 ms.Rmd
Small plotting change.
Match capitilisation in workflow names.

@@
   Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence (darker blue showing higher probability of occurrence).
-  White areas are masked due to being marine or due the MESS mask [@Elith2010]
+  White areas are masked due to being marine in the marine environment or due to the MESS mask [@Elith2010]
 '
@@
 ```{r fengChangeWorkflow, eval = TRUE, fig.show = 'hide'}
-Feng_Papes_interactive <- 
+Feng_Papes_Interactive <- 
   ChangeWorkflow(workflow = Feng_Papes,
@@
 
-All of these have now been uploaded to the ZOON modules repository (under the names `CarolinaWrenPO`, `CarolinaWrenPA`, and `MaxLike`), though we diusplay the code required for the module `CarolinaWrenPO` here as an illustration of the minimal overhead required in writing `zoon` software modules:
+All of these have now been uploaded to the ZOON modules repository (under the names `CarolinaWrenPO`, `CarolinaWrenPA`, and `MaxLike`), though we display the code required for the module `CarolinaWrenPO` here as an illustration of the minimal overhead required in writing `zoon` software modules:
 
@@
 ```{r merowworkflow, eval = FALSE}
-merow_and_silander <- workflow(
+Merow_Silander <- workflow(
     occurrence = CarolinaWrenPO,
@@
     process = ...,
-    model = list(MaxEnt, MaxLike),
+    model = list(LogisticRegression, MaxEnt, MaxLike),
     output = ...)
@@
 ```{r mewoChangeWorkflow, eval = FALSE}
-merow_and_silander_spatial <- ChangeWorkflow(
+Merow_Silander_Spatial <- ChangeWorkflow(
     merow_and_silander,

### 2016-03-31 ms.Rmd
Split Merow into a plan and working workflow for now.

@@
 
+```{r merowpreamblel, eval = TRUE, echo = FALSE}
+
+# Load the data into the global environment
+data(carw, envir = environment())
+
+rl <- lapply(carw.data$raster.data, function(x) {
+   m <- matrix(x, nrow=carw.data$dim[1], ncol=carw.data$dim[2],
+               byrow=TRUE)
+   r <- raster(m)
+   extent(r) <- c(-124.83, -66.67, 24.61, 49.41)
+   r
+})
+
+# Stack and name them
+CarolinaRasters <- stack(rl[[1]], rl[[2]], rl[[3]], rl[[4]], rl[[5]], rl[[6]])
+names(CarolinaRasters) <- names(carw.data$raster.data)
+
+```
+
+```{r merowWorkingCopy, eval = TRUE}
+
+Merow_Silander <- workflow(
+    occurrence = Chain(LocalOccurrenceData('occurrenceType = 'presence' ),
+                       LocalOccurrenceData(occurrenceType = 'presence/absence', externalValidation = TRUE)),
+     covariate = LocalRasterData('CarolinaRasters'),
+       process = Chain(Transform(trans = 'square', replace = FALSE), 
+                       Background(n = 10000)),
+         model = list(LogisticRegression, MaxEnt, MaxLike),
+        output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap)
+
+```
+
 ```{r merowworkflow, eval = FALSE}
+
 Merow_Silander <- workflow(
-    occurrence = CarolinaWrenPO,
-    covariate = ...,
-    process = ...,
+    occurrence = Chain(LocalOccurrenceData(occurrenceType = 'presence' ),
+                       LocalOccurrenceData(occurrenceType = 'presence/absence', externalValidation = TRUE)),
+     covariate = LocalRasterData('CarolinaRasters'),
+       process = Chain(Transform(trans = 'square', replace = FALSE), 
+                       Background(n = 10000)),
          model = list(LogisticRegression, MaxEnt, MaxLike),
-    output = ...)
+        output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap, Coefficients)
 ```

### 2016-03-31 ms.Rmd
Set the seed.

@@
 
+set.seed(1633)
+
 ```
@@
                               MESSMask,  
-                              Background(n = 1000, bias = 200), 
+                              Background(n = 10000, bias = 200), 
                               Crossvalidate(k = 5)),

### 2016-05-17 ms.Rmd
A new draft for teh introduction. More to follow in the next few days. Hope I am getting the hang of working outside Word and Google docs :)

@@
 
-```{r knitrOpts, echo = FALSE, cache = FALSE, eval = TRUE}
+```{r knitrOpts, echo = FALSE, cache = FALSE, eval = FALSE}
 opts_chunk$set(fig.path = 'figs/', message = FALSE, dev = c('pdf', 'png'), cache = TRUE)
@@
 
-... reproducibility crisis in SDM.
+Like many areas of quantitative science, Species Distribution Modelling SDM has grown rapidly - fuelled by the availability of diverse modelling software and data.
+Over the last 20 years, SDM has developed into a very large research community and vast literature, and is now one of “the most widely-reviewed topics in the ecological literature” (Araújo & Peterson XXXX).
+Most of the software that enables SDM research, and that of other computational sub-disciplines, focuses on the analytical tasks of science (such as modelling and statistics) rather than tasks that are essential to the scientific enterprise, such as communication and producing reproducible outputs.
 
-<!-- from Greg, will need trimming down -->
-In 2008, @Beale2008 suggested that “we are currently unable to build useful distribution models for many species”. 
-Using null models, @Beale2008 found that chance associations between climate and species’ distributions could account for the modelled relationships which they then suggested undermined the rationale for SDM.
-Amongst the replies to this article was a re-analysis by @Araujo2009 that disputed this finding. 
-@Beale2008 put the alternative results to differences in distribution data, climate data and spatial domain, but emphasised the use of different null model distributions. 
-Whilst some code was provided in the original publication by @Beale2008, the following papers did not. 
-The resolution of this very important discussion was hindered by the lack of data and code sharing.
+This has presented computational science with a very modern version of the reproducibility problem as computational research is largely inaccessible ().
+If research materials and outputs are shared, they are rarely in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g.
+Fig Share, Git Hub).
+The ZOӦN R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
+The module-workflow structure offers a format by which scientists can more easily to create and share components of their analysis, and then access, modify, reuse and combine the components of others (see below and Figure 1).
 
-The second example comes from the study by @Elith2006, that was exceptional in its scope by comparing “16 modelling methods over 226 species from 6 regions of the world”. 
-This paper provided benchmarks that both figured highly in many researchers selection of method (and software). 
-However, as the data and code is not available this much needed set of benchmarks cannot be updated by the community. 
-For example, when a new modelling method is produced (e.g. GRAF, Golding…) authors cannot compare the outcomes to transparent set of community contributed modelling benchmarks and standards. 
-In this case, the lack of data and code sharing hinder the legacy of one of the key SDM studies to date.
+To date, the lack of sharing has reduced the capacity for innovation and validation in SDM, as well as its capacity for self-correction ().
+For example, a fundamental dispute over the ability of SDM to detect environmental associations (Beale et al.
+2008, Araujou et al.
+2009, Beale et al.
+2009) was left unresolved as only code in the original publication was published (Beale et al.
+2008).
+Without the resources being made open, the critical differences in these analysis are assumed rather than proven.
 
-Thirdly, @Soininen2014 “assembled over 4900 published AUC (area under the curve of a receiver operating characteristic plot) values” to examine the predictability of SDM in relation to species traits.
-Having conducted this labour intensive review they suggested that researchers should “give precise reports of model calibration and evaluation methods” to enable subsequent meta-analyses. 
-Relatedly, the methodological review of @Yackulic2013 investigated the use of the MAXENT algorithm in 108 articles, and have a similar concern regarding the “transparency and usefulness” in the reporting of SDM.
-In these studies there is little to no opportunity afforded by the original studies to investigate the sensitivity of the conclusion to choices in model building and data selection. 
-These studies point to a far wider reduction in scientific productivity due to reporting of SDM as ‘closed knowledge’.
+A second example comes from Elith et al.
+(2006), who reported a study of exceptional scope(“16 modelling methods over 226 species from 6 regions of the world”).
+The modelling benchmarks that were reported have been highly influential, yet cannot be updated as the code and data were not shared.
+The community cannot update or modify the benchmarks when new modelling method is produced (e.g.
+GRAF, Golding…).
+A third example comes from meta-analyses () and methodological reviews (), which require laborious searches of publications in the absence of shared code and modelling outputs.
+These findings cannot be turned into reanalyses that would yield truly informative methodological conclusions.
 
@@
 Each are crucial examples of the barriers to this science.
+
 Science has to be this way and written statements of the modelling carried out do not facilitate a productive science. 

### 2016-05-20 ms.Rmd
cleaning out some bugs in the doc

@@
 ```{r knitrOpts, echo = FALSE, cache = FALSE, eval = FALSE}
-opts_chunk$set(fig.path = 'figs/', message = FALSE, dev = c('pdf', 'png'), cache = TRUE)
+# set up knitr options
+opts_chunk$set(fig.path = 'figs/',
+               message = FALSE,
+               dev = c('pdf', 'png'),
+               cache = TRUE)
 
+```
+
+```{r raster_dir, echo = FALSE, cache = FALSE, eval = TRUE}
+# set a place for the rasters to be downloaded to (to stop them being grabbed
+# again each time)
+raster_dir <- './raster_data'
+if (!dir.exists(raster_dir)) dir.create(raster_dir)
+options("rasterDataDir" = raster_dir)
 ```
@@
 Fig Share, Git Hub).
-The ZOӦN R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
+The ZOON R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
 The module-workflow structure offers a format by which scientists can more easily to create and share components of their analysis, and then access, modify, reuse and combine the components of others (see below and Figure 1).
@@
 
-![The interactive map module overlay the raw data and predicted distribution, allowing users to interactively explore their results.\label{fig:interactive}](./figs/interactive_map.png)
+![The interactive map module overlays the raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results.\label{fig:interactive}](./figs/interactive_map.png)
 
@@
 # Load the data into the global environment
-data(carw, envir = environment())
+data(carw, envir = environment(), package = 'maxlike')
 
@@
 
-```{r merowWorkingCopy, eval = TRUE}
+```{r merowWorkingCopy, eval = FALSE}
 
 Merow_Silander <- workflow(
-    occurrence = Chain(LocalOccurrenceData('occurrenceType = 'presence' ),
-                       LocalOccurrenceData(occurrenceType = 'presence/absence', externalValidation = TRUE)),
-     covariate = LocalRasterData('CarolinaRasters'),
-       process = Chain(Transform(trans = 'square', replace = FALSE), 
+    occurrence = Chain(LocalOccurrenceData(occurrenceType = 'presence'),
+                       LocalOccurrenceData(occurrenceType = 'presence/absence',
+                                           externalValidation = TRUE)),
+     covariate = LocalRaster('CarolinaRasters'),
+       process = Chain(Transform(trans = 'square',
+                                 replace = FALSE), 
                        Background(n = 10000)),
          model = list(LogisticRegression, MaxEnt, MaxLike),
-        output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap)
+        output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap))
 

### 2016-05-20 ms.Rmd
fixed figure legend that was screwing the Rmd display in GitHub

@@
 
-```{r plotzoonoverview, echo = FALSE, fig.height = 9, fig.cap = "The modular SDM structure encoded by a zoon workflow.
+```{r plotzoonoverviewcaptions, eval = TRUE, echo = FALSE}
+pzocap <- "
+The modular SDM structure encoded by a zoon workflow.
 A) Flow diagram representing the module types.
@@
 C) Chaining and listing modules of the same type.
-D) The structure of a workflow object.\\label{fig:workflows}"}
+D) The structure of a workflow object.\\label{fig:workflows}
+"
+```
+
+```{r plotzoonoverview, echo = FALSE, fig.height = 9, fig.cap = pzocap}
 
@@
 '
-
 ```

### 2016-05-20 ms.Rmd
trying another way of rendering the figures

@@
                message = FALSE,
-               dev = c('pdf', 'png'),
+               dev = c('png'),
                cache = TRUE)
@@
 
-```{r plotzoonoverview, echo = FALSE, fig.height = 9, fig.cap = pzocap}
+```{r plotzoonoverview, echo = FALSE, fig.height = 9, dpi = 300, fig.cap = pzocap}
 
@@
 
-```{r fengworkCaptions, eval = TRUE, echo = FALSE}
-fengworkcap <- '
-  Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence (darker blue showing higher probability of occurrence).
-  White areas are masked due to being marine in the marine environment or due to the MESS mask [@Elith2010]
-'
-```
-
-```{r fengworkflow, eval = TRUE, results = 'hide', message = FALSE, fig.cap = fengworkcap}
+```{r fengworkflow, eval = TRUE, results = 'hide', fig.show = 'hide', message = FALSE}
 Feng_Papes <- 
@@
 ```
+
+![Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence (darker blue showing higher probability of occurrence).
+White areas are masked due to being marine in the marine environment or due to the MESS mask.\label{fig:fengmap](./figs/fengworkflow-1.png)
 
@@
 
-```{r mewoChangeWorkflow, eval = FALSE}
+```{r merowChangeWorkflow, eval = FALSE}
 Merow_Silander_Spatial <- ChangeWorkflow(

### 2016-05-20 ms.Rmd
resolving figure legend issues & adding higher resolution figures

@@
 
-```{r knitrOpts, echo = FALSE, cache = FALSE, eval = FALSE}
+```{r knitrOpts, echo = FALSE, cache = FALSE, eval = TRUE}
 # set up knitr options
-opts_chunk$set(fig.path = 'figs/',
+knitr::opts_chunk$set(fig.path = 'figs/',
                message = FALSE,
@@
                cache = TRUE)
-
 ```
@@
 
-```{r plotzoonoverviewcaptions, eval = TRUE, echo = FALSE}
-pzocap <- "
-The modular SDM structure encoded by a zoon workflow.
-A) Flow diagram representing the module types.
-B) The required inputs and outputs for each module types (full details given in the `zoon` vignette 'Building a module').
-C) Chaining and listing modules of the same type.
-D) The structure of a workflow object.\\label{fig:workflows}
-"
-```
-
-```{r plotzoonoverview, echo = FALSE, fig.height = 9, dpi = 300, fig.cap = pzocap}
+```{r plotzoonoverview, echo = FALSE, fig.show = 'hide', fig.height = 9, dpi = 300}
 
@@
 ```
+![The modular SDM structure encoded by a zoon workflow.
+A) Flow diagram representing the module types.
+B) The required inputs and outputs for each module types (full details given in the `zoon` vignette 'Building a module').
+C) Chaining and listing modules of the same type.
+D) The structure of a workflow object.\label{fig:workflows}](./figs/plotzoonoverview-1.png)
 
@@
 
-```{r fengworkflow, eval = TRUE, results = 'hide', fig.show = 'hide', message = FALSE}
+```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = 'hide', message = FALSE}
 Feng_Papes <- 
@@
 ![Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence (darker blue showing higher probability of occurrence).
-White areas are masked due to being marine in the marine environment or due to the MESS mask.\label{fig:fengmap](./figs/fengworkflow-1.png)
+White areas are masked due to being marine in the marine environment or due to the MESS mask.\label{fig:fengmap}](./figs/fengworkflow-1.png)
 

### 2016-05-20 ms.Rmd
more tidying up of chunks, and putting Greg's sentences on newlines

@@
                message = FALSE,
+               warning = FALSE,
                dev = c('png'),
@@
 
-SDM boom & growth of methods literature [@Barbosa2015]
-
-Like many areas of quantitative science, Species Distribution Modelling SDM has grown rapidly - fuelled by the availability of diverse modelling software and data.
+Like many areas of quantitative science, Species Distribution Modelling SDM has grown rapidly [@Barbosa2015] - fuelled by the availability of diverse modelling software and data.
 Over the last 20 years, SDM has developed into a very large research community and vast literature, and is now one of “the most widely-reviewed topics in the ecological literature” (Araújo & Peterson XXXX).
@@
 If research materials and outputs are shared, they are rarely in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g.
-Fig Share, Git Hub).
+Figshare, GitHub).
 The ZOON R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
@@
 
-A second example comes from Elith et al.
-(2006), who reported a study of exceptional scope(“16 modelling methods over 226 species from 6 regions of the world”).
+A second example comes from Elith et al. (2006), who reported a study of exceptional scope(“16 modelling methods over 226 species from 6 regions of the world”).
 The modelling benchmarks that were reported have been highly influential, yet cannot be updated as the code and data were not shared.
-The community cannot update or modify the benchmarks when new modelling method is produced (e.g.
-GRAF, Golding…).
+The community cannot update or modify the benchmarks when new modelling method is produced (e.g. GRAF, Golding…).
 A third example comes from meta-analyses () and methodological reviews (), which require laborious searches of publications in the absence of shared code and modelling outputs.
@@
 Modules are simply R function definitions, with a set of required inputs and outputs.
-For example, the`Bioclim` module, which uses the `raster` R package [@raster] to download the widely used bioclim [@Hijmans2005] covariate layers in the correct format, is defined as:
+For example the`Bioclim` module, which uses the `raster` R package [@raster] to download the widely used bioclim [@Hijmans2005] covariate layers in the correct format, is (excluding some error-handling code for brevity) defined as:
 
@@
                      layers = 1:19) {
-
-    if(!(resolution %in% c(2.5, 5, 10))){
-      stop('only 2.5, 5 and 10 degree resolutions 
-            are supported currently')
-    }
-
-    if (length(layers) < 1 |
-        length(layers) > 19 |
-        !all(layers %in% 1:19)) {
-      stop ('layers must be a vector of integers between 
-             1 and 19 indicating layers to obtain')
-    }
-
-    stopifnot(length(extent) == 4)
-    stopifnot(all(is.numeric(extent)))
-
     world <- getData('worldclim', var = 'bio', res = resolution)
@@
 
-```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = 'hide', message = FALSE}
+```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = 'hide', results = 'hide', message = FALSE}
 Feng_Papes <- 

### 2016-05-20 ms.Rmd

@@
 author: "Nick Golding, Tom August, Tim C.D. Lucas, Emiel Van Loon & Greg McInerny"
-date: "21 March 2016"
 bibliography: zoon_app.bib
@@
 
-#plot the saved image in the first panel
+# plot the saved image in the first panel
 library(raster)
@@
                 model = MaxEnt,
-               output = Chain(PerformanceMeasures, PrintMap))
+               output = PrintMap)
 
@@
 
-![Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence (darker blue showing higher probability of occurrence).
-White areas are masked due to being marine in the marine environment or due to the MESS mask.\label{fig:fengmap}](./figs/fengworkflow-1.png)
+![Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap` module in the workflow `Feng_Papes`.
+White areas are masked due to being in the sea or outside the MESS mask.
+\label{fig:fengmap}](./figs/fengworkflow-1.png)
 
@@
 ```{r fengChangeWorkflow, eval = TRUE, fig.show = 'hide'}
-Feng_Papes_Interactive <- 
+Feng_Papes_Update <- 
   ChangeWorkflow(workflow = Feng_Papes,
-                   output = Chain(PerformanceMeasures, 
-                                  PrintMap, 
-                                  InteractiveMap, 
-                                  ResponseCurve))
+                   output = Chain(InteractiveMap, 
+                                  ResponseCurve(cov = 1)))
 ```
 
-![The interactive map module overlays the raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results.\label{fig:interactive}](./figs/interactive_map.png)
+```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 300, fig.show = 'hide', fig.height = 3.2}
+# combine the printed map and first effect plot
+
+# load both images
+r_map <- brick('figs/interactive_map.png')
+r_resp <- brick('figs/fengChangeWorkflow-2.png')
+
+# set up layout so that heights are the same for both panels
+plot_height <- min(nrow(r_map), nrow(r_resp))
+
+# get rescaling factor for each
+rescale_map <- plot_height / nrow(r_map)
+rescale_resp <- plot_height / nrow(r_resp)
+
+# get widths
+width_map <- ncol(r_map) * rescale_map
+width_resp <- ncol(r_resp) * rescale_resp
+
+# set up layout (maximum total of 200 columns) with a gap in between
+# mar won't work for these
+gap <- 2
+widths <- c(width_map, width_resp)
+widths <- round(widths * ((200-gap) / sum(widths)))
+widths <- c(widths[1], gap, widths[2])
+mat <- matrix(rep(1:3, widths),
+              nrow = 1)
+layout(mat)
+
+# plot interactive map
+plotRGB(r_map, maxpixels = Inf)
+
+# add a panel letter
+mtext(text = 'A',
+        side = 3,
+        line = -3,
+        adj = 0)
+
+# gap
+plot.new()
+
+# plot the response curve
+plotRGB(r_resp, maxpixels = Inf)
+
+# add a panel letter
+mtext(text = 'B',
+        side = 3,
+        line = -3,
+        adj = 0)
+
+```
+
+![The outputs of the updated workflow object `Feng_Papes_Update`.
+A) A screenshot of the interactive map produced by the `InteractiveMap`, displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results.
+B) A response curve produced by the `ResponseCurve` for the first covariate, bio1.
+\label{fig:interactive}](./figs/combinefengupdate-1.png)
 

### 2016-05-21 ms.Rmd
tidying up and re-running higher res Feng_Papes workflow

@@
 
-<!--
-covariates Feng & Papes used:
-1    annual mean temperature
-2    mean diurnal range
-3    isothermality
-4    temperature seasonality
-6    minimum temperature of the coldest month
-9    mean temperature of the driest quarter
-10   mean temperature of the warmest quarter
-12   annual precipitation
-15   precipitation seasonality
--->
-
-<!---
-Notes:
-
--->
-
 ```{r loadzoon, eval = TRUE, echo = FALSE, cache = FALSE}
@@
 
-```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = 'hide', results = 'hide', message = FALSE}
+```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = 'hide', results = 'hide', message = FALSE, fig.height = 9, fig.width = 9}
 Feng_Papes <- 
@@
                output = PrintMap)
-
 ```

### 2016-05-21 ms.Rmd
fixing weird greying issue with response curve

@@
 r_map <- brick('figs/interactive_map.png')
-r_resp <- brick('figs/fengChangeWorkflow-2.png')
+r_resp <- raster('figs/fengChangeWorkflow-2.png')
+r_resp <- brick(r_resp, r_resp, r_resp)
 
@@
 # plot the response curve
-plotRGB(r_resp, maxpixels = Inf)
+plotRGB(r_resp, maxpixels = Inf,
+        scale = max(maxValue(r_resp)))
 

### 2016-05-21 ms.Rmd
incorporating new maxlike modules and removing excess text

@@
 
-All of these have now been uploaded to the ZOON modules repository (under the names `CarolinaWrenPO`, `CarolinaWrenPA`, and `MaxLike`), though we display the code required for the module `CarolinaWrenPO` here as an illustration of the minimal overhead required in writing `zoon` software modules:
-
-```{r carolinamodule, eval = FALSE}
-CarolinaWrenPO <- function() {
-
-  # load maxlike package (installing if it isn't in the library)
-  zoon::GetPackage('maxlike')
-
-  # load the dataset from the maxlike package
-  data(carw, envir = environment())
-  occ <- na.omit(carw.data$pa.data)
-
-  # keep only presence records
-  occ <- occ[occ$y == 1, ]
-
-  # build the occurrence dataset and return
-  data.frame(longitude = occ$Lon,
-             latitude = occ$Lat,
-             value = occ$y,
-             type = 'presence'),
-             fold = 1)
-
-}
-```
-
-This function meets the requirements of an *occurrence* module; providing as output a dataframe with the required columns and datatypes.
-
-We can run automated checks on this module
+All of these have now been uploaded to the ZOON modules repository under the names `CarolinaWrenPO`, `CarolinaWrenPA`, `CarolinaWrenRasters` and `MaxLike`.
 
@@
 
-```{r merowpreamblel, eval = TRUE, echo = FALSE}
-
-# Load the data into the global environment
-data(carw, envir = environment(), package = 'maxlike')
-
-rl <- lapply(carw.data$raster.data, function(x) {
-   m <- matrix(x, nrow=carw.data$dim[1], ncol=carw.data$dim[2],
-               byrow=TRUE)
-   r <- raster(m)
-   extent(r) <- c(-124.83, -66.67, 24.61, 49.41)
-   r
-})
-
-# Stack and name them
-CarolinaRasters <- stack(rl[[1]], rl[[2]], rl[[3]], rl[[4]], rl[[5]], rl[[6]])
-names(CarolinaRasters) <- names(carw.data$raster.data)
-
-```
-
-```{r merowWorkingCopy, eval = FALSE}
-
+```{r merowworkflow, eval = TRUE}
 Merow_Silander <- workflow(
-    occurrence = Chain(LocalOccurrenceData(occurrenceType = 'presence'),
-                       LocalOccurrenceData(occurrenceType = 'presence/absence',
-                                           externalValidation = TRUE)),
-     covariate = LocalRaster('CarolinaRasters'),
-       process = Chain(Transform(trans = 'square',
-                                 replace = FALSE), 
-                       Background(n = 10000)),
+    occurrence = CarolinaWrenPO,
+    covariate = CarolinaWrenRasters,
+    process = Background(n = 10000),
     model = list(LogisticRegression, MaxEnt, MaxLike),
-        output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap))
-
+    output = Chain(PerformanceMeasures, PrintMap))
 ```
@@
 Merow_Silander <- workflow(
-    occurrence = Chain(LocalOccurrenceData(occurrenceType = 'presence' ),
-                       LocalOccurrenceData(occurrenceType = 'presence/absence', externalValidation = TRUE)),
-     covariate = LocalRasterData('CarolinaRasters'),
-       process = Chain(Transform(trans = 'square', replace = FALSE), 
+    occurrence = CarolinaWrenPO,
+    covariate = CarolinaWrenRasters,
+    process = Chain(Transform(trans = 'square',
+                                 replace = FALSE), 
                        Background(n = 10000)),
     model = list(LogisticRegression, MaxEnt, MaxLike),
-        output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap, Coefficients)
+    output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap, Coefficients))
 ```

### 2016-05-22 figs/zoondiag.R
removing old figure 1 components

@@ -1,45 +1,0 @@
-# code to construct a simple diagram of the workflow layout
-# because the output is html, I can't find a programmatic way to convert to
-# a static image format for inclusion in the pdf.
-# Instead, I'm running this in RStudio, and manually saving as png
-
-library(DiagrammeR)
-
-grViz("
-      digraph DAG {
-      
-      # Intialization of graph attributes
-      graph [overlap = true]
-      
-      # Initialization of node attributes
-      node [
-        shape = box,
-        fontname = Helvetica,
-        color = DimGray,
-        type = box,
-        width = 1.2,
-        fixedsize = true,
-        style = filled,
-        penwidth = 0.5
-      ]
-      
-      # Node statements
-      occurrence [fillcolor = LightBlue]
-      covariate [fillcolor = LightGreen]
-      process [fillcolor = LightGray]
-      model [fillcolor = lightsalmon]
-      output [fillcolor = wheat]
-
-
-      # Initialization of edge attributes
-      edge [
-        color = DimGray,
-        rel = yields,
-        arrowsize = 0.8
-      ]
-      
-      # Edge statements
-      occurrence->process; covariate->process; process->model; model->output
-      
-      }
-      ")

### 2016-05-22 ms.Rmd

@@
 
-```{r plotzoonoverview, echo = FALSE, fig.show = 'hide', fig.height = 9, dpi = 300}
+<!--
+the figures for this panel are generated in powerpoint (I know, I know, but it's quick and easy) and saved in the figs folder.
+the powerpoint file itself is not version controlled, but can be access here: https://www.dropbox.com/s/afjpv8v6na9y56e/diagrams.pptx?dl=0
+-->
 
-# four panel plot
-par(mfrow = c(2, 2))
-
-# plot the saved image in the first panel
-library(raster)
-r <- brick('figs/workflow_basic.png')
-par(mar = c(1, 1, 2, 1))
-plotRGB(r, maxpixels = Inf)
-
-# add a panel letter
-# this is an awful hack and completely contingent on the screenshot size :/
-text(40, 940, labels = LETTERS[1],
-      xpd = NA,
-     cex = 1.2)
-
-par(mar = c(1, 1, 2, 1))
-# placeholder plots for each panel
-for (i in 2:4) {
-
-  # set up plotting window
-  plot.new()
-  plot.window(0:1, 0:1)
-
-  # grey rectangle (no background in vector pdf?)
-  rect(-1, -1, 2, 2, col = grey(0.9), border = NA)
-
-  # text over the top
-  text(0.5, 0.5, 'PLACEHOLDER',
-       col = grey(0.3),
-       cex = -13,
-       xpd = NA)
-
-  # add a panel letter
-  mtext(text = LETTERS[i],
-        side = 3,
-        line = 0.5,
-        adj = 0)  
-}
-```
 ![The modular SDM structure encoded by a zoon workflow.
-A) Flow diagram representing the module types.
-B) The required inputs and outputs for each module types (full details given in the `zoon` vignette 'Building a module').
-C) Chaining and listing modules of the same type.
-D) The structure of a workflow object.\label{fig:workflows}](./figs/plotzoonoverview-1.png)
+A) Flow diagram representing the five module types and intermediate objects: df - an occurrence dataframe; ras - a RasterStack object; mod - a ZoonModel object; ?
+- any user-defined inputs or outputs.
+B) The flow diagram implied by chaining modules of the same type.
+C) The flow diagram implied by listing modules of the same type.
+Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the `zoon` vignette 'Building a module'.
+\label{fig:workflows}](./figs/diagrams.png)
 
@@
 
-```{r merowworkflow, eval = TRUE}
+```{r merowworkflowtemp, eval = TRUE}
 Merow_Silander <- workflow(

### 2016-05-22 ms.Rmd
Re-organising the structure of the main section

@@
 # Introduction
-
-<!-- The Problem -->
 
@@
 
-This paper introduces version 0.4-22 of the `zoon` R package, which encodes the majority of this functionality.
+This paper introduces version 0.4-22 of the `zoon` R package, which can be dowloaded from CRAN and which encodes the majority of this functionality.
 First, we describe the modular structure of zoon workflows and demonstrate how completed `zoon` workflows can be shared, evaluated and extended.
@@
 
-<!-- The Solution: ZOOOOOON!  -->
 # The `zoon` R package
 
-Versions 0.4-22 and later of the `zoon` R package can be installed directly from CRAN.
+The `zoon` R package encodes SDM analyses as a simple workflow of five key steps, with each step encoded as one or more software modules.
+Users combine these modules via a call to the `workflow` function, which executes each module in turn, before returning a `zoonWorkflow` object - a shareable, extensible and fully reproducible documentation of the SDM analysis.
+Figure \ref{fig:workflows} illustrates the structure of a `zoon` workflow comprising the five module types and illustrates how multiple modules may be combined at each stage.
+The module types and module combination options are described in more detail below.
 
-### Constructing a `zoon` workflow
-
-The `zoon` R package encodes SDM analyses as a simple workflow of five steps: obtaining occurrence data, obtaining covariate data, pre-processing these data, fitting a model, and generating outputs.
-Each of these steps is carried out by one or more community-contributed software modules.
-For each of these steps, the user selects one or more 'modules' and combined them in a call to `workflow`.
-
-Figure \ref{fig:workflows} illustrates the structure of a `zoon` workflow.
-
-<!--
-the figures for this panel are generated in powerpoint (I know, I know, but it's quick and easy) and saved in the figs folder.
-the powerpoint file itself is not version controlled, but can be access here: https://www.dropbox.com/s/afjpv8v6na9y56e/diagrams.pptx?dl=0
--->
+<!— link to powerpoint doc: https://www.dropbox.com/s/afjpv8v6na9y56e/diagrams.pptx?dl=0-->
 
@@
 
+####
+
 #### Module types
 
-* *Occurrence* - Usually presence-absence data or presence-only data, though abundance data is also used.
-* *Covariates* - Predictor variables or covariates (typically environmental covariates) are required and the values of these covariates, at the locations of the occurrence data, must be extracted.
-* *Process* - Processes  applied to the occurrence and covariate data.
+The five module types, corresponding to the five key steps of an SDM anaysis are:
+
+<!— needs work—>
+
+- *Occurrence* - Usually presence-absence data or presence-only data, though abundance data is also used.
+- *Covariates* - Predictor variables or covariates (typically environmental covariates) are required and the values of these covariates, at the locations of the occurrence data, must be extracted.
+- *Process* - Processes  applied to the occurrence and covariate data.
 These processes include data cleaning, data thinning to account for spatial biases, feature selection using PCA or association tests and the splitting of data into training and test sets or cross validation folds.
-* *Model* - Once the data has been suitable manipulated a model is fitted to estimate the relationships between the covariates and occurrence data.
+- *Model* - Once the data has been suitable manipulated a model is fitted to estimate the relationships between the covariates and occurrence data.
 These models include simple statistical models such as GLMs as well as modern, flexible machine-learning methods such as MaxEnt and boosted regression trees.
-* *Output* - The predictive power and goodness of fit must be assessed and the model parameters or response curves must be examined.
+- *Output* - The predictive power and goodness of fit must be assessed and the model parameters or response curves must be examined.
 The model is likely to be used to predict species occurrence, either in the vicinity of the occurrence data or elsewhere, or into the past or future.
@@
 #### Lists and Chains
+
+<!— needs work—>
 
@@
 
-#### Crossvalidation
-
-`workflow` sensibly handles cross-validation.
-Process modules are used to split the data into a training and test set or into multiple folds so that each fold in turn will be held back and used as the test for the predictive ability of any fitted models.
-The same test/train splits will be used for each analysis if the multiple models or methods are being used so that model performance measures are comparable.
-Furthermore, the test/train splits are saved along with the workflow so that other researchers adding to a workflow can make fair comparisons. 
-After models are fitted to these subsets of the data, they are also fitted to the entire dataset as this will provide the best models for predictions from further new data.
-
 ### Inspecting, sharing and extending a `zoon` workflow
 
-Structure of workflow objects:
+A `zoonWorkflow` objectStructure of workflow objects:
 
@@
 
+#### Additional features
+
+##### Cross-validation
+
+<!— needs work—>
+
+`workflow` sensibly handles cross-validation.
+Process modules are used to split the data into a training and test set or into multiple folds so that each fold in turn will be held back and used as the test for the predictive ability of any fitted models.
+The same test/train splits will be used for each analysis if the multiple models or methods are being used so that model performance measures are comparable.
+Furthermore, the test/train splits are saved along with the workflow so that other researchers adding to a workflow can make fair comparisons. 
+After models are fitted to these subsets of the data, they are also fitted to the entire dataset as this will provide the best models for predictions from further new data.
+
+##### Sharing workflows
+
+`zoon` provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform [`figshare.com`]().
+`ZoonFigshare` takes a workflow object and some minimal metadata and uploads the workflow as an RData object to the user's figshare profile, for others to download, inspect and modify.
+
 ### Example Applications
 
-We demonstrate the `zoon` R package by recreating two SDM analyses from published research articles and extending them.
+We demonstrate the `zoon` R package by recreating two SDM analyses from published research articles as `zoon` workflows and showing how these workflow objects can easily be extended with new modules.
 Workflow objects created by these analyses can be accessed at [http://figshare.com/articles/zoon_applications_paper_workflows](http://figshare.com/articles/zoon_applications_paper_workflows).
 We encourage readers to download, interrogate and alter these workflows for themselves.
-Full code and metadata for all of the modules used in the examples below, can be found at [https://github.com/zoonproject/modules/R](https://github.com/zoonproject/modules/R)
+Full code and metadata for all of the modules used in the examples below can be found at [https://github.com/zoonproject/modules/R](https://github.com/zoonproject/modules/R)
 
@@
 
-```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = 'hide', results = 'hide', message = FALSE, fig.height = 9, fig.width = 9}
+```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = &#39;hide&#39;, results = &#39;hide&#39;, message = FALSE, fig.height = 9, fig.width = 9}
 Feng_Papes <- 
@@
 
-```{r fengChangeWorkflow, eval = TRUE, fig.show = 'hide'}
+```{r fengChangeWorkflow, eval = TRUE, fig.show = &#39;hide&#39;}
 Feng_Papes_Update <- 
@@
 
-```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 300, fig.show = 'hide', fig.height = 3.2}
+```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 300, fig.show = &#39;hide&#39;, fig.height = 3.2}
 # combine the printed map and first effect plot
@@
 
+<!— point to github development site—>
+
 ### References

### 2016-05-22 ms.Rmd
tidying up formatting error, switching BioClim to PPM example

@@
 Modules are simply R function definitions, with a set of required inputs and outputs.
-For example the`Bioclim` module, which uses the `raster` R package [@raster] to download the widely used bioclim [@Hijmans2005] covariate layers in the correct format, is (excluding some error-handling code for brevity) defined as:
-
-```{r bioclimmodule, eval = FALSE}
-Bioclim <- function (extent = c(-180, 180, -90, 90),
-                     resolution = 10,
-                     layers = 1:19) {
-    world <- getData('worldclim', var = 'bio', res = resolution)
-    world <- world[[layers]]
-    cropped <- crop(world, extent(extent))
-    return (cropped)
-  }
-```
-
-This function meets the minimal requirements of a *covariate* module (see Figure \ref{fig:workflows} Panel B); returning a `Raster\*` object of the correct dimensions.
 
@@
 
-```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = &#39;hide&#39;, results = &#39;hide&#39;, message = FALSE, fig.height = 9, fig.width = 9}
+```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = "hide", results = "hide", message = FALSE, fig.height = 9, fig.width = 9}
 Feng_Papes <- 
@@
 
-```{r fengChangeWorkflow, eval = TRUE, fig.show = &#39;hide&#39;}
+```{r fengChangeWorkflow, eval = TRUE, fig.show = "hide"}
 Feng_Papes_Update <- 
@@
 
-```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 300, fig.show = &#39;hide&#39;, fig.height = 3.2}
+```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 300, fig.show = "hide", fig.height = 3.2}
 # combine the printed map and first effect plot
@@
 
-#### Example 2. Evaluating MaxLike
+#### Example 2. Building a point process model module
+
+Recent SDM research has identified an equivalence between MaxEnt and Poisson point process models (PPMs) [] and noted that PPMs can be fitted using Poisson modelling software [Warton & Renner, Renner et al.].
+This means that other modelling approaches can be fitted as PPMs, after a bit of data preparation.
+We illustrate how to build a `zoon` model module by defining a module to fit a PPM using a Generalized Additive Model (GAM).
+
+The code for our model is as follows (see Renner et al., 2014 for details of this approach):
+
+```{r bioclimmodule, eval = FALSE}
+PPM_GAM <- function () {
+    # to do...
+  }
+```
+
+The only complicated aspect of this code is construction of a `zoonModel` object, which contains code to make predictions from the model in a standardised way.
+The other module types do not require this step, so are simpler to implement.
+The code also meets the minimal requirements of a *model* module, accepting the required argument `.df` and returning the `zoonModel` object.
+
+To convert this code into a ZOON module, we need to write it to a standalone file with the required metadata.
+the `zoon` function `BuildModule` helps with this step, and can also run checks to make sure we got everything right:
+
+```{r buildmodule, eval = FALSE} 
+BuildModule(PPM_GAM, ...)
+```
+
+This module can now be used in a `zoon` workflow and shared with others. To upload the module to the `zoon` modules repository, [TBD].
+
+#### Example 3. Evaluating MaxLike
 

### 2016-05-23 ms.Rmd
switching to spThin for module example, combining figures for two Feng workflows into one multi-panel figure

@@
 
+##### Accessing module documentation
+
+<!-- to do -->
+
 ##### Cross-validation
@@
 
-![Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap` module in the workflow `Feng_Papes`.
-White areas are masked due to being in the sea or outside the MESS mask.
-\label{fig:fengmap}](./figs/fengworkflow-1.png)
-
+This module plots a static map of the predicted distribution, and the underlying data.
+This map is shown in \ref{fig:feng_papes_plots}A.
 The resulting workflow contains all the code required to re-run the workflow, the input data and the results of executing each module.
@@
 Next, we update the workflow to produce an interactive map enabling anyone to inspect the data and predictions on a zoomable map, and to inspect the response curves of the fitted model.
-These outputs are shown in Figure \ref{fig:interactive}.
+These outputs are shown in Figure This map is shown in \ref{fig:feng_papes_plots}, panels B and C.
 
-```{r fengChangeWorkflow, eval = TRUE, fig.show = "hide"}
+```{r fengChangeWorkflow, eval = TRUE, fig.show = "hide", fig.height = 9}
 Feng_Papes_Update <- 
@@
 
-```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 300, fig.show = "hide", fig.height = 3.2}
-# combine the printed map and first effect plot
+```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 3.2}
+# combine the static and interactive maps and first effect plot
 
 # load both images
+r_map1 <- brick('figs/fengworkflow-1.png')
 r_map <- brick('figs/interactive_map.png')
-r_resp <- raster('figs/fengChangeWorkflow-2.png')
+r_resp <- raster('figs/fengChangeWorkflow-1.png')
 r_resp <- brick(r_resp, r_resp, r_resp)
@@
 # set up layout so that heights are the same for both panels
-plot_height <- min(nrow(r_map), nrow(r_resp))
+plot_height <- min(nrow(r_map1), nrow(r_map), nrow(r_resp))
 
 # get rescaling factor for each
+rescale_map1 <- plot_height / nrow(r_map1)
 rescale_map <- plot_height / nrow(r_map)
@@
 # get widths
+width_map1 <- ncol(r_map1) * rescale_map1
 width_map <- ncol(r_map) * rescale_map
@@
 gap <- 2
-widths <- c(width_map, width_resp)
-widths <- round(widths * ((200-gap) / sum(widths)))
-widths <- c(widths[1], gap, widths[2])
-mat <- matrix(rep(1:3, widths),
+widths <- c(width_map1, width_map, width_resp)
+widths <- round(widths * ((200 - (2 * gap)) / sum(widths)))
+widths <- c(widths[1], gap, widths[2], gap, widths[3])
+mat <- matrix(rep(1:5, widths),
               nrow = 1)
 layout(mat)
+
+# plot statis map
+plotRGB(r_map1, maxpixels = Inf)
+
+# add a panel letter
+mtext(text = 'A',
+        side = 3,
+        line = -3,
+        adj = 0)
+
+# gap
+plot.new()
 
@@
 # add a panel letter
-mtext(text = 'A',
+mtext(text = 'B',
         side = 3,
@@
 # add a panel letter
-mtext(text = 'B',
+mtext(text = 'C',
         side = 3,
@@
 
-![The outputs of the updated workflow object `Feng_Papes_Update`.
-A) A screenshot of the interactive map produced by the `InteractiveMap`, displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results.
-B) A response curve produced by the `ResponseCurve` for the first covariate, bio1.
-\label{fig:interactive}](./figs/combinefengupdate-1.png)
+![Outputs of the workflow objects `Feng_Papes` and `Feng_Papes_Update`.
+A) Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap` module in the workflow `Feng_Papes`.
+White areas are masked due to being in the sea or outside the MESS mask.
+B) A screenshot of the interactive map produced by the `InteractiveMap` moduels in the workflow `Feng_Papes_Update`, displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results.
+C) A response curve produced by the `ResponseCurve` module for the first covariate, bio1 in the workflow `Feng_Papes_Update`.
+\label{fig:feng_papes_plots}](./figs/combinefengupdate-1.png)
 
@@
 
-#### Example 2. Building a point process model module
+#### Example 2. Building a spatial thinning module
 
-Recent SDM research has identified an equivalence between MaxEnt and Poisson point process models (PPMs) [] and noted that PPMs can be fitted using Poisson modelling software [Warton & Renner, Renner et al.].
-This means that other modelling approaches can be fitted as PPMs, after a bit of data preparation.
-We illustrate how to build a `zoon` model module by defining a module to fit a PPM using a Generalized Additive Model (GAM).
+@Aiello-Lammens2015 proposed an approach for dealing with spatial sampling bias in presence-only data, by 'thinning' the presence data, and provide an R package `spThin` to implement their procedure [@spThin].
+We can incorporate this approach in a workflow, by defining a simple *process* module that adapts the `zoon` data into the spThin format, uses this package to apply the algorithm, and converts data back into `zoon`'s expected format again:
 
-The code for our model is as follows (see Renner et al., 2014 for details of this approach):
+```{r spthinmodule, eval = TRUE}
+spThin <- function (.data, thin = 50) {
 
-```{r bioclimmodule, eval = FALSE}
-PPM_GAM <- function () {
-    # to do...
+  # check these are presence-background data
+  stopifnot(all(.data$df$type %in% c('presence', 'background')))
+
+  # install & load the package
+  zoon::GetPackage('spThin')
+
+  # get dataframe & index to presence data
+  df <- .data$df
+  pres_idx <- which(df$type == 'presence')
+
+  # prepare presence data subset and apply thinning
+  sub_df <- data.frame(LAT = df$latitude[pres_idx],
+                       LONG = df$longitude[pres_idx],
+                       SPEC = NA)
+  th <- thin(loc.data = sub_df,
+             thin.par = thin,
+             reps = 1,
+             locs.thinned.list.return = TRUE,
+             write.files = FALSE,
+             write.log.file = FALSE)
+
+  # get index to rows in sub_df, update the full dataset and return
+  pres_keep_idx <- rownames(th[[1]])
+  .data$df <- rbind(df[pres_idx[pres_keep_idx], ],
+                    df[-pres_idx, ])
+  return (.data)
 }
@@
 
-The only complicated aspect of this code is construction of a `zoonModel` object, which contains code to make predictions from the model in a standardised way.
-The other module types do not require this step, so are simpler to implement.
-The code also meets the minimal requirements of a *model* module, accepting the required argument `.df` and returning the `zoonModel` object.
+To convert this code into a ZOON module, we need to write it to a standalone file (named `spThin.R`) with the metadata required to build the module documentation.
+The `zoon` function `BuildModule()` helps with this step, and can also run checks to make sure we got everything right:
 
-To convert this code into a ZOON module, we need to write it to a standalone file with the required metadata.
-the `zoon` function `BuildModule` helps with this step, and can also run checks to make sure we got everything right:
-
-```{r buildmodule, eval = FALSE} 
-BuildModule(PPM_GAM, ...)
+```{r buildmodule, eval = TRUE, results = 'hide'} 
+BuildModule(object = spThin,
+            type = 'process',
+            title = 'Spatial thinning of Presence-only Data',
+            description = paste('Apply the stochastic spatial thinning',
+                                'algorithm implemented in the spThin',
+                                'package to presence data in a',
+                                'presence-background dataset'),
+            details = paste('Full details of the algorithm are available in',
+                            'the open-access article by Aiello-Lammens',
+                            'et al. (2015): dx.doi.org/10.1111/ecog.01132'),
+            author = 'ZOON Developers',
+            email = 'zoonproject@gmail.com',
+            paras = list(thin = paste('Thinning parameter - the required',
+                         'minimum distance (in kilometres) between points',
+                         'after applying the thinning procedure')),
+            dataType = 'presence-only',
+            check = TRUE)
 ```
 
-This module can now be used in a `zoon` workflow and shared with others. To upload the module to the `zoon` modules repository, [TBD].
+This module can now be shared so that others can use it in their ZOON workflows.
+The bexst way of sharing it is via the `zoon` modules repository.
+Modules can uploaded there, via the module submission system at [URL TBD].
 

### 2016-05-24 ms.Rmd

@@
                 model = MaxEnt,
-               output = PrintMap)
+               output = PrintMap(points = FALSE))
 ```
@@
     model = list(LogisticRegression, MaxEnt, MaxLike),
-    output = Chain(PerformanceMeasures, PrintMap))
+    output = Chain(PerformanceMeasures, PrintMap(points = FALSE)))
 ```
@@
     model = list(LogisticRegression, MaxEnt, MaxLike),
-    output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap, Coefficients))
+    output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap(points = FALSE), Coefficients))
 ```

### 2016-09-13 ms.Rmd
spThin module code updated so that it passes tha package checks. There was an issue with indexing in the last lines and I think NAs might have been causing an issue too.

@@
 Unfortunately the latter data have not been made publically available, so here we use only data from GBIF.
-If the the additional data had been made available it would be straightforward to incorporate them, for example using the `LocalOccurrenceData` module.
+If the additional data had been made available it would be straightforward to incorporate them, for example using the `LocalOccurrenceData` module.
 
@@
   # get dataframe & index to presence data
-  df <- .data$df
+  df <- na.omit(.data$df)
   pres_idx <- which(df$type == 'presence')
@@
   pres_keep_idx <- rownames(th[[1]])
-  .data$df <- rbind(df[pres_idx[pres_keep_idx], ],
+  .data$df <- rbind(df[pres_idx,][pres_keep_idx, ],
                     df[-pres_idx, ])
@@
 
-```{r buildmodule, eval = TRUE, results = 'hide'} 
+```{r buildmodule, eval = TRUE, results = 'hide', fig.keep = 'none'} 
 BuildModule(object = spThin,
@@
 This module can now be shared so that others can use it in their ZOON workflows.
-The bexst way of sharing it is via the `zoon` modules repository.
+The best way of sharing it is via the `zoon` modules repository.
 Modules can uploaded there, via the module submission system at [URL TBD].

### 2016-11-01 ms.Rmd
add intro text

--- 2016-05-24 ms.Rmd
@@
 
-Like many areas of quantitative science, Species Distribution Modelling SDM has grown rapidly [@Barbosa2015] - fuelled by the availability of diverse modelling software and data.
+Like many areas of quantitative science Species Distribution Modelling (SDM) has grown rapidly [@Barbosa2015] fueled by the availability of diverse modelling software and data.
 Over the last 20 years, SDM has developed into a very large research community and vast literature, and is now one of “the most widely-reviewed topics in the ecological literature” (Araújo & Peterson XXXX).
-Most of the software that enables SDM research, and that of other computational sub-disciplines, focuses on the analytical tasks of science (such as modelling and statistics) rather than tasks that are essential to the scientific enterprise, such as communication and producing reproducible outputs.
+Most SDM software, and that enabling other other computational sub-disciplines, focus on  analytical tasks (such as modelling and statistics) rather than enabling scientists to produce reproducible and modifiable analyses that are essential to the scientific enterprise.
 
-This has presented computational science with a very modern version of the reproducibility problem as computational research is largely inaccessible ().
-If research materials and outputs are shared, they are rarely in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g.
+Computational sciences are then presented with a modern version of the reproducibility problem as computational research is largely inaccessible let alone reusable.
+If research materials and outputs are shared, they are rarely disseminated in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g.
 Figshare, GitHub).
 The ZOON R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
-The module-workflow structure offers a format by which scientists can more easily to create and share components of their analysis, and then access, modify, reuse and combine the components of others (see below and Figure 1).
+The module-workflow structure enables scientists to more easily to create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure 1).
 
-To date, the lack of sharing has reduced the capacity for innovation and validation in SDM, as well as its capacity for self-correction ().
+A lack of sharing reduces the capacity for innovation and validation in SDM, as well as its capacity to be a "self-correcting process" [@roysoc_report].
 For example, a fundamental dispute over the ability of SDM to detect environmental associations (Beale et al.
@@
 2009, Beale et al.
-2009) was left unresolved as only code in the original publication was published (Beale et al.
+2009) was left unresolved as only the original publication shared their analytical code (Beale et al.
 2008).
-Without the resources being made open, the critical differences in these analysis are assumed rather than proven.
+Without the resources being made open the critical differences in these analytical steps are assumed rather than proven.
 
 A second example comes from Elith et al. (2006), who reported a study of exceptional scope(“16 modelling methods over 226 species from 6 regions of the world”).
-The modelling benchmarks that were reported have been highly influential, yet cannot be updated as the code and data were not shared.
-The community cannot update or modify the benchmarks when new modelling method is produced (e.g. GRAF, Golding…).
-A third example comes from meta-analyses () and methodological reviews (), which require laborious searches of publications in the absence of shared code and modelling outputs.
-These findings cannot be turned into reanalyses that would yield truly informative methodological conclusions.
+The modelling benchmarks that were reported have been highly influential, yet cannot be updated or modified when new modelling methods are produced as the code and data were not shared.
+A third example comes from meta-analyses [] and methodological reviews [@Yackulic?], which rely on laborious searches through text publications in the absence of shared code and modelling outputs.
+SDM then has limited abilities to derive truly informative methodological conclusions that can be acted upon.
 
-These three scenarios are illustrative of some of the need for the data and modelling aspects of SDM to become more repeatable, reproducible, accessible, modifiable and updateable.
-Each are crucial examples of the barriers to this science.
+These three examples broadly illustrate why the data and models underpinning SDM need to become more repeatable, reproducible, accessible, modifiable and updateable.
+A productive science relies on the accumulation of knowledge that can be reused and appropriated to advance understanding and resolve methodological issues.
+Thus while many software are pitched as being ‘easy to use’ or ‘user friendly’, this is infrequently the case for the secondary user.
 
-Science has to be this way and written statements of the modelling carried out do not facilitate a productive science. 
-SDM, as a science, is not extensible given current working practices.
+This paper introduces version 0.4-22 of the `ZOON` R package, which enables users to address issues of compatibility, accessibility, sharing and reproducibility for their SDM analyses.
+Firstly, we describe the modular structure of ZOON workflows and demonstrate how completed workflows can be shared, evaluated and extended.
+We then illustrate these concepts, and the potential contributions of ZOON to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.
+To assist users a number of tutorials are provided on the github repository and as vignettes distributed with zoon – including  guides on creating workflows and modules, as well as full technical details for module developers.
 
-### Solutions to these problems
-
-The core, soluble problems are:
-
-Difficult for new SDM users to access latest methods.
-
-Difficult for anyone to compare methods effectively.
-
-Equally difficult for methods developers to disseminate their advances.
-
-<!-- Past work towards this goal & where it falls short - the USP of ZOON and requirements -->
 ### Reproducible software projects

### 2016-11-01 ms.Rmd
add software review text

@@
 
-Most of the SDM users replying to a survey in 2015 (Ahmed et al. 2015) were using either MaxEnt or R as the first choice software for their analyses. 
-The software market is much diverse, however, including a variety of software developed specifically for SDM (e.g.
-MaxEnt (), the BioMod () and dismo () packages for R, SDMtoolbox (Brown 2014), openModeller (), BioEnsembles (), ModEco ()) and software that has been appropriated for SDM studies, such as statistical software (R (), WinBugs (), OpenBugs (), Python()) and other less general software  (Domain (), Canoco (), MARS()).
+Most of the SDM users replying to a survey in 2015 (Ahmed et al. 2015) were using either MaxEnt [@Phillips2006] or R [@R] as the first choice software for their analyses (also see @Touchon2016). 
+The software market is diverse, however, including a variety of software developed specifically for SDM (e.g.
+MaxEnt [@Phillips2006], the BioMod and BioMod2 [Thuiller2009] and dismo [@Hijmans?] packages for R, SDMtoolbox [@Brown2014], openModeller [@deSouzaMunoz2009], BioEnsembles [@Diniz-Filho2009], ModEco [@Guo2010]) and software that has been appropriated for SDM studies, such as statistical software (WinBugs and OpenBugs [@Lunn2000]) and other, more general ecological software  (Domain [Carpenter1993], Canoco [@TerBraak2002]).
 
-Reproducibility (and the issues discussed above) were not a primary requirement of these software, and whilst some have the capacity for an analysis to be repeated, many do not. 
-Software such as BioMod, OpenModeller, BioEnsembles, ModEco were developed to enable users to carry out analyses with multiple models.
+Of the SDM software for multi model comparisons (e.g. OpenModeller, BioEnsembles, ModEco), BioMod (as an open-source R package) has the greatest potential to be modified and re-appropriated.
+In essence, BioMod is an SDM environment accessed from R, rather than necessarily being a SDM tool kit for R.
+The R package called sdm [@Naimi2016] implements a variety of analytical functionality in a simplifying framework.
+There is scope for incorporating new techniques but not necessarily a straight forward system for sharing those contributions.
+The BIOVEL workflow environment [@DeGiovanni2015] is an exception in terms of reproducibility, sharing many of the general principles of ZOON.
+Adoption of workflow environments (e.g.
+Kepler, Vis-Trails, Taverna) is low however, presumably due to the users’ unfamiliarity, and the requirement of embedding R within a workflow environment rather than embedding a workflow system within R.
 
-The BIOVEL software [@DeGiovanni2015] is an exception, sharing many of the general principles of ZOON, such as the increased accessibility to science via community contributed resources and increased sustainability by creating an e-infrastructure for this science.
-There are other workflow systems that can be used to wrap around SDM analyses (e.g. Kepler, Vis-Trails, Taverna), yet there has not been a large uptake by the community.
-Perhaps this is because of the unfamiliarity or assumed complexity of these tools.
-
-The SDM package [@Naimi2016] implements a number of modern SDM approaches in a simple modular framework.
-This package also has some functionality for users to incorporate new modelling methods into their analyses.
-There is, however still a signifiact technical barrier to those wishing to incorporate their methods, and to share them more widely.
+Examples of subject specific frameworks are found within the Bioinformatics community, for example the Bioconductor infrastructure that uses R (Gentleman et al.
+2004, https://www.bioconductor.org/), and BioJS, a repository for JavaScript components (Corpas et al.
+2014, https://www.biojs.net/).
+Both software aim to increase the compatibility and sharing of computational tools in order to increase the reuse and robustness of tools.
+In terms of robustness, the CHASTE project for cardiac modelling (Mirams et al.
+2013, http://www.cs.ox.ac.uk/chaste/) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling (Cooper et al.
+2016).
+These frameworks increase the access, usability and robustness of ultimately scientific research.
+Of course, R users could point to the R environment itself, but the stringent requirements of R packages are likely to prevent much of the SDM community from sharing code in a discoverable or usable format.
 

### 2016-11-01 ms.Rmd
add zoon package text, update figure

@@
 Users can create new modules for any modelling method and integrate it into another analysis. 
-
 Similarly to programming languages such as R and Python, facilitating wider development of software will enable the whole research community to develop the software that it requires, rather than rely on the developers of the software they use to do it for them.
 Unlike software packages for programming languages however, the ZOON project enforces a modular structure that ensures interoperability between contributed software modules.
-This also helps to lower the bar for creating modules (each module is simply an R function with a set of required inputs and outputs) and facilitates automated testing of modules ot ensure software quality .
-
-This paper introduces version 0.4-22 of the `zoon` R package, which can be dowloaded from CRAN and which encodes the majority of this functionality.
-First, we describe the modular structure of zoon workflows and demonstrate how completed `zoon` workflows can be shared, evaluated and extended.
-Then we illustrate these concepts, and the contribution we hope `zoon` will make to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.
+This also helps to lower the bar for creating modules (each module is simply an R function with a set of required inputs and outputs) and facilitates automated testing of modules ot ensure software quality.
 
@@
 
-<!— link to powerpoint doc: https://www.dropbox.com/s/afjpv8v6na9y56e/diagrams.pptx?dl=0-->
+<!— link to powerpoint doc: https://www.dropbox.com/s/p7rqww2v0h3a7np/diagrams.pptx?dl=0-->
 
@@
 
-The five module types, corresponding to the five key steps of an SDM anaysis are:
+The five module types, corresponding to the five key steps of an SDM analysis are:
 

### 2016-11-01 ms.Rmd
add future developments text

@@
 
-Tutorials on how to create workflows and modules, as well as full technical details for module developers, are provided as vignettes distributed with `zoon`.
+Reproducibility, open-data and open-science are topics that receive ever more attention all across science.
+Journals (McNutt 2016) and funders (European Commission 2016) are moving towards openly shared data and code resources as a requirement, as was set out in the executive order issued by Barak Obama in 2013 (https://www.whitehouse.gov/the-press-office/2013/05/09/executive-order-making-open-and-machine-readable-new-default-government-).
+These trends strongly suggest researchers must at least find ways of documenting and disseminating code and data, if not ways that enable end users.
+Numerous platforms for publishing analytical outputs exist (e.g.
+GitHub, FigShare, Dryad, CRAN), but end users will still face the demands of locating, deciphering and managing different resources across different platforms.
+Then, the real challenge of open and reproducible science, that has an increased capacity for innovation, is the challenge of increasing the usability of code and data.
 
-The `zoon` packages provides solutions to some of the technical barriers to the ultimate goal of more open, productive, reproducible SDM research.
-Achieving this goal will also depend on overcoming the social hurdle of developing a common tool for comparing and quantitatively evaluating SDMs.
+However, “scientific software is developed either by domain scientists who have little training in user interface (UI) or experience (UX) design, or by software engineers who build broad systems that do not suit the niche needs or practices of domain scientists” (Thomer et al.
+2016).
+Then, there is a need for environments that can address niche needs (e.g. of a modelling discipline such as SDM) and that address usability issues for diverse outputs produced by diverse scientists.
+The SDM community is large enough to warrant a discipline specific platform such as zoon.
+The zoon framework offers a new opportunity for the SDM community to develop community modelling benchmarks, and resolve scientific and methodological questions in ways that the current culture of publishing cannot achieve.
+It may be that zoon can also be used for other modelling activities that have similar workflows, or use similar data and processing steps.
 
-This will require achieving a critical mass of `zoon` modules, so that there is an incentive for species distribution modellers to use the software, and a critical mass of users, so that there is an incentive for SDM methods developers to encode their proposed methods as modules in the community repo.
-At this point `zoon` will provide a common environment in which to evaluate SDM methods and enable the SDM research community to make evidence-based decisions about best practice and the goals for development of the field.
-
-In order to facilitate this social aspect, future work will develop an online platform to enable exploring contributed modules and workflows.
-This platform will also provide an online space to discuss, and openly evaluate, proposed best practices in SDM.
-The zoon R package therefore represents a step towards a more reproducible ecosystem of SDM software.
-
-<!— point to github development site—>
+During the development of ZOON we have improvised a user-centred, participatory design process that builds on our prior experiences with software development and soliciting user feedback at 3 stages.
+An ideation workshop set the initial requirements before any code was written, and two further workshops have allowed users input on an initial prototype distributed via GitHub and then a stable R package.
+This is by no means a full UX approach (Pavelin et al. 2012), but it has directly informed the design. 
+Continued user engagement will be vital to developing a critical mass of zoon modules that can incentivise modellers to contribute to community resources that enable evidence-based decisions about best practice.
+An important step would be the publication of a large model inter-comparison within the zoon environment, such as that produced by Elith et al. (2006) or more recently in Beaumont et al. (2016).
+At present, studies such as these might be considered as ‘closed knowledge’ even if they were reproducible from the publication.
+In order to facilitate a user –friendly approach to distributed working, future work will include the development of an online interface for accessing a repository of contributed modules and modifying workflows.
+This interface might also host methodological discourse and openly evaluate proposed best practices in SDM.
+In sum, we hope that the zoon R package represents a step towards a more reproducible ecosystem of SDM software.
 

### 2016-11-01 ms.Rmd
sorting out citations

@@
 Like many areas of quantitative science Species Distribution Modelling (SDM) has grown rapidly [@Barbosa2015] fueled by the availability of diverse modelling software and data.
-Over the last 20 years, SDM has developed into a very large research community and vast literature, and is now one of “the most widely-reviewed topics in the ecological literature” (Araújo & Peterson XXXX).
+Over the last 20 years, SDM has developed into a very large research community and vast literature, and is now one of “the most widely-reviewed topics in the ecological literature” [@Araujo2012].
 Most SDM software, and that enabling other other computational sub-disciplines, focus on  analytical tasks (such as modelling and statistics) rather than enabling scientists to produce reproducible and modifiable analyses that are essential to the scientific enterprise.
@@
 The ZOON R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
-The module-workflow structure enables scientists to more easily to create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure 1).
+The module-workflow structure enables scientists to more easily to create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure \ref{fig:workflows}).
 
 A lack of sharing reduces the capacity for innovation and validation in SDM, as well as its capacity to be a "self-correcting process" [@roysoc_report].
-For example, a fundamental dispute over the ability of SDM to detect environmental associations (Beale et al.
-2008, Araujou et al.
-2009, Beale et al.
-2009) was left unresolved as only the original publication shared their analytical code (Beale et al.
-2008).
+For example, a fundamental dispute over the ability of SDM to detect environmental associations [@Beale2008,@Araujo2009,@Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].
 Without the resources being made open the critical differences in these analytical steps are assumed rather than proven.
 
-A second example comes from Elith et al. (2006), who reported a study of exceptional scope(“16 modelling methods over 226 species from 6 regions of the world”).
+A second example comes from [@Elith2006], who reported a study of exceptional scope(“16 modelling methods over 226 species from 6 regions of the world”).
 The modelling benchmarks that were reported have been highly influential, yet cannot be updated or modified when new modelling methods are produced as the code and data were not shared.
-A third example comes from meta-analyses [] and methodological reviews [@Yackulic?], which rely on laborious searches through text publications in the absence of shared code and modelling outputs.
+A third example comes from meta-analyses [] and methodological reviews [@Yackulic2013], which rely on laborious searches through text publications in the absence of shared code and modelling outputs.
 SDM then has limited abilities to derive truly informative methodological conclusions that can be acted upon.
@@
 We then illustrate these concepts, and the potential contributions of ZOON to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.
-To assist users a number of tutorials are provided on the github repository and as vignettes distributed with zoon – including  guides on creating workflows and modules, as well as full technical details for module developers.
+To assist users a number of tutorials are provided on the GitHub repository and as vignettes distributed with zoon – including  guides on creating workflows and modules, as well as full technical details for module developers.
 
@@
 
-Most of the SDM users replying to a survey in 2015 (Ahmed et al. 2015) were using either MaxEnt [@Phillips2006] or R [@R] as the first choice software for their analyses (also see @Touchon2016). 
+Most of the SDM users replying to a survey in 2015 [@Ahmed2015] were using either MaxEnt [@Phillips2006] or R [@R] as the first choice software for their analyses (also see @Touchon2016). 
 The software market is diverse, however, including a variety of software developed specifically for SDM (e.g.
-MaxEnt [@Phillips2006], the BioMod and BioMod2 [Thuiller2009] and dismo [@Hijmans?] packages for R, SDMtoolbox [@Brown2014], openModeller [@deSouzaMunoz2009], BioEnsembles [@Diniz-Filho2009], ModEco [@Guo2010]) and software that has been appropriated for SDM studies, such as statistical software (WinBugs and OpenBugs [@Lunn2000]) and other, more general ecological software  (Domain [Carpenter1993], Canoco [@TerBraak2002]).
+MaxEnt [@Phillips2006], the BioMod and BioMod2 [@Thuiller2009] and dismo [@dismo] packages for R, SDMtoolbox [@Brown2014], openModeller [@deSouza2011], BioEnsembles [@Diniz2009], ModEco [@Guo2010]) and software that has been appropriated for SDM studies, such as statistical software (WinBugs and OpenBugs [@Lunn2000]) and other, more general ecological software  (Domain [@Carpenter1993], Canoco [@TerBraak2002]).
 
@@
 
-Examples of subject specific frameworks are found within the Bioinformatics community, for example the Bioconductor infrastructure that uses R (Gentleman et al.
-2004, https://www.bioconductor.org/), and BioJS, a repository for JavaScript components (Corpas et al.
-2014, https://www.biojs.net/).
+Examples of subject specific frameworks are found within the Bioinformatics community, for example the Bioconductor infrastructure that uses R (@Gentleman2004, https://www.bioconductor.org/), and BioJS, a repository for JavaScript components (@Corpas2014, https://www.biojs.net/).
 Both software aim to increase the compatibility and sharing of computational tools in order to increase the reuse and robustness of tools.
-In terms of robustness, the CHASTE project for cardiac modelling (Mirams et al.
-2013, http://www.cs.ox.ac.uk/chaste/) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling (Cooper et al.
-2016).
+In terms of robustness, the CHASTE project for cardiac modelling (@Mirams2013, http://www.cs.ox.ac.uk/chaste/) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling [@Cooper2016].
 These frameworks increase the access, usability and robustness of ultimately scientific research.
@@
 
-<!--
-Possible experiments:
+<!-- Possible experiments:
 * evaluating with different criteria (e.g. deviance or pseudo-r-squared which measure calibration capacity)
@@
 Reproducibility, open-data and open-science are topics that receive ever more attention all across science.
-Journals (McNutt 2016) and funders (European Commission 2016) are moving towards openly shared data and code resources as a requirement, as was set out in the executive order issued by Barak Obama in 2013 (https://www.whitehouse.gov/the-press-office/2013/05/09/executive-order-making-open-and-machine-readable-new-default-government-).
+Journals [@McNutt2016] and funders [@EC2016] are moving towards openly shared data and code resources as a requirement, as was set out in the executive order issued by Barak Obama in 2013 (https://www.whitehouse.gov/the-press-office/2013/05/09/executive-order-making-open-and-machine-readable-new-default-government-).
 These trends strongly suggest researchers must at least find ways of documenting and disseminating code and data, if not ways that enable end users.
@@
 
-However, “scientific software is developed either by domain scientists who have little training in user interface (UI) or experience (UX) design, or by software engineers who build broad systems that do not suit the niche needs or practices of domain scientists” (Thomer et al.
-2016).
+However, “scientific software is developed either by domain scientists who have little training in user interface (UI) or experience (UX) design, or by software engineers who build broad systems that do not suit the niche needs or practices of domain scientists” [@Thomer2016].
 Then, there is a need for environments that can address niche needs (e.g. of a modelling discipline such as SDM) and that address usability issues for diverse outputs produced by diverse scientists.
@@
 An ideation workshop set the initial requirements before any code was written, and two further workshops have allowed users input on an initial prototype distributed via GitHub and then a stable R package.
-This is by no means a full UX approach (Pavelin et al. 2012), but it has directly informed the design. 
+This is by no means a full UX approach [@Pavelin2012], but it has directly informed the design. 
 Continued user engagement will be vital to developing a critical mass of zoon modules that can incentivise modellers to contribute to community resources that enable evidence-based decisions about best practice.
-An important step would be the publication of a large model inter-comparison within the zoon environment, such as that produced by Elith et al. (2006) or more recently in Beaumont et al. (2016).
+An important step would be the publication of a large model inter-comparison within the zoon environment, such as that produced by @Elith2006 or more recently in @Beaumont2016.
 At present, studies such as these might be considered as ‘closed knowledge’ even if they were reproducible from the publication.

### 2016-11-01 ms.Rmd
sorting out figure captions

@@
                warning = FALSE,
+               fig.align = 'center',
                dev = c('png'),
@@
 The ZOON R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
-The module-workflow structure enables scientists to more easily to create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure \ref{fig:workflows}).
+The module-workflow structure enables scientists to more easily to create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure 1).
 
 A lack of sharing reduces the capacity for innovation and validation in SDM, as well as its capacity to be a "self-correcting process" [@roysoc_report].
-For example, a fundamental dispute over the ability of SDM to detect environmental associations [@Beale2008,@Araujo2009,@Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].
+For example, a fundamental dispute over the ability of SDM to detect environmental associations [@Beale2008, @Araujo2009, @Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].
 Without the resources being made open the critical differences in these analytical steps are assumed rather than proven.
@@
 Users combine these modules via a call to the `workflow` function, which executes each module in turn, before returning a `zoonWorkflow` object - a shareable, extensible and fully reproducible documentation of the SDM analysis.
-Figure \ref{fig:workflows} illustrates the structure of a `zoon` workflow comprising the five module types and illustrates how multiple modules may be combined at each stage.
+Figure 1 illustrates the structure of a `zoon` workflow comprising the five module types and illustrates how multiple modules may be combined at each stage.
 The module types and module combination options are described in more detail below.
@@
 
-![The modular SDM structure encoded by a zoon workflow.
+```{r, echo = FALSE, out.width = "350px", fig.cap = "The modular SDM structure encoded by a zoon workflow.
 A) Flow diagram representing the five module types and intermediate objects: df - an occurrence dataframe; ras - a RasterStack object; mod - a ZoonModel object; ?
@@
 C) The flow diagram implied by listing modules of the same type.
-Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the `zoon` vignette 'Building a module'.
-\label{fig:workflows}](./figs/diagrams.png)
+Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the `zoon` vignette 'Building a module'."}
+knitr::include_graphics("figs/diagrams.png")
+```
 

### 2016-11-01 ms.Rmd
bugfixes and tidying captions

@@
 
-A second example comes from [@Elith2006], who reported a study of exceptional scope(“16 modelling methods over 226 species from 6 regions of the world”).
+A second example comes from [@Elith2006], who reported a study of exceptional scope (“16 modelling methods over 226 species from 6 regions of the world”).
 The modelling benchmarks that were reported have been highly influential, yet cannot be updated or modified when new modelling methods are produced as the code and data were not shared.
@@
 
-```{r, echo = FALSE, out.width = "350px", fig.cap = "The modular SDM structure encoded by a zoon workflow.
+```{r flow_diagram, echo = FALSE, out.width = "375px", fig.cap = "The modular SDM structure encoded by a zoon workflow.
 A) Flow diagram representing the five module types and intermediate objects: df - an occurrence dataframe; ras - a RasterStack object; mod - a ZoonModel object; ?
@@
 ```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = "hide", results = "hide", message = FALSE, fig.height = 9, fig.width = 9}
-Feng_Papes <- 
+FengPapes <- 
   workflow(occurrence = SpOcc('Dasypus novemcinctus',
@@
 This module plots a static map of the predicted distribution, and the underlying data.
-This map is shown in \ref{fig:feng_papes_plots}A.
+This map is shown in Figure 2A.
 The resulting workflow contains all the code required to re-run the workflow, the input data and the results of executing each module.
-The object `Feng_Papes` could therefore be saved as a binary file and shared as a reproducible representation of this research.
+The object `FengPapes` could therefore be saved as a binary file and shared as a reproducible representation of this research.
 
 Next, we update the workflow to produce an interactive map enabling anyone to inspect the data and predictions on a zoomable map, and to inspect the response curves of the fitted model.
-These outputs are shown in Figure This map is shown in \ref{fig:feng_papes_plots}, panels B and C.
+These outputs are shown in Figure This map is shown in Figure 2, panels B and C.
 
 ```{r fengChangeWorkflow, eval = TRUE, fig.show = "hide", fig.height = 9}
-Feng_Papes_Update <- 
-  ChangeWorkflow(workflow = Feng_Papes,
+FengPapesUpdate <- 
+  ChangeWorkflow(workflow = FengPapes,
                    output = Chain(InteractiveMap, 
@@
 
-![Outputs of the workflow objects `Feng_Papes` and `Feng_Papes_Update`.
-A) Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap` module in the workflow `Feng_Papes`.
+```{r feng_papes_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Outputs of the workflow objects `FengPapes' and `FengPapesUpdate'.
+A) Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap' module in the workflow `FengPapes'.
 White areas are masked due to being in the sea or outside the MESS mask.
-B) A screenshot of the interactive map produced by the `InteractiveMap` moduels in the workflow `Feng_Papes_Update`, displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results.
-C) A response curve produced by the `ResponseCurve` module for the first covariate, bio1 in the workflow `Feng_Papes_Update`.
-\label{fig:feng_papes_plots}](./figs/combinefengupdate-1.png)
-
-<!-- add the original and the response curves to this figure too -->
+B) A screenshot of the interactive map produced by the `InteractiveMap' modules in the workflow `FengPapesUpdate', displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results.
+C) A response curve produced by the `ResponseCurve' module for the first covariate, bio1 in the workflow `FengPapesUpdate'."}
+knitr::include_graphics("figs/combinefengupdate-1.png")
+```
 
@@
   # get index to rows in sub_df, update the full dataset and return
-  pres_keep_idx <- rownames(th[[1]])
+  pres_keep_idx <- as.numeric(rownames(th[[1]]))
   .data$df <- rbind(df[pres_idx,][pres_keep_idx, ],

### 2016-11-01 ms.Rmd
re-organising Figure 2

@@
   ChangeWorkflow(workflow = FengPapes,
-                   output = Chain(InteractiveMap, 
-                                  ResponseCurve(cov = 1)))
+                   output = Chain(ResponseCurve(cov = 1),
+                                  InteractiveMap))
 ```
 
-```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 3.2}
+```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 4.5, fig.width = 4.5}
 # combine the static and interactive maps and first effect plot
@@
 
-# set up layout so that heights are the same for both panels
+# set up layout so that heights are the same for all panels
 plot_height <- min(nrow(r_map1), nrow(r_map), nrow(r_resp))
@@
 gap <- 2
-widths <- c(width_map1, width_map, width_resp)
-widths <- round(widths * ((200 - (2 * gap)) / sum(widths)))
-widths <- c(widths[1], gap, widths[2], gap, widths[3])
-mat <- matrix(rep(1:5, widths),
-              nrow = 1)
+
+# top row: map1 and resp
+widths_top <- c(width_map1, width_resp)
+widths_top <- round(widths_top * ((200 - (1 * gap)) / sum(widths_top)))
+widths_top <- c(widths_top[1], gap, widths_top[2])
+
+top_row <- rep(1:3, widths_top)
+bottom_row <- rep(5, length(top_row))
+
+heights <- c(plot_height + 1, plot_height)
+heights <- round(heights * ((50 - (1)) / sum(heights)))
+heights <- c(heights[1], 1, heights[2])
+
+top_matrix <- do.call(rbind, replicate(heights[1], top_row, simplify = FALSE))
+middle_matrix <- matrix(4, nrow = heights[2], ncol = ncol(top_matrix))
+bottom_matrix <- matrix(5, nrow = heights[3], ncol = ncol(top_matrix))
+
+mat <- rbind(top_matrix, middle_matrix, bottom_matrix)
 layout(mat)
 
-# plot statis map
+# plot static map
 plotRGB(r_map1, maxpixels = Inf)
@@
         side = 3,
-        line = -3,
-        adj = 0)
-
-# gap
-plot.new()
-
-# plot interactive map
-plotRGB(r_map, maxpixels = Inf)
-
-# add a panel letter
-mtext(text = 'B',
-        side = 3,
-        line = -3,
+        line = -1.5,
         adj = 0)
@@
 # add a panel letter
+mtext(text = 'B',
+        side = 3,
+        line = -1.5,
+        adj = 0)
+
+# gap
+plot.new()
+
+# plot interactive map
+plotRGB(r_map, maxpixels = Inf)
+
+# add a panel letter
 mtext(text = 'C',
         side = 3,
-        line = -3,
+        line = -1.5,
         adj = 0)
@@
 White areas are masked due to being in the sea or outside the MESS mask.
-B) A screenshot of the interactive map produced by the `InteractiveMap' modules in the workflow `FengPapesUpdate', displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results.
-C) A response curve produced by the `ResponseCurve' module for the first covariate, bio1 in the workflow `FengPapesUpdate'."}
+B) A response curve produced by the `ResponseCurve' module for the first covariate, bio1 in the workflow `FengPapesUpdate'.
+C) A screenshot of the interactive map produced by the `InteractiveMap' modules in the workflow `FengPapesUpdate', displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results."}
 knitr::include_graphics("figs/combinefengupdate-1.png")

### 2016-11-01 ms.Rmd
more tidying, and increasing figure resolution

@@
 
-Examples of subject specific frameworks are found within the Bioinformatics community, for example the Bioconductor infrastructure that uses R (@Gentleman2004, https://www.bioconductor.org/), and BioJS, a repository for JavaScript components (@Corpas2014, https://www.biojs.net/).
+Examples of subject specific frameworks are found within the Bioinformatics community, for example the Bioconductor infrastructure that uses R (@Gentleman2004, www.bioconductor.org), and BioJS, a repository for JavaScript components (@Corpas2014, www.biojs.net).
 Both software aim to increase the compatibility and sharing of computational tools in order to increase the reuse and robustness of tools.
-In terms of robustness, the CHASTE project for cardiac modelling (@Mirams2013, http://www.cs.ox.ac.uk/chaste/) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling [@Cooper2016].
+In terms of robustness, the CHASTE project for cardiac modelling (@Mirams2013, www.cs.ox.ac.uk/chaste) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling [@Cooper2016].
 These frameworks increase the access, usability and robustness of ultimately scientific research.
@@
 
-```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = "hide", results = "hide", message = FALSE, fig.height = 9, fig.width = 9}
+```{r fengworkflow, eval = TRUE, dpi = 600, fig.show = "hide", results = "hide", message = FALSE, fig.height = 9, fig.width = 9}
 FengPapes <- 
@@
 
-```{r merowworkflowtemp, eval = TRUE}
+```{r merowworkflowtemp, dpi = 600, eval = TRUE}
 Merow_Silander <- workflow(
@@
 
-```{r merowworkflow, eval = FALSE}
+```{r merowworkflow,dpi = 600, eval = FALSE}
 
@@
 
-```{r merowChangeWorkflow, eval = FALSE}
+```{r merowChangeWorkflow, dpi = 600, eval = FALSE}
 Merow_Silander_Spatial <- ChangeWorkflow(
@@
 Reproducibility, open-data and open-science are topics that receive ever more attention all across science.
-Journals [@McNutt2016] and funders [@EC2016] are moving towards openly shared data and code resources as a requirement, as was set out in the executive order issued by Barak Obama in 2013 (https://www.whitehouse.gov/the-press-office/2013/05/09/executive-order-making-open-and-machine-readable-new-default-government-).
+Journals [@McNutt2016] and funders [@EC2016] are moving towards openly shared data and code resources as a requirement, as was set out in the executive order issued by US President Barack Obama in 2013 [@Obama2013].
 These trends strongly suggest researchers must at least find ways of documenting and disseminating code and data, if not ways that enable end users.

### 2016-11-01 ms.Rmd

@@
 
-5. We demonstrate `zoon` by recreating SDM analyses from two published research articles as zoon workflows, which readers can interrogate and extend.
+5. We demonstrate `zoon` by recreating SDM analyses from two published research articles as `zoon` workflows, which readers can interrogate and extend.
 
@@
 If research materials and outputs are shared, they are rarely disseminated in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g.
-Figshare, GitHub).
-The ZOON R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
+figshare, GitHub).
+The `zoon` R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
 The module-workflow structure enables scientists to more easily to create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure 1).
@@
 
-This paper introduces version 0.4-22 of the `ZOON` R package, which enables users to address issues of compatibility, accessibility, sharing and reproducibility for their SDM analyses.
-Firstly, we describe the modular structure of ZOON workflows and demonstrate how completed workflows can be shared, evaluated and extended.
-We then illustrate these concepts, and the potential contributions of ZOON to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.
-To assist users a number of tutorials are provided on the GitHub repository and as vignettes distributed with zoon – including  guides on creating workflows and modules, as well as full technical details for module developers.
+This paper introduces version 0.4-22 of the `zoon` R package, which enables users to address issues of compatibility, accessibility, sharing and reproducibility for their SDM analyses.
+Firstly, we describe the modular structure of `zoon` workflows and demonstrate how completed workflows can be shared, evaluated and extended.
+We then illustrate these concepts, and the potential contributions of `zoon` to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.
+To assist users a number of tutorials are provided on the GitHub repository and as vignettes distributed with `zoon` – including  guides on creating workflows and modules, as well as full technical details for module developers.
 
@@
 There is scope for incorporating new techniques but not necessarily a straight forward system for sharing those contributions.
-The BIOVEL workflow environment [@DeGiovanni2015] is an exception in terms of reproducibility, sharing many of the general principles of ZOON.
+The BIOVEL workflow environment [@DeGiovanni2015] is an exception in terms of reproducibility, sharing many of the general principles of `zoon`.
 Adoption of workflow environments (e.g.
@@
 
-Examples of subject specific frameworks are found within the Bioinformatics community, for example the Bioconductor infrastructure that uses R (@Gentleman2004, www.bioconductor.org), and BioJS, a repository for JavaScript components (@Corpas2014, www.biojs.net).
+Examples of subject specific frameworks are found within the Bioinformatics community, for example the Bioconductor infrastructure that uses R (@Gentleman2004, `bioconductor.org`), and BioJS, a repository for JavaScript components (@Corpas2014, `biojs.net`).
 Both software aim to increase the compatibility and sharing of computational tools in order to increase the reuse and robustness of tools.
-In terms of robustness, the CHASTE project for cardiac modelling (@Mirams2013, www.cs.ox.ac.uk/chaste) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling [@Cooper2016].
+In terms of robustness, the CHASTE project for cardiac modelling (@Mirams2013, `cs.ox.ac.uk/chaste`) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling [@Cooper2016].
 These frameworks increase the access, usability and robustness of ultimately scientific research.
@@
 
-### The ZOON project
+### The Zoon Project
 
-The ZOON project takes a different approach, instead defining only the core architecture of workflows, and interfacing with an open repository of community-contributed software modules.
+The Zoon Project takes a different approach, instead defining only the core architecture of workflows, and interfacing with an open repository of community-contributed software modules.
 Users can create new modules for any modelling method and integrate it into another analysis. 
 Similarly to programming languages such as R and Python, facilitating wider development of software will enable the whole research community to develop the software that it requires, rather than rely on the developers of the software they use to do it for them.
-Unlike software packages for programming languages however, the ZOON project enforces a modular structure that ensures interoperability between contributed software modules.
+Unlike software packages for programming languages however, The Zoon Project enforces a modular structure that ensures interoperability between contributed software modules.
 This also helps to lower the bar for creating modules (each module is simply an R function with a set of required inputs and outputs) and facilitates automated testing of modules ot ensure software quality.
@@
 `zoon` workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.
-A zoon workflow returns an object of class `zoonWorkflow`.
+A `zoon` workflow returns an object of class `zoonWorkflow`.
 This object contains all the information needed to reproduce the analysis.
@@
 
-As long as a valid module function is defined in an R session, it can be used in workflows in combination with modules downloaded from the ZOON module repository.
+As long as a valid module function is defined in an R session, it can be used in workflows in combination with modules downloaded from the `zoon` module repository.
 Whilst this makes it easy to develop modules locally, the strength of `zoon` comes from the ability to upload modules to the online repository so that others can access them.
@@
 
-`zoon` provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform [`figshare.com`]().
+`zoon` provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform figshare.
 `ZoonFigshare` takes a workflow object and some minimal metadata and uploads the workflow as an RData object to the user's figshare profile, for others to download, inspect and modify.
@@
 We demonstrate the `zoon` R package by recreating two SDM analyses from published research articles as `zoon` workflows and showing how these workflow objects can easily be extended with new modules.
-Workflow objects created by these analyses can be accessed at [http://figshare.com/articles/zoon_applications_paper_workflows](http://figshare.com/articles/zoon_applications_paper_workflows).
+Workflow objects created by these analyses can be accessed at [`figshare.com/articles/zoon_applications_paper_workflows`](http://figshare.com/articles/zoon_applications_paper_workflows).
 We encourage readers to download, interrogate and alter these workflows for themselves.
-Full code and metadata for all of the modules used in the examples below can be found at [https://github.com/zoonproject/modules/R](https://github.com/zoonproject/modules/R)
+Full code and metadata for all of the modules used in the examples below can be found at [`github.com/zoonproject/modules/R`](https://github.com/zoonproject/modules/R)
 
@@
 
-Such a model can be quickly and easily re-constructed as a `zoon` workflow using modules available in the ZOON module repository.
+Such a model can be quickly and easily re-constructed as a `zoon` workflow using modules available in the `zoon` module repository.
 @Feng2015 used a combination of occurrence data from GBIF, and additional occurrence data manually collected from the published literature.
@@
 
-To convert this code into a ZOON module, we need to write it to a standalone file (named `spThin.R`) with the metadata required to build the module documentation.
+To convert this code into a `zoon` module, we need to write it to a standalone file (named `spThin.R`) with the metadata required to build the module documentation.
 The `zoon` function `BuildModule()` helps with this step, and can also run checks to make sure we got everything right:
@@
                             'et al. (2015): dx.doi.org/10.1111/ecog.01132'),
-            author = 'ZOON Developers',
+            author = 'zoon Developers',
             email = 'zoonproject@gmail.com',
@@
 
-This module can now be shared so that others can use it in their ZOON workflows.
+This module can now be shared so that others can use it in their `zoon` workflows.
 The best way of sharing it is via the `zoon` modules repository.
@@
 
-All of these have now been uploaded to the ZOON modules repository under the names `CarolinaWrenPO`, `CarolinaWrenPA`, `CarolinaWrenRasters` and `MaxLike`.
+All of these have now been uploaded to the `zoon` modules repository under the names `CarolinaWrenPO`, `CarolinaWrenPA`, `CarolinaWrenRasters` and `MaxLike`.
 
@@
 Numerous platforms for publishing analytical outputs exist (e.g.
-GitHub, FigShare, Dryad, CRAN), but end users will still face the demands of locating, deciphering and managing different resources across different platforms.
+GitHub, figshare, Dryad, CRAN), but end users will still face the demands of locating, deciphering and managing different resources across different platforms.
 Then, the real challenge of open and reproducible science, that has an increased capacity for innovation, is the challenge of increasing the usability of code and data.
@@
 Then, there is a need for environments that can address niche needs (e.g. of a modelling discipline such as SDM) and that address usability issues for diverse outputs produced by diverse scientists.
-The SDM community is large enough to warrant a discipline specific platform such as zoon.
-The zoon framework offers a new opportunity for the SDM community to develop community modelling benchmarks, and resolve scientific and methodological questions in ways that the current culture of publishing cannot achieve.
+The SDM community is large enough to warrant a discipline specific platform such as `zoon`.
+The Zoon Project offers a new opportunity for the SDM community to develop community modelling benchmarks, and resolve scientific and methodological questions in ways that the current culture of publishing cannot achieve.
 It may be that zoon can also be used for other modelling activities that have similar workflows, or use similar data and processing steps.
 
-During the development of ZOON we have improvised a user-centred, participatory design process that builds on our prior experiences with software development and soliciting user feedback at 3 stages.
+During the development of `zoon` we have improvised a user-centred, participatory design process that builds on our prior experiences with software development and soliciting user feedback at 3 stages.
 An ideation workshop set the initial requirements before any code was written, and two further workshops have allowed users input on an initial prototype distributed via GitHub and then a stable R package.
 This is by no means a full UX approach [@Pavelin2012], but it has directly informed the design. 
-Continued user engagement will be vital to developing a critical mass of zoon modules that can incentivise modellers to contribute to community resources that enable evidence-based decisions about best practice.
-An important step would be the publication of a large model inter-comparison within the zoon environment, such as that produced by @Elith2006 or more recently in @Beaumont2016.
+Continued user engagement will be vital to developing a critical mass of `zoon` modules that can incentivise modellers to contribute to community resources that enable evidence-based decisions about best practice.
+An important step would be the publication of a large model inter-comparison within the `zoon` environment, such as that produced by @Elith2006 or more recently in @Beaumont2016.
 At present, studies such as these might be considered as ‘closed knowledge’ even if they were reproducible from the publication.
@@
 This interface might also host methodological discourse and openly evaluate proposed best practices in SDM.
-In sum, we hope that the zoon R package represents a step towards a more reproducible ecosystem of SDM software.
+In sum, we hope that the `zoon` R package represents a step towards a more reproducible ecosystem of SDM software.
 

### 2016-11-01 ms.Rmd

@@
 
-A lack of sharing reduces the capacity for innovation and validation in SDM, as well as its capacity to be a "self-correcting process" [@roysoc_report].
+A lack of sharing reduces the capacity for innovation and validation in SDM, as well as its capacity to be a "self-correcting process" [@Boulton2012].
 For example, a fundamental dispute over the ability of SDM to detect environmental associations [@Beale2008, @Araujo2009, @Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].

### 2016-11-11 ms.Rmd
toms initial changes

--- 2016-11-01 ms.Rmd
@@
                warning = FALSE,
-               fig.align = 'center',
                dev = c('png'),
@@
 
-Like many areas of quantitative science Species Distribution Modelling (SDM) has grown rapidly [@Barbosa2015] fueled by the availability of diverse modelling software and data.
-Over the last 20 years, SDM has developed into a very large research community and vast literature, and is now one of “the most widely-reviewed topics in the ecological literature” [@Araujo2012].
-Most SDM software, and that enabling other other computational sub-disciplines, focus on  analytical tasks (such as modelling and statistics) rather than enabling scientists to produce reproducible and modifiable analyses that are essential to the scientific enterprise.
+Like many areas of quantitative science, Species Distribution Modelling (SDM) has grown rapidly [@Barbosa2015] - fuelled by the availability of diverse modelling software and data.
+Over the last 20 years, SDM has developed into a very large research community and vast literature, and is now one of “the most widely-reviewed topics in the ecological literature” (Araújo & Peterson XXXX).
+Most of the software that enables SDM research, and that of other computational sub-disciplines, focuses on the analytical tasks of science (such as modelling and statistics) rather than tasks that are essential to the scientific enterprise, such as communication and producing reproducible outputs.
 
-Computational sciences are then presented with a modern version of the reproducibility problem as computational research is largely inaccessible let alone reusable.
-If research materials and outputs are shared, they are rarely disseminated in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g.
+This has presented computational science with a very modern version of the reproducibility problem as computational research is largely inaccessible ().
+If research materials and outputs are shared, they are rarely in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g.
 Figshare, GitHub).
 The ZOON R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
-The module-workflow structure enables scientists to more easily to create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure 1).
+The module-workflow structure offers a format by which scientists can more easily to create and share components of their analysis, and then access, modify, reuse and combine the components of others (see below and Figure 1).
 
-A lack of sharing reduces the capacity for innovation and validation in SDM, as well as its capacity to be a "self-correcting process" [@roysoc_report].
-For example, a fundamental dispute over the ability of SDM to detect environmental associations [@Beale2008, @Araujo2009, @Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].
-Without the resources being made open the critical differences in these analytical steps are assumed rather than proven.
+To date, the lack of sharing has reduced the capacity for innovation and validation in SDM, as well as its capacity for self-correction ().
+For example, a fundamental dispute over the ability of SDM to detect environmental associations (Beale et al.
+2008, Araujou et al.
+2009, Beale et al.
+2009) was left unresolved as only code in the original publication was published (Beale et al.
+2008).
+Without the resources being made open, the critical differences in these analysis are assumed rather than proven.
 
-A second example comes from [@Elith2006], who reported a study of exceptional scope (“16 modelling methods over 226 species from 6 regions of the world”).
-The modelling benchmarks that were reported have been highly influential, yet cannot be updated or modified when new modelling methods are produced as the code and data were not shared.
-A third example comes from meta-analyses [] and methodological reviews [@Yackulic2013], which rely on laborious searches through text publications in the absence of shared code and modelling outputs.
-SDM then has limited abilities to derive truly informative methodological conclusions that can be acted upon.
+A second example comes from Elith et al. (2006), who reported a study of exceptional scope(“16 modelling methods over 226 species from 6 regions of the world”).
+The modelling benchmarks that were reported have been highly influential, yet cannot be updated as the code and data were not shared.
+The community cannot update or modify the benchmarks when new modelling method is produced (e.g. GRAF, Golding…).
+A third example comes from meta-analyses () and methodological reviews (), which require laborious searches of publications in the absence of shared code and modelling outputs.
+These findings cannot be turned into reanalyses that would yield truly informative methodological conclusions.
 
-These three examples broadly illustrate why the data and models underpinning SDM need to become more repeatable, reproducible, accessible, modifiable and updateable.
-A productive science relies on the accumulation of knowledge that can be reused and appropriated to advance understanding and resolve methodological issues.
-Thus while many software are pitched as being ‘easy to use’ or ‘user friendly’, this is infrequently the case for the secondary user.
+These three scenarios are illustrative of some of the need for the data and modelling aspects of SDM to become more repeatable, reproducible, accessible, modifiable and updateable.
+Each are crucial examples of the barriers to this science.
 
-This paper introduces version 0.4-22 of the `ZOON` R package, which enables users to address issues of compatibility, accessibility, sharing and reproducibility for their SDM analyses.
-Firstly, we describe the modular structure of ZOON workflows and demonstrate how completed workflows can be shared, evaluated and extended.
-We then illustrate these concepts, and the potential contributions of ZOON to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.
-To assist users a number of tutorials are provided on the GitHub repository and as vignettes distributed with zoon – including  guides on creating workflows and modules, as well as full technical details for module developers.
+Science has to be this way and written statements of the modelling carried out do not facilitate a productive science. 
+SDM, as a science, is not extensible given current working practices.
 
+### Solutions to these problems
+
+The core, soluble problems are:
+
+Difficult for new SDM users to access latest methods.
+
+Difficult for anyone to compare methods effectively.
+
+Equally difficult for methods developers to disseminate their advances.
+
+<!-- Past work towards this goal & where it falls short - the USP of ZOON and requirements -->
 ### Reproducible software projects
 
-Most of the SDM users replying to a survey in 2015 [@Ahmed2015] were using either MaxEnt [@Phillips2006] or R [@R] as the first choice software for their analyses (also see @Touchon2016). 
-The software market is diverse, however, including a variety of software developed specifically for SDM (e.g.
-MaxEnt [@Phillips2006], the BioMod and BioMod2 [@Thuiller2009] and dismo [@dismo] packages for R, SDMtoolbox [@Brown2014], openModeller [@deSouza2011], BioEnsembles [@Diniz2009], ModEco [@Guo2010]) and software that has been appropriated for SDM studies, such as statistical software (WinBugs and OpenBugs [@Lunn2000]) and other, more general ecological software  (Domain [@Carpenter1993], Canoco [@TerBraak2002]).
+Most of the SDM users replying to a survey in 2015 (Ahmed et al. 2015) were using either MaxEnt or R as the first choice software for their analyses. 
+The software market is much more diverse, however, including a variety of software developed specifically for SDM (e.g.
+MaxEnt (), the BioMod () and dismo () packages for R, SDMtoolbox (Brown 2014), openModeller (), BioEnsembles (), ModEco ()) and software that has been appropriated for SDM studies, such as statistical software (R (), WinBugs (), OpenBugs (), Python()) and other less general software  (Domain (), Canoco (), MARS()).
 
-Of the SDM software for multi model comparisons (e.g. OpenModeller, BioEnsembles, ModEco), BioMod (as an open-source R package) has the greatest potential to be modified and re-appropriated.
-In essence, BioMod is an SDM environment accessed from R, rather than necessarily being a SDM tool kit for R.
-The R package called sdm [@Naimi2016] implements a variety of analytical functionality in a simplifying framework.
-There is scope for incorporating new techniques but not necessarily a straight forward system for sharing those contributions.
-The BIOVEL workflow environment [@DeGiovanni2015] is an exception in terms of reproducibility, sharing many of the general principles of ZOON.
-Adoption of workflow environments (e.g.
-Kepler, Vis-Trails, Taverna) is low however, presumably due to the users’ unfamiliarity, and the requirement of embedding R within a workflow environment rather than embedding a workflow system within R.
+Reproducibility (and the issues discussed above) were not a primary requirement of these software, and whilst some have the capacity for an analysis to be repeated, many do not. 
+Software such as BioMod, OpenModeller, BioEnsembles, ModEco were developed to enable users to carry out analyses with multiple models.
 
-Examples of subject specific frameworks are found within the Bioinformatics community, for example the Bioconductor infrastructure that uses R (@Gentleman2004, www.bioconductor.org), and BioJS, a repository for JavaScript components (@Corpas2014, www.biojs.net).
-Both software aim to increase the compatibility and sharing of computational tools in order to increase the reuse and robustness of tools.
-In terms of robustness, the CHASTE project for cardiac modelling (@Mirams2013, www.cs.ox.ac.uk/chaste) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling [@Cooper2016].
-These frameworks increase the access, usability and robustness of ultimately scientific research.
-Of course, R users could point to the R environment itself, but the stringent requirements of R packages are likely to prevent much of the SDM community from sharing code in a discoverable or usable format.
+The BIOVEL software [@DeGiovanni2015] is an exception, sharing many of the general principles of ZOON, such as the increased accessibility to science via community contributed resources and increased sustainability by creating an e-infrastructure for this science.
+There are other workflow systems that can be used to wrap around SDM analyses (e.g. Kepler, Vis-Trails, Taverna), yet there has not been a large uptake by the community.
+Perhaps this is because of the unfamiliarity or assumed complexity of these tools.
+
+The SDM package [@Naimi2016] implements a number of modern SDM approaches in a simple modular framework.
+This package also has some functionality for users to incorporate new modelling methods into their analyses.
+There is, however still a signifiact technical barrier to those wishing to incorporate their methods, and to share them more widely.
 
@@
 
-The ZOON project takes a different approach, instead defining only the core architecture of workflows, and interfacing with an open repository of community-contributed software modules.
+The ZOON project takes a different approach, defining only the core architecture of workflows, and interfacing with an open repository of community-contributed software modules.
 Users can create new modules for any modelling method and integrate it into another analysis.
+
 Similarly to programming languages such as R and Python, facilitating wider development of software will enable the whole research community to develop the software that it requires, rather than rely on the developers of the software they use to do it for them.
 Unlike software packages for programming languages however, the ZOON project enforces a modular structure that ensures interoperability between contributed software modules.
-This also helps to lower the bar for creating modules (each module is simply an R function with a set of required inputs and outputs) and facilitates automated testing of modules ot ensure software quality.
+This also helps to lower the bar for creating modules (each module is simply an R function with a set of required inputs and outputs) and facilitates automated testing of modules to ensure software quality.
+
+This paper introduces version 0.4-22 of the `zoon` R package, which can be dowloaded from CRAN and which encodes the majority of this functionality.
+First, we describe the modular structure of zoon workflows and demonstrate how completed `zoon` workflows can be shared, evaluated and extended.
+Then we illustrate these concepts, and the contribution we hope `zoon` will make to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.
 
@@
 Users combine these modules via a call to the `workflow` function, which executes each module in turn, before returning a `zoonWorkflow` object - a shareable, extensible and fully reproducible documentation of the SDM analysis.
-Figure 1 illustrates the structure of a `zoon` workflow comprising the five module types and illustrates how multiple modules may be combined at each stage.
+Figure \ref{fig:workflows} illustrates the structure of a `zoon` workflow comprising the five module types and illustrates how multiple modules may be combined at each stage.
 The module types and module combination options are described in more detail below.
 
-<!— link to powerpoint doc: https://www.dropbox.com/s/p7rqww2v0h3a7np/diagrams.pptx?dl=0-->
+<!— link to powerpoint doc: https://www.dropbox.com/s/afjpv8v6na9y56e/diagrams.pptx?dl=0-->
 
-```{r flow_diagram, echo = FALSE, out.width = "375px", fig.cap = "The modular SDM structure encoded by a zoon workflow.
+![The modular SDM structure encoded by a zoon workflow.
 A) Flow diagram representing the five module types and intermediate objects: df - an occurrence dataframe; ras - a RasterStack object; mod - a ZoonModel object; ?
@@
 C) The flow diagram implied by listing modules of the same type.
-Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the `zoon` vignette 'Building a module'."}
-knitr::include_graphics("figs/diagrams.png")
-```
+Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the `zoon` vignette 'Building a module'.
+\label{fig:workflows}](./figs/diagrams.png)
 
@@
 
-The five module types, corresponding to the five key steps of an SDM analysis are:
+The five module types, corresponding to the five key steps of an SDM anaysis are:
 
@@
 
-#### Lists and Chains
+#### Lists, Chains and Replicates
 
@@
 This displays the names of all modules and the broader structure of the workflow as well as displaying which modules are not available in the online repository (Figure ).
+
+```{r plotworkflow, eval = TRUE, echo = FALSE, fig.height = 4, fig.width = 9, dpi = 300}
+
+work <- workflow(occurrence = UKAnophelesPlumbeus,
+                 covariate  = UKAir,
+                 process    = Chain(Background(n = 70), Crossvalidate(k = 5)),
+                 model      = list(LogisticRegression, RandomForest, QuickGRaF),
+                 output     = PerformanceMeasures)
+
+plot(work)
+```
 
@@
 
-```{r fengworkflow, eval = TRUE, dpi = 600, fig.show = "hide", results = "hide", message = FALSE, fig.height = 9, fig.width = 9}
-FengPapes <- 
+```{r fengworkflow, eval = TRUE, dpi = 300, fig.show = "hide", results = "hide", message = FALSE, fig.height = 9, fig.width = 9}
+Feng_Papes <- 
   workflow(occurrence = SpOcc('Dasypus novemcinctus',
@@
 This module plots a static map of the predicted distribution, and the underlying data.
-This map is shown in Figure 2A.
+This map is shown in \ref{fig:feng_papes_plots}A.
 The resulting workflow contains all the code required to re-run the workflow, the input data and the results of executing each module.
-The object `FengPapes` could therefore be saved as a binary file and shared as a reproducible representation of this research.
+The object `Feng_Papes` could therefore be saved as a binary file and shared as a reproducible representation of this research.
 
 Next, we update the workflow to produce an interactive map enabling anyone to inspect the data and predictions on a zoomable map, and to inspect the response curves of the fitted model.
-These outputs are shown in Figure This map is shown in Figure 2, panels B and C.
+These outputs are shown in Figure This map is shown in \ref{fig:feng_papes_plots}, panels B and C.
 
 ```{r fengChangeWorkflow, eval = TRUE, fig.show = "hide", fig.height = 9}
-FengPapesUpdate <- 
-  ChangeWorkflow(workflow = FengPapes,
-                   output = Chain(ResponseCurve(cov = 1),
-                                  InteractiveMap))
+Feng_Papes_Update <- 
+  ChangeWorkflow(workflow = Feng_Papes,
+                   output = Chain(InteractiveMap, 
+                                  ResponseCurve(cov = 1)))
 ```
 
-```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 4.5, fig.width = 4.5}
+```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 3.2}
 # combine the static and interactive maps and first effect plot
@@
 
-# set up layout so that heights are the same for all panels
+# set up layout so that heights are the same for both panels
 plot_height <- min(nrow(r_map1), nrow(r_map), nrow(r_resp))
@@
 gap <- 2
-
-# top row: map1 and resp
-widths_top <- c(width_map1, width_resp)
-widths_top <- round(widths_top * ((200 - (1 * gap)) / sum(widths_top)))
-widths_top <- c(widths_top[1], gap, widths_top[2])
-
-top_row <- rep(1:3, widths_top)
-bottom_row <- rep(5, length(top_row))
-
-heights <- c(plot_height + 1, plot_height)
-heights <- round(heights * ((50 - (1)) / sum(heights)))
-heights <- c(heights[1], 1, heights[2])
-
-top_matrix <- do.call(rbind, replicate(heights[1], top_row, simplify = FALSE))
-middle_matrix <- matrix(4, nrow = heights[2], ncol = ncol(top_matrix))
-bottom_matrix <- matrix(5, nrow = heights[3], ncol = ncol(top_matrix))
-
-mat <- rbind(top_matrix, middle_matrix, bottom_matrix)
+widths <- c(width_map1, width_map, width_resp)
+widths <- round(widths * ((200 - (2 * gap)) / sum(widths)))
+widths <- c(widths[1], gap, widths[2], gap, widths[3])
+mat <- matrix(rep(1:5, widths),
+              nrow = 1)
 layout(mat)
 
-# plot static map
+# plot statis map
 plotRGB(r_map1, maxpixels = Inf)
@@
         side = 3,
-        line = -1.5,
+        line = -3,
+        adj = 0)
+
+# gap
+plot.new()
+
+# plot interactive map
+plotRGB(r_map, maxpixels = Inf)
+
+# add a panel letter
+mtext(text = 'B',
+        side = 3,
+        line = -3,
         adj = 0)
@@
 # add a panel letter
-mtext(text = 'B',
-        side = 3,
-        line = -1.5,
-        adj = 0)
-
-# gap
-plot.new()
-
-# plot interactive map
-plotRGB(r_map, maxpixels = Inf)
-
-# add a panel letter
 mtext(text = 'C',
         side = 3,
-        line = -1.5,
+        line = -3,
         adj = 0)
@@
 
-```{r feng_papes_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Outputs of the workflow objects `FengPapes' and `FengPapesUpdate'.
-A) Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap' module in the workflow `FengPapes'.
+![Outputs of the workflow objects `Feng_Papes` and `Feng_Papes_Update`.
+A) Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap` module in the workflow `Feng_Papes`.
 White areas are masked due to being in the sea or outside the MESS mask.
-B) A response curve produced by the `ResponseCurve' module for the first covariate, bio1 in the workflow `FengPapesUpdate'.
-C) A screenshot of the interactive map produced by the `InteractiveMap' modules in the workflow `FengPapesUpdate', displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results."}
-knitr::include_graphics("figs/combinefengupdate-1.png")
-```
+B) A screenshot of the interactive map produced by the `InteractiveMap` moduels in the workflow `Feng_Papes_Update`, displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results.
+C) A response curve produced by the `ResponseCurve` module for the first covariate, bio1 in the workflow `Feng_Papes_Update`.
+\label{fig:feng_papes_plots}](./figs/combinefengupdate-1.png)
+
+<!-- add the original and the response curves to this figure too -->
 
@@
   # get index to rows in sub_df, update the full dataset and return
-  pres_keep_idx <- as.numeric(rownames(th[[1]]))
+  pres_keep_idx <- rownames(th[[1]])
   .data$df <- rbind(df[pres_idx,][pres_keep_idx, ],
@@
 
-<!-- Possible experiments:
+<!--
+Possible experiments:
 * evaluating with different criteria (e.g. deviance or pseudo-r-squared which measure calibration capacity)
@@
 
-```{r merowworkflowtemp, dpi = 600, eval = TRUE}
+```{r merowworkflowtemp, eval = TRUE}
 Merow_Silander <- workflow(
@@
 
-```{r merowworkflow,dpi = 600, eval = FALSE}
+```{r merowworkflow, eval = FALSE}
 
@@
 
-```{r merowChangeWorkflow, dpi = 600, eval = FALSE}
+```{r merowChangeWorkflow, eval = FALSE}
 Merow_Silander_Spatial <- ChangeWorkflow(
@@
 
-Reproducibility, open-data and open-science are topics that receive ever more attention all across science.
-Journals [@McNutt2016] and funders [@EC2016] are moving towards openly shared data and code resources as a requirement, as was set out in the executive order issued by US President Barack Obama in 2013 [@Obama2013].
-These trends strongly suggest researchers must at least find ways of documenting and disseminating code and data, if not ways that enable end users.
-Numerous platforms for publishing analytical outputs exist (e.g.
-GitHub, FigShare, Dryad, CRAN), but end users will still face the demands of locating, deciphering and managing different resources across different platforms.
-Then, the real challenge of open and reproducible science, that has an increased capacity for innovation, is the challenge of increasing the usability of code and data.
+Tutorials on how to create workflows and modules, as well as full technical details for module developers, are provided as vignettes distributed with `zoon`.
 
-However, “scientific software is developed either by domain scientists who have little training in user interface (UI) or experience (UX) design, or by software engineers who build broad systems that do not suit the niche needs or practices of domain scientists” [@Thomer2016].
-Then, there is a need for environments that can address niche needs (e.g. of a modelling discipline such as SDM) and that address usability issues for diverse outputs produced by diverse scientists.
-The SDM community is large enough to warrant a discipline specific platform such as zoon.
-The zoon framework offers a new opportunity for the SDM community to develop community modelling benchmarks, and resolve scientific and methodological questions in ways that the current culture of publishing cannot achieve.
-It may be that zoon can also be used for other modelling activities that have similar workflows, or use similar data and processing steps.
+The `zoon` packages provides solutions to some of the technical barriers to the ultimate goal of more open, productive, reproducible SDM research.
+Achieving this goal will also depend on overcoming the social hurdle of developing a common tool for comparing and quantitatively evaluating SDMs.
 
-During the development of ZOON we have improvised a user-centred, participatory design process that builds on our prior experiences with software development and soliciting user feedback at 3 stages.
-An ideation workshop set the initial requirements before any code was written, and two further workshops have allowed users input on an initial prototype distributed via GitHub and then a stable R package.
-This is by no means a full UX approach [@Pavelin2012], but it has directly informed the design. 
-Continued user engagement will be vital to developing a critical mass of zoon modules that can incentivise modellers to contribute to community resources that enable evidence-based decisions about best practice.
-An important step would be the publication of a large model inter-comparison within the zoon environment, such as that produced by @Elith2006 or more recently in @Beaumont2016.
-At present, studies such as these might be considered as ‘closed knowledge’ even if they were reproducible from the publication.
-In order to facilitate a user –friendly approach to distributed working, future work will include the development of an online interface for accessing a repository of contributed modules and modifying workflows.
-This interface might also host methodological discourse and openly evaluate proposed best practices in SDM.
-In sum, we hope that the zoon R package represents a step towards a more reproducible ecosystem of SDM software.
+This will require achieving a critical mass of `zoon` modules, so that there is an incentive for species distribution modellers to use the software, and a critical mass of users, so that there is an incentive for SDM methods developers to encode their proposed methods as modules in the community repo.
+At this point `zoon` will provide a common environment in which to evaluate SDM methods and enable the SDM research community to make evidence-based decisions about best practice and the goals for development of the field.
+
+In order to facilitate this social aspect, future work will develop an online platform to enable exploring contributed modules and workflows.
+This platform will also provide an online space to discuss, and openly evaluate, proposed best practices in SDM.
+The zoon R package therefore represents a step towards a more reproducible ecosystem of SDM software.
+
+<!— point to github development site—>
 

### 2016-11-11 ms.Rmd

--- 2016-11-01 ms.Rmd
@@
 The Zoon Project takes a different approach, instead defining only the core architecture of workflows, and interfacing with an open repository of community-contributed software modules.
+
 Users can create new modules for any modelling method and integrate it into another analysis. 
 Similarly to programming languages such as R and Python, facilitating wider development of software will enable the whole research community to develop the software that it requires, rather than rely on the developers of the software they use to do it for them.
+
 Unlike software packages for programming languages however, The Zoon Project enforces a modular structure that ensures interoperability between contributed software modules.

### 2016-11-30 ms.Rmd
Minor edits to text plus add threshold to Feng example

@@
 
-Computational sciences are then presented with a modern version of the reproducibility problem as computational research is largely inaccessible let alone reusable.
+Computational sciences are then presented with a modern version of the reproducibility problem [reference?] as computational research is largely inaccessible let alone reusable.
 If research materials and outputs are shared, they are rarely disseminated in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g.
@@
 In terms of robustness, the CHASTE project for cardiac modelling (@Mirams2013, `cs.ox.ac.uk/chaste`) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling [@Cooper2016].
-These frameworks increase the access, usability and robustness of ultimately scientific research.
+These frameworks increase the access, usability and robustness of scientific research.
 Of course, R users could point to the R environment itself, but the stringent requirements of R packages are likely to prevent much of the SDM community from sharing code in a discoverable or usable format.
@@
 Unlike software packages for programming languages however, The Zoon Project enforces a modular structure that ensures interoperability between contributed software modules.
-This also helps to lower the bar for creating modules (each module is simply an R function with a set of required inputs and outputs) and facilitates automated testing of modules ot ensure software quality.
+This also helps to lower the bar for creating modules (each module is simply an R function with a set of required inputs and outputs) and facilitates automated testing of modules to ensure software quality.
 
@@
 `list` can be used to split the analysis at any point: listing multiple occurrence modules together will run the workflow on different occurrence datasets while listing multiple model modules will fit different models to a shared, identical dataset but then apply output modules to each model separately.
+In cases where modules containing randomness need to be run many times - such as running a workflow with hundreds of simulated datasets - `replicate` can be used.
+This simply creates a `list` of the specified module, repeated a given number of times.
 
-The complete workflow, with multiple modules defined with the `Chain` function and parallel analyses being defined with the `list` function can be visualised by calling `plot` with the workflow object as it's argument.
+The complete workflow, with multiple modules defined with the `Chain` function and parallel analyses being defined with the `list` or `replicate` function can be visualised by calling `plot` with the workflow object as it's argument.
 This displays the names of all modules and the broader structure of the workflow as well as displaying which modules are not available in the online repository (Figure ).
@@
 
+All modules, created by the zoon authors or contributed by others, have documentation explaining their methods and how they are used.
+This documentation is easily accessed in R by using the function `ModuleHelp`.
+This function fetched the documentation from the module repository.
+
 ##### Cross-validation
@@
 `zoon` provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform figshare.
-`ZoonFigshare` takes a workflow object and some minimal metadata and uploads the workflow as an RData object to the user's figshare profile, for others to download, inspect and modify.
+`ZoonFigshare` takes a workflow object and some minimal metadata and uploads the workflow as an RData object, along with a metadata text file, to the user's figshare profile for others to download, inspect and modify.
 
@@
                 model = MaxEnt,
-               output = PrintMap(points = FALSE))
+               output = PrintMap(points = FALSE,
+                                 threshold = 0.05))
 ```

### 2016-12-01 ms.Rmd
Update to thte feng example - in particular the output figure.

@@
                            MESSMask,  
-                              Background(n = 10000, bias = 200), 
+                           Background(n = 10000,
+                                      bias = 200), 
                            Crossvalidate(k = 5)),
@@
            output = PrintMap(points = FALSE,
-                                 threshold = 0.05))
+                             threshold = 0.05,
+                             xlim = c(-130,-70),
+                             ylim = c(20, 50)))
+
 ```
 
-This module plots a static map of the predicted distribution, and the underlying data.
-This map is shown in Figure 2A.
+This workflow plots a static map of the predicted distribution, and the underlying data.
+This map is shown in Figure 2A and corresponds to figure 3 in @Feng2015.
 The resulting workflow contains all the code required to re-run the workflow, the input data and the results of executing each module.
@@
 # combine the static and interactive maps and first effect plot
-
 # load both images
@@
         adj = 0)  
-
 ```

### 2016-12-19 ms.Rmd
re-writing abstract

@@
 
-1. The diverse array of software and methodological research available for species distribution modelling (SDM) hinders independent evaluation of new methods and their dissemination to SDM users.
-
+1.
+The rapid growth of species distribution modelling (SDM) as an ecological discipline has resulted in a large and diverse set of methods and software for constructing and evaluating SDMs.
+The disjointed nature of the current SDM research environment hinders evaluation of new methods, synthesis of current knowledge, and the dissemination of new methods to SDM users.
 2.
-The `zoon` R package encodes SDM analyses as a simple, but fully reproducible workflow of five steps: obtaining occurrence data, obtaining covariate data, pre-processing these data, fitting a model, and generating outputs.
-
+The zoon R package aims to overcome these problems by providing a modular framework for constructing SDM workflows.
+zoon modules are interoperable snippets of R code, each carrying out an SDM method, that zoon combines into a single analysis.
 3.
-Each of these steps is carried out by one or more community-contributed software modules maintained in a version-controlled online repository and built upon existing SDM code form R-packages and the research community.
-
-4. `zoon` workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.
-
-5. We demonstrate `zoon` by recreating SDM analyses from two published research articles as `zoon` workflows, which readers can interrogate and extend.
+Rather than defining these modules, zoon instead draws modules from an open, version-controlled online repository.
+zoon makes it easy for SDM researchers to contribute modules to this repository, enabling others to rapidly add deploy new methods into their own workflow, or to compare a competing methods.
+4.
+The workflows returned by zoon are re-runnable records of the data, code and results of an entire SDM analysis.
+these can be easily reproduced, scrutinized and extended by the whole SDM research community.
+5.
+We introduce zoon, explain how it works and demonstrate how it can be used to construct a complete SDM analyses, create and disseminate a new software module, and perform a methodological comparison study.
 
@@
                              threshold = 0.05,
-                             xlim = c(-130,-70),
+                             xlim = c(-130, -70),
                              ylim = c(20, 50)))

### 2016-12-19 ms.Rmd
massaging the intro

@@
 
-1.
-The rapid growth of species distribution modelling (SDM) as an ecological discipline has resulted in a large and diverse set of methods and software for constructing and evaluating SDMs.
+1. The rapid growth of species distribution modelling (SDM) as an ecological discipline has resulted in a large and diverse set of methods and software for constructing and evaluating SDMs.
 The disjointed nature of the current SDM research environment hinders evaluation of new methods, synthesis of current knowledge, and the dissemination of new methods to SDM users.
-2.
-The zoon R package aims to overcome these problems by providing a modular framework for constructing SDM workflows.
-zoon modules are interoperable snippets of R code, each carrying out an SDM method, that zoon combines into a single analysis.
-3.
-Rather than defining these modules, zoon instead draws modules from an open, version-controlled online repository.
-zoon makes it easy for SDM researchers to contribute modules to this repository, enabling others to rapidly add deploy new methods into their own workflow, or to compare a competing methods.
-4.
-The workflows returned by zoon are re-runnable records of the data, code and results of an entire SDM analysis.
-these can be easily reproduced, scrutinized and extended by the whole SDM research community.
-5.
-We introduce zoon, explain how it works and demonstrate how it can be used to construct a complete SDM analyses, create and disseminate a new software module, and perform a methodological comparison study.
+
+2. The zoon R package aims to overcome these problems by providing a modular framework for constructing reproducible SDM workflows.
+zoon modules are interoperable snippets of R code, each carrying out an SDM method, that zoon combines into a single analysis object.
+
+3. Rather than defining these modules, zoon instead draws modules from an open, version-controlled online repository.
+zoon makes it easy for SDM researchers to contribute modules to this repository, enabling others to rapidly deploy new methods in their own workflows, or to compare alternative methods.
+
+4. Each workflow object created by zoon is a re-runnable record of the data, code and results of an entire SDM analysis.
+This can then be easily shared, scrutinized, reproduced and extended by the whole SDM research community.
+
+5. We explain how zoon works and demonstrate how it can be used to construct a complete reproducible SDM analyses, create and share a new module, and perform a methodological comparison study.
 
@@
 
-Like many areas of quantitative science Species Distribution Modelling (SDM) has grown rapidly [@Barbosa2015] fueled by the availability of diverse modelling software and data.
+Like many areas of quantitative science species distribution modelling (SDM) has grown rapidly, fueled by the availability of diverse modelling software and data [@Barbosa2015].
 Over the last 20 years, SDM has developed into a very large research community and vast literature, and is now one of “the most widely-reviewed topics in the ecological literature” [@Araujo2012].
-Most SDM software, and that enabling other other computational sub-disciplines, focus on  analytical tasks (such as modelling and statistics) rather than enabling scientists to produce reproducible and modifiable analyses that are essential to the scientific enterprise.
+As in other computational sub-disciplines, most SDM software focusses on analytical tasks such as constructing models, rather than enabling scientists to share software, and to produce reproducible and modifiable analyses that are essential to the scientific enterprise.
 
-Computational sciences are then presented with a modern version of the reproducibility problem [reference?] as computational research is largely inaccessible let alone reusable.
-If research materials and outputs are shared, they are rarely disseminated in a readily modifiable or extensible format despite the great potential for distributing these resources digitally (e.g.
-figshare, GitHub).
-The `zoon` R package aims to address these issues for Species Distribution Modelling by allowing users to encode SDM analyses as workflows comprised of independently executable modules.
-The module-workflow structure enables scientists to more easily to create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure 1).
+Consequently, computational sciences like SDM suffer from a modern version of the reproducibility problem, as computational research is largely inaccessible, let alone reusable [reference?].
+When SDM research materials have been shared, they have rarely been disseminated in a format that enables them to be readily modified or adapted.
+This is despite the many available methods for distributing these resources digitally, e,g, via services such as Figshare and GitHub.
 
-A lack of sharing reduces the capacity for innovation and validation in SDM, as well as its capacity to be a "self-correcting process" [@Boulton2012].
+A lack of sharing reduces the capacity for innovation and validation in SDM, as well as its capacity to be a self-correcting process [@Boulton2012].
 For example, a fundamental dispute over the ability of SDM to detect environmental associations [@Beale2008, @Araujo2009, @Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].
-Without the resources being made open the critical differences in these analytical steps are assumed rather than proven.
+Because the resources required to scrutinize and test each piece of research were not shared, the critical analytical differences between these conflicting studies remain unclear.
 
-A second example comes from [@Elith2006], who reported a study of exceptional scope (“16 modelling methods over 226 species from 6 regions of the world”).
-The modelling benchmarks that were reported have been highly influential, yet cannot be updated or modified when new modelling methods are produced as the code and data were not shared.
-A third example comes from meta-analyses [] and methodological reviews [@Yackulic2013], which rely on laborious searches through text publications in the absence of shared code and modelling outputs.
-SDM then has limited abilities to derive truly informative methodological conclusions that can be acted upon.
+The inability to share code and data also prevents broad-scale methodological comparisons.
+The landmark SDM modelling comparison of @Elith2006 has been highly influential in SDM research, having been cited over xxx times.
+However in the decade since that study was published there have been no comparisons of a similar scope, despite a rapid increase in the number of available methods for fitting and evaluating SDMs.
+The community's inability to repeat this comprehensive analysis is due to both lack of access to the original data and the difficulty in learning and applying the many different pieces of software for model fitting.
+Consequently, we currently have a very poor understanding of how different models perform for SDM under different circumstances.
 
-These three examples broadly illustrate why the data and models underpinning SDM need to become more repeatable, reproducible, accessible, modifiable and updateable.
-A productive science relies on the accumulation of knowledge that can be reused and appropriated to advance understanding and resolve methodological issues.
-Thus while many software are pitched as being ‘easy to use’ or ‘user friendly’, this is infrequently the case for the secondary user.
+Our inability to synthesise findings across the SDM literature is bolstered by the discontinuous nature of collaboration in SDM research.
+Most SDM researchers collaborate with researchers who use the same SDM software [@Ahmed2015].
+This limits dissemination of new methods throughout the SDM community due to the need to switch software.
+The existence of these research silos may also hinder dissemination of methodological findings, and contribute to widespread misunderstanding about how apply and interpret different modelling approaches [@Yackulic2013].
 
-This paper introduces version 0.4-22 of the `zoon` R package, which enables users to address issues of compatibility, accessibility, sharing and reproducibility for their SDM analyses.
-Firstly, we describe the modular structure of `zoon` workflows and demonstrate how completed workflows can be shared, evaluated and extended.
-We then illustrate these concepts, and the potential contributions of `zoon` to SDM research, by recreating and extending two recent SDM analyses using `zoon` workflows.
-To assist users a number of tutorials are provided on the GitHub repository and as vignettes distributed with `zoon` – including  guides on creating workflows and modules, as well as full technical details for module developers.
+In order to overcome these problems, the data and models underpinning SDM need to become more accessible, reproducible and modifiable.
+This can be achieved by faciliting access to methods developed across the whole research community [open science reference?], and making it easier to share and modify fully reproducible research objects [open science reference?].
+The zoon R package aims to address these issues for SDM by allowing users to encode entire SDM analyses as reproducible workflows comprised of independently executable, community-contributed modules.
+The module-workflow structure enables scientists to more easily create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure 1).
+
+This paper introduces version 0.4-22 of the zoon R package.
+Firstly, we describe the modular structure of zoon workflows and how they can be constructed, shared, reproduced and modified.
+We then illustrate how these concepts enable better SDM research by reproducing a published SDM analysis; converting a recently proposed method into a zoon module; and extending a published methodological comparison.
 
@@
 
+To assist users a number of tutorials are provided on the GitHub repository and as vignettes distributed with `zoon` – including  guides on creating workflows and modules, as well as full technical details for module developers.
+
 ### References

### 2016-12-19 ms.Rmd
merging software comparison in introduction and tweaking modules and chains etc. sections

@@
 Like many areas of quantitative science species distribution modelling (SDM) has grown rapidly, fueled by the availability of diverse modelling software and data [@Barbosa2015].
-Over the last 20 years, SDM has developed into a very large research community and vast literature, and is now one of “the most widely-reviewed topics in the ecological literature” [@Araujo2012].
+Over the last 20 years, SDM has developed into a very large research community and vast literature.
+SDM is now one of “the most widely-reviewed topics in the ecological literature” [@Araujo2012] and the growth of this literature is accelerating [@Barbosa2015].
+
+The SDM software market is similarly diverse [@Ahmed2015].
+It comprises many software packages developed specifically for SDM such as the BioMod, BioMod2 [@Thuiller2009], dismo [@dismo] and sdm [@Naimi2016] R packages, the SDMtoolbox python package [@Brown2014] and GUI applications including MaxEnt [@Phillips2006], openModeller [@deSouza2011], BioEnsembles [@Diniz2009] and ModEco [@Guo2010].
+A range of other software has also been appropriated for SDM studies, such as statistical software including WinBugs and OpenBugs [@Lunn2000], and more general ecological software such as Domain [@Carpenter1993], Canoco [@TerBraak2002].
+However most SDM users replying to a survey in 2015 [@Ahmed2015] used either MaxEnt [@Phillips2006] or R [@R] as the first choice software for their analyses.
 As in other computational sub-disciplines, most SDM software focusses on analytical tasks such as constructing models, rather than enabling scientists to share software, and to produce reproducible and modifiable analyses that are essential to the scientific enterprise.
@@
 
+### Comparison with existing software
+
+The modular nature of zoon workflows will be familiar to users of other SDM R packages, such as BioMod2 [@Thuiller2009] and sdm [@Naimi2016].
+Whilst BioMod2 modules and sdm modules are created by the developers of those packages and contained in the software itself,zoon's modules are drawn from an open repository that any SDM user can contribute to.
+zoon makes module writing easy for SDM users who are not expert R developers, and provides functionality to help document, test and upload modules to the repository.
+zoon workflows are also designed as reproducible research objects which can be shared, inspected, re-run and modified by swapping in new modules.
+
+This focus on reproducibile and modifiable workflows shares similarities with general-purpose workflow systems such as Kepler [@], Vis-Trails [@], Taverna [@].
+The Taverna platform has recently been extended for biodiversity research via the BIOVEL system [@DeGiovanni2015] and provides some SDM functionality via its ENM Components [@DeGiovanni2016].
+However adoption of these workflow environments in the SDM research community is low.
+Presumably this is due to SDM users’ unfamiliarity with workflow software and the limited number of methods already implemented.
+This hurdle is created by requiring users to embed R code within a workflow software, rather than embedding a workflow system within R.
+
+By constrast with existing software, zoon has been custom-built to improve reproducibility of SDM research in R.
+By providing a single, extensible access point to various SDM software, it is designed to facilitate a more connected and effective SDM research community.
+
+# The `zoon` R package
+
 This paper introduces version 0.4-22 of the zoon R package.
-Firstly, we describe the modular structure of zoon workflows and how they can be constructed, shared, reproduced and modified.
+We describe the modular structure of zoon workflows and how they can be constructed, shared, reproduced and modified.
 We then illustrate how these concepts enable better SDM research by reproducing a published SDM analysis; converting a recently proposed method into a zoon module; and extending a published methodological comparison.
 
-### Reproducible software projects
-
-Most of the SDM users replying to a survey in 2015 [@Ahmed2015] were using either MaxEnt [@Phillips2006] or R [@R] as the first choice software for their analyses (also see @Touchon2016). 
-The software market is diverse, however, including a variety of software developed specifically for SDM (e.g.
-MaxEnt [@Phillips2006], the BioMod and BioMod2 [@Thuiller2009] and dismo [@dismo] packages for R, SDMtoolbox [@Brown2014], openModeller [@deSouza2011], BioEnsembles [@Diniz2009], ModEco [@Guo2010]) and software that has been appropriated for SDM studies, such as statistical software (WinBugs and OpenBugs [@Lunn2000]) and other, more general ecological software  (Domain [@Carpenter1993], Canoco [@TerBraak2002]).
-
-Of the SDM software for multi model comparisons (e.g. OpenModeller, BioEnsembles, ModEco), BioMod (as an open-source R package) has the greatest potential to be modified and re-appropriated.
-In essence, BioMod is an SDM environment accessed from R, rather than necessarily being a SDM tool kit for R.
-The R package called sdm [@Naimi2016] implements a variety of analytical functionality in a simplifying framework.
-There is scope for incorporating new techniques but not necessarily a straight forward system for sharing those contributions.
-The BIOVEL workflow environment [@DeGiovanni2015] is an exception in terms of reproducibility, sharing many of the general principles of `zoon`.
-Adoption of workflow environments (e.g.
-Kepler, Vis-Trails, Taverna) is low however, presumably due to the users’ unfamiliarity, and the requirement of embedding R within a workflow environment rather than embedding a workflow system within R.
-
-Examples of subject specific frameworks are found within the Bioinformatics community, for example the Bioconductor infrastructure that uses R (@Gentleman2004, `bioconductor.org`), and BioJS, a repository for JavaScript components (@Corpas2014, `biojs.net`).
-Both software aim to increase the compatibility and sharing of computational tools in order to increase the reuse and robustness of tools.
-In terms of robustness, the CHASTE project for cardiac modelling (@Mirams2013, `cs.ox.ac.uk/chaste`) has developed into a model repository that offers community benchmarks for the last 50 years of cardiac modelling [@Cooper2016].
-These frameworks increase the access, usability and robustness of scientific research.
-Of course, R users could point to the R environment itself, but the stringent requirements of R packages are likely to prevent much of the SDM community from sharing code in a discoverable or usable format.
-
-### The Zoon Project
-
-The Zoon Project takes a different approach, instead defining only the core architecture of workflows, and interfacing with an open repository of community-contributed software modules.
-
-Users can create new modules for any modelling method and integrate it into another analysis. 
-Similarly to programming languages such as R and Python, facilitating wider development of software will enable the whole research community to develop the software that it requires, rather than rely on the developers of the software they use to do it for them.
-
-Unlike software packages for programming languages however, The Zoon Project enforces a modular structure that ensures interoperability between contributed software modules.
-This also helps to lower the bar for creating modules (each module is simply an R function with a set of required inputs and outputs) and facilitates automated testing of modules to ensure software quality.
-
-# The `zoon` R package
+#### Module types
 
@@
 
-####
-
-#### Module types
-
 The five module types, corresponding to the five key steps of an SDM analysis are:
@@
 
+Users can query the modules hosted on the repository from their R session using the function `GetModuleList()`, and pull up the documentation for any module in the repository with `ModuleHelp()`.
+
 #### Lists, Chains and Replicates
@@
 <!— needs work—>
+
+Many SDM analyses apply more than one method for each of these steps.
+zoon enables more complex workflows by enabling users to pass multiple modules of each type via the List, Chain and Replicate functions.
 
@@
 
-The complete workflow, with multiple modules defined with the `Chain` function and parallel analyses being defined with the `list` or `replicate` function can be visualised by calling `plot` with the workflow object as it's argument.
-This displays the names of all modules and the broader structure of the workflow as well as displaying which modules are not available in the online repository (Figure ).
+The following workflow demonstrates use of the `Chain`s and `list`s to fit three different SDMs to a presence-background dataset for a UK mosquito species, after standardising the covariates.
+The results of the three models are then returned in Shiny app allowing the user to interactively explore the occurrence and covariate data, model summaries and prediction maps.
 
-```{r plotworkflow, eval = TRUE, echo = FALSE, fig.height = 4, fig.width = 9, dpi = 300}
-
+```{r plotworkflow, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 9, dpi = 300}
 work <- workflow(occurrence = UKAnophelesPlumbeus,
-                 covariate  = UKAir,
-                 process    = Chain(Background(n = 70), Crossvalidate(k = 5)),
-                 model      = list(LogisticRegression, RandomForest, QuickGRaF),
-                 output     = PerformanceMeasures)
-
-plot(work)
+                 covariate  = UKBioclim,
+                 process    = Chain(Background(n = 500),
+                                    StandardiseCov),
+                 model      = list(MaxEnt,
+                                   GBM,
+                                   RandomForest),
+                 output     = Appify)
 ```
@@
 
-A `zoonWorkflow` objectStructure of workflow objects:
+A `zoonWorkflow` object
+Structure of workflow objects:
 

### 2016-12-20 ms.Rmd
remove codeface from 'zoon' in text

@@
 ---
-title: "The `zoon` R package for reproducible and shareable species distribution modelling"
+title: "The zoon R package for reproducible and shareable species distribution modelling"
 author: "Nick Golding, Tom August, Tim C.D. Lucas, Emiel Van Loon & Greg McInerny"
@@
 
-# The `zoon` R package
+# The zoon R package
 
@@
 
-The `zoon` R package encodes SDM analyses as a simple workflow of five key steps, with each step encoded as one or more software modules.
+The zoon R package encodes SDM analyses as a simple workflow of five key steps, with each step encoded as one or more software modules.
 Users combine these modules via a call to the `workflow` function, which executes each module in turn, before returning a `zoonWorkflow` object - a shareable, extensible and fully reproducible documentation of the SDM analysis.
-Figure 1 illustrates the structure of a `zoon` workflow comprising the five module types and illustrates how multiple modules may be combined at each stage.
+Figure 1 illustrates the structure of a zoon workflow comprising the five module types and illustrates how multiple modules may be combined at each stage.
 The module types and module combination options are described in more detail below.
@@
 C) The flow diagram implied by listing modules of the same type.
-Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the `zoon` vignette 'Building a module'."}
+Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the zoon vignette 'Building a module'."}
 knitr::include_graphics("figs/diagrams.png")
@@
 
-<!— needs work—>
+<!— remove this, it's covered in the figure —>
 
@@
 In methodological work, it is important to compare new methods or models to benchmarks.
-To divide a `zoon` workflow into multiple parallel analyses the `list` function is used.
+To divide a zoon workflow into multiple parallel analyses the `list` function is used.
 `list` can be used to split the analysis at any point: listing multiple occurrence modules together will run the workflow on different occurrence datasets while listing multiple model modules will fit different models to a shared, identical dataset but then apply output modules to each model separately.
@@
 
-### Inspecting, sharing and extending a `zoon` workflow
+### Inspecting, sharing and extending a zoon workflow
 
@@
 
-`zoon` workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.
-A `zoon` workflow returns an object of class `zoonWorkflow`.
+zoon workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.
+A zoon workflow returns an object of class `zoonWorkflow`.
 This object contains all the information needed to reproduce the analysis.
@@
 
-`zoon` provides functionality to quickly update an existing workflow object (even one created by someone else) by switching out modules.
+zoon provides functionality to quickly update an existing workflow object (even one created by someone else) by switching out modules.
 
@@
 
-As long as a valid module function is defined in an R session, it can be used in workflows in combination with modules downloaded from the `zoon` module repository.
-Whilst this makes it easy to develop modules locally, the strength of `zoon` comes from the ability to upload modules to the online repository so that others can access them.
+As long as a valid module function is defined in an R session, it can be used in workflows in combination with modules downloaded from the zoon module repository.
+Whilst this makes it easy to develop modules locally, the strength of zoon comes from the ability to upload modules to the online repository so that others can access them.
 
 To do this, it first is necessary to provide additional metadata including the module tile and description, module type, author name and definitions for any non-mandatory arguments.
-`zoon` provides the function `BuildModule()` to facilitate entering this metadata and checking that it matches the module.
+zoon provides the function `BuildModule()` to facilitate entering this metadata and checking that it matches the module.
 
-The `zoon` vignette *Building modules* provides a detailed tutorial for building modules of each type.
+The zoon vignette *Building modules* provides a detailed tutorial for building modules of each type.
 
@@
 
-`zoon` provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform figshare.
+zoon provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform figshare.
 `ZoonFigshare` takes a workflow object and some minimal metadata and uploads the workflow as an RData object, along with a metadata text file, to the user's figshare profile for others to download, inspect and modify.
@@
 
-We demonstrate the `zoon` R package by recreating two SDM analyses from published research articles as `zoon` workflows and showing how these workflow objects can easily be extended with new modules.
+We demonstrate the zoon R package by recreating two SDM analyses from published research articles as zoon workflows and showing how these workflow objects can easily be extended with new modules.
 Workflow objects created by these analyses can be accessed at [`figshare.com/articles/zoon_applications_paper_workflows`](http://figshare.com/articles/zoon_applications_paper_workflows).
@@
 
-Such a model can be quickly and easily re-constructed as a `zoon` workflow using modules available in the `zoon` module repository.
+Such a model can be quickly and easily re-constructed as a zoon workflow using modules available in the zoon module repository.
 @Feng2015 used a combination of occurrence data from GBIF, and additional occurrence data manually collected from the published literature.
@@
 @Aiello-Lammens2015 proposed an approach for dealing with spatial sampling bias in presence-only data, by 'thinning' the presence data, and provide an R package `spThin` to implement their procedure [@spThin].
-We can incorporate this approach in a workflow, by defining a simple *process* module that adapts the `zoon` data into the spThin format, uses this package to apply the algorithm, and converts data back into `zoon`'s expected format again:
+We can incorporate this approach in a workflow, by defining a simple *process* module that adapts the zoon data into the spThin format, uses this package to apply the algorithm, and converts data back into zoon's expected format again:
 
@@
 
-To convert this code into a `zoon` module, we need to write it to a standalone file (named `spThin.R`) with the metadata required to build the module documentation.
-The `zoon` function `BuildModule()` helps with this step, and can also run checks to make sure we got everything right:
+To convert this code into a zoon module, we need to write it to a standalone file (named `spThin.R`) with the metadata required to build the module documentation.
+The zoon function `BuildModule()` helps with this step, and can also run checks to make sure we got everything right:
 
@@
 
-This module can now be shared so that others can use it in their `zoon` workflows.
-The best way of sharing it is via the `zoon` modules repository.
+This module can now be shared so that others can use it in their zoon workflows.
+The best way of sharing it is via the zoon modules repository.
 Modules can uploaded there, via the module submission system at [URL TBD].
@@
 
-In this example we illustrate how `zoon` can help contribute to discussions such as these by enabling rapid evaluation of a new method against a variety of datasets and test procedures.
+In this example we illustrate how zoon can help contribute to discussions such as these by enabling rapid evaluation of a new method against a variety of datasets and test procedures.
 
@@
 
-All of these have now been uploaded to the `zoon` modules repository under the names `CarolinaWrenPO`, `CarolinaWrenPA`, `CarolinaWrenRasters` and `MaxLike`.
+All of these have now been uploaded to the zoon modules repository under the names `CarolinaWrenPO`, `CarolinaWrenPA`, `CarolinaWrenRasters` and `MaxLike`.
 
@@
 Then, there is a need for environments that can address niche needs (e.g. of a modelling discipline such as SDM) and that address usability issues for diverse outputs produced by diverse scientists.
-The SDM community is large enough to warrant a discipline specific platform such as `zoon`.
+The SDM community is large enough to warrant a discipline specific platform such as zoon.
 The Zoon Project offers a new opportunity for the SDM community to develop community modelling benchmarks, and resolve scientific and methodological questions in ways that the current culture of publishing cannot achieve.
@@
 
-During the development of `zoon` we have improvised a user-centred, participatory design process that builds on our prior experiences with software development and soliciting user feedback at 3 stages.
+During the development of zoon we have improvised a user-centred, participatory design process that builds on our prior experiences with software development and soliciting user feedback at 3 stages.
 An ideation workshop set the initial requirements before any code was written, and two further workshops have allowed users input on an initial prototype distributed via GitHub and then a stable R package.
 This is by no means a full UX approach [@Pavelin2012], but it has directly informed the design. 
-Continued user engagement will be vital to developing a critical mass of `zoon` modules that can incentivise modellers to contribute to community resources that enable evidence-based decisions about best practice.
-An important step would be the publication of a large model inter-comparison within the `zoon` environment, such as that produced by @Elith2006 or more recently in @Beaumont2016.
+Continued user engagement will be vital to developing a critical mass of zoon modules that can incentivise modellers to contribute to community resources that enable evidence-based decisions about best practice.
+An important step would be the publication of a large model inter-comparison within the zoon environment, such as that produced by @Elith2006 or more recently in @Beaumont2016.
 At present, studies such as these might be considered as ‘closed knowledge’ even if they were reproducible from the publication.
@@
 This interface might also host methodological discourse and openly evaluate proposed best practices in SDM.
-In sum, we hope that the `zoon` R package represents a step towards a more reproducible ecosystem of SDM software.
+In sum, we hope that the zoon R package represents a step towards a more reproducible ecosystem of SDM software.
 
-To assist users a number of tutorials are provided on the GitHub repository and as vignettes distributed with `zoon` – including  guides on creating workflows and modules, as well as full technical details for module developers.
+To assist users a number of tutorials are provided on the GitHub repository and as vignettes distributed with zoon – including  guides on creating workflows and modules, as well as full technical details for module developers.
 

### 2016-12-20 ms.Rmd
slimming down software comparison section

@@
 
-### Comparison with existing software
-
+By constrast with existing software, zoon has been custom-built to improve reproducibility of SDM research in R.
 The modular nature of zoon workflows will be familiar to users of other SDM R packages, such as BioMod2 [@Thuiller2009] and sdm [@Naimi2016].
-Whilst BioMod2 modules and sdm modules are created by the developers of those packages and contained in the software itself,zoon's modules are drawn from an open repository that any SDM user can contribute to.
-zoon makes module writing easy for SDM users who are not expert R developers, and provides functionality to help document, test and upload modules to the repository.
-zoon workflows are also designed as reproducible research objects which can be shared, inspected, re-run and modified by swapping in new modules.
-
-This focus on reproducibile and modifiable workflows shares similarities with general-purpose workflow systems such as Kepler [@], Vis-Trails [@], Taverna [@].
-The Taverna platform has recently been extended for biodiversity research via the BIOVEL system [@DeGiovanni2015] and provides some SDM functionality via its ENM Components [@DeGiovanni2016].
-However adoption of these workflow environments in the SDM research community is low.
-Presumably this is due to SDM users’ unfamiliarity with workflow software and the limited number of methods already implemented.
-This hurdle is created by requiring users to embed R code within a workflow software, rather than embedding a workflow system within R.
-
-By constrast with existing software, zoon has been custom-built to improve reproducibility of SDM research in R.
-By providing a single, extensible access point to various SDM software, it is designed to facilitate a more connected and effective SDM research community.
+zoon differs from these packages in that zoon modules are drawn from an open repository that any SDM user can contribute to, and zoon makes it easy for SDM users who are not expert R developers to contribute.
+zoon's focus on reproducibile and modifiable workflows also shares similarities with general-purpose workflow systems such as Taverna [@] and its extensions for biodiversity research via the BIOVEL system [@DeGiovanni2015,@DeGiovanni2016].
+However unlike these workflow environments in which users may embed R code, zoon embeds an SDM-centric workflow system within R.
+zoon therefore provides the benefits of workflow systems to SDM users without them needing to learn complex workflow systems.
 

### 2016-12-20 ms.Rmd
more swingeing cuts to the introduction

@@
 
-Like many areas of quantitative science species distribution modelling (SDM) has grown rapidly, fueled by the availability of diverse modelling software and data [@Barbosa2015].
-Over the last 20 years, SDM has developed into a very large research community and vast literature.
-SDM is now one of “the most widely-reviewed topics in the ecological literature” [@Araujo2012] and the growth of this literature is accelerating [@Barbosa2015].
+Species distribution modelling (SDM) has grown rapidly over the last 20 years.
+It is now one of “the most widely-reviewed topics in the ecological literature” [@Araujo2012] and the growth of this literature is still accelerating [@Barbosa2015].
+The SDM software market is similarly large and diverse [@Ahmed2015].
+Whilst most SDM users rely either on the MaxEnt standalone application [@Phillips2006] or the R programming language [@R] as the first choice software for their analyses [@Ahmed2015], there are many competing SDM-specific R packages, each implementing different approaches to fitting SDMs [@Thuiller2009; @dismo; @Naimi2016], as well as a range of alternative standalone applications.
 
-The SDM software market is similarly diverse [@Ahmed2015].
-It comprises many software packages developed specifically for SDM such as the BioMod, BioMod2 [@Thuiller2009], dismo [@dismo] and sdm [@Naimi2016] R packages, the SDMtoolbox python package [@Brown2014] and GUI applications including MaxEnt [@Phillips2006], openModeller [@deSouza2011], BioEnsembles [@Diniz2009] and ModEco [@Guo2010].
-A range of other software has also been appropriated for SDM studies, such as statistical software including WinBugs and OpenBugs [@Lunn2000], and more general ecological software such as Domain [@Carpenter1993], Canoco [@TerBraak2002].
-However most SDM users replying to a survey in 2015 [@Ahmed2015] used either MaxEnt [@Phillips2006] or R [@R] as the first choice software for their analyses.
-As in other computational sub-disciplines, most SDM software focusses on analytical tasks such as constructing models, rather than enabling scientists to share software, and to produce reproducible and modifiable analyses that are essential to the scientific enterprise.
+As a result, the SDM community is divided into research silos, with groups of researchers collaborating primarily with others who use the same software, even within specific R packages [@Ahmed2015], presenting a barrier to disseminating new SDM findings and methods.
+This is compounded by the almost all available SDM software being focussed on analytical tasks such as constructing models, rather than enabling scientists to produce analyses that others can reproduce and modify.
+The inability of SDM researchers to reproduce, scrutinise, and build on others' research prevents rigorous peer review and synthesis of research findings across studies, as well as its capacity to be a self-correcting process [@Boulton2012].
 
-Consequently, computational sciences like SDM suffer from a modern version of the reproducibility problem, as computational research is largely inaccessible, let alone reusable [reference?].
-When SDM research materials have been shared, they have rarely been disseminated in a format that enables them to be readily modified or adapted.
-This is despite the many available methods for distributing these resources digitally, e,g, via services such as Figshare and GitHub.
+For example, a fundamental dispute over the ability of SDM to detect environmental associations [@Beale2008, @Araujo2009, @Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].
+Similarly, in the past decade the SDM community has failed to carry out any large scale comparisons of SDM methods, still relying on the work of @Elith2006 to make decisions, despite the development of hundreds of new methods as well as improved evaluation procedures [@Roberts2016].
+The community's inability to repeat this comprehensive analysis is due to both lack of access to comparable data and the difficulty in learning and applying the many different pieces of software for model fitting.
 
-A lack of sharing reduces the capacity for innovation and validation in SDM, as well as its capacity to be a self-correcting process [@Boulton2012].
-For example, a fundamental dispute over the ability of SDM to detect environmental associations [@Beale2008, @Araujo2009, @Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].
-Because the resources required to scrutinize and test each piece of research were not shared, the critical analytical differences between these conflicting studies remain unclear.
+The SDM community's inability to compare models and repeat analyses has left the field unable to decisively answer even basic questions about how best to do SDM.
+It may also have contributed to widespread misunderstandings about how apply and interpret different modelling approaches [@Yackulic2013].
+In order to overcome these problems, the data and methods underpinning SDM research need to be made more accessible, reproducible and modifiable by the whole research community.
+This can be achieved by faciliting access to methods[open science reference?], and making it easier to share and modify fully reproducible research objects [open science reference?].
 
-The inability to share code and data also prevents broad-scale methodological comparisons.
-The landmark SDM modelling comparison of @Elith2006 has been highly influential in SDM research, having been cited over xxx times.
-However in the decade since that study was published there have been no comparisons of a similar scope, despite a rapid increase in the number of available methods for fitting and evaluating SDMs.
-The community's inability to repeat this comprehensive analysis is due to both lack of access to the original data and the difficulty in learning and applying the many different pieces of software for model fitting.
-Consequently, we currently have a very poor understanding of how different models perform for SDM under different circumstances.
-
-Our inability to synthesise findings across the SDM literature is bolstered by the discontinuous nature of collaboration in SDM research.
-Most SDM researchers collaborate with researchers who use the same SDM software [@Ahmed2015].
-This limits dissemination of new methods throughout the SDM community due to the need to switch software.
-The existence of these research silos may also hinder dissemination of methodological findings, and contribute to widespread misunderstanding about how apply and interpret different modelling approaches [@Yackulic2013].
-
-In order to overcome these problems, the data and models underpinning SDM need to become more accessible, reproducible and modifiable.
-This can be achieved by faciliting access to methods developed across the whole research community [open science reference?], and making it easier to share and modify fully reproducible research objects [open science reference?].
-The zoon R package aims to address these issues for SDM by allowing users to encode entire SDM analyses as reproducible workflows comprised of independently executable, community-contributed modules.
+The zoon R package has been developed specifically to improve reproducibility and comparability of SDM in R by allowing users to encode entire SDM analyses as reproducible workflows of independently executable, community-contributed modules.
 The module-workflow structure enables scientists to more easily create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure 1).
-
-By constrast with existing software, zoon has been custom-built to improve reproducibility of SDM research in R.
-The modular nature of zoon workflows will be familiar to users of other SDM R packages, such as BioMod2 [@Thuiller2009] and sdm [@Naimi2016].
-zoon differs from these packages in that zoon modules are drawn from an open repository that any SDM user can contribute to, and zoon makes it easy for SDM users who are not expert R developers to contribute.
-zoon's focus on reproducibile and modifiable workflows also shares similarities with general-purpose workflow systems such as Taverna [@] and its extensions for biodiversity research via the BIOVEL system [@DeGiovanni2015,@DeGiovanni2016].
-However unlike these workflow environments in which users may embed R code, zoon embeds an SDM-centric workflow system within R.
-zoon therefore provides the benefits of workflow systems to SDM users without them needing to learn complex workflow systems.
+Whilst zoon's modular nature is similar to other SDM R packages such as BioMod2 [@Thuiller2009] and sdm [@Naimi2016], zoon instead pulls each module from an open repository that any SDM user can contribute to, and makes it easy for non-developers to contribute modules.
+zoon's focus on reproducibile and modifiable workflows is inspired by workflow systems such as Taverna [@Wolstencroft2013a] and the BIOVEL system [@DeGiovanni2015; @DeGiovanni2016], but embeds an SDM-specific workflow system within R, making it much more accessible to the SDM community.
 
@@
 
-This paper introduces version 0.4-22 of the zoon R package.
+This paper introduces version 0.5-3 of the zoon R package.
 We describe the modular structure of zoon workflows and how they can be constructed, shared, reproduced and modified.
-We then illustrate how these concepts enable better SDM research by reproducing a published SDM analysis; converting a recently proposed method into a zoon module; and extending a published methodological comparison.
+We then illustrate how these concepts enable better SDM research by reproducing a published SDM analysis; converting a recently proposed method into a zoon module; and performing a methodological comparison.
 
@@
 The module types and module combination options are described in more detail below.
-
-<!— link to powerpoint doc: https://www.dropbox.com/s/p7rqww2v0h3a7np/diagrams.pptx?dl=0-->
 
@@
 ```
-
-The five module types, corresponding to the five key steps of an SDM analysis are:
-
-<!— remove this, it's covered in the figure —>
-
-- *Occurrence* - Usually presence-absence data or presence-only data, though abundance data is also used.
-- *Covariates* - Predictor variables or covariates (typically environmental covariates) are required and the values of these covariates, at the locations of the occurrence data, must be extracted.
-- *Process* - Processes applied to the occurrence and covariate data.
-These processes include data cleaning, data thinning to account for spatial biases, feature selection using PCA or association tests and the splitting of data into training and test sets or cross validation folds.
-- *Model* - Once the data has been suitable manipulated a model is fitted to estimate the relationships between the covariates and occurrence data.
-These models include simple statistical models such as GLMs as well as modern, flexible machine-learning methods such as MaxEnt and boosted regression trees.
-- *Output* - The predictive power and goodness of fit must be assessed and the model parameters or response curves must be examined.
-The model is likely to be used to predict species occurrence, either in the vicinity of the occurrence data or elsewhere, or into the past or future.
-
-Users can query the modules hosted on the repository from their R session using the function `GetModuleList()`, and pull up the documentation for any module in the repository with `ModuleHelp()`.
 

### 2016-12-20 ms.Rmd
polishing module section

@@
 
-# The zoon R package
-
 This paper introduces version 0.5-3 of the zoon R package.
@@
 
-#### Module types
+# Building a workflow
 
-The zoon R package encodes SDM analyses as a simple workflow of five key steps, with each step encoded as one or more software modules.
+zoon encodes SDM analyses as a simple workflow of five key steps: obtaining occurrence data; obtaining covariate data; applying some processes to these data; fitting one or more models; and generating any outputs (Figure 1A).
+Each of these steps is encoded as one or more software modules of a given type.
 Users combine these modules via a call to the `workflow` function, which executes each module in turn, before returning a `zoonWorkflow` object - a shareable, extensible and fully reproducible documentation of the SDM analysis.
-Figure 1 illustrates the structure of a zoon workflow comprising the five module types and illustrates how multiple modules may be combined at each stage.
-The module types and module combination options are described in more detail below.
+Figure 1B illustrates how data and outputs are passed between modules of the five different types in a a zoon workflow.
 
 ```{r flow_diagram, echo = FALSE, out.width = "375px", fig.cap = "The modular SDM structure encoded by a zoon workflow.
-A) Flow diagram representing the five module types and intermediate objects: df - an occurrence dataframe; ras - a RasterStack object; mod - a ZoonModel object; ?
-- any user-defined inputs or outputs.
-B) The flow diagram implied by chaining modules of the same type.
-C) The flow diagram implied by listing modules of the same type.
-Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the zoon vignette 'Building a module'."}
+A) Description of the five module types.
+B) Flow diagram illustrating how objects are passed between different module types: `data frame' - an occurrence dataframe; `ras' - a RasterStack object; `mod' - a ZoonModel object; `?' - any user-defined inputs or outputs.
+C) The flow diagram implied by chaining two `process' modules.
+D) The flow diagram implied by listing three `model' modules.
+Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the zoon vignette `Building a module'."}
 knitr::include_graphics("figs/diagrams.png")
@@
 
-#### Lists, Chains and Replicates
+## Lists, Chains and Replicates
 
-<!— needs work—>
+Many SDM analyses apply more than one method in one of the five SDM steps, for example to combine multiple covariates and processing steps, to compare various models on the same dataset, or to run the same procedure for several species.
+zoon enables these more complex workflows by enabling users to pass multiple modules of each type via the `list`, `Chain` and `Replicate` functions.
 
-Many SDM analyses apply more than one method for each of these steps.
-zoon enables more complex workflows by enabling users to pass multiple modules of each type via the List, Chain and Replicate functions.
+### Chain
 
-To combine multiple modules of the same type we provide the `Chain` command.
-For occurrence and covariate modules, this command takes multiple modules and simply combines the data acquired by each module.
-Chained process models are run sequentially.
-For example if a user wants to generate background or pseudo absence data and then split the data into crossvalidation folds, modules implementing these two seperate process would be chained in that order.
-Finally, chained output modules are simply all run seperately allowing the user to create multiple maps and summary figures, calcualte performance metrics and create other model outputs in one workflow.
-Model modules cannot be chained.
+`Chain` runs multiple modules of the same type *sequentially*, as illustrated in Figure 1C.
+For example, in a presence-only analysis a chain of two process modules could be used to generate background data and then to standardise the covariate rasters using the presence/background dataset.
+Chains can also be applied to occurrence or covariate modules, to combine multiple datasets (e.g. occurrences from different databases) into one.
+Chaining output modules simply runs each module separately, allowing the user to create multiple maps and summary figures, calcualte performance metrics and create other model outputs in one workflow.
+Model modules are the only module type that may not be chained, since their outputs and inputs are always different.
 
-Many SDM analyses require running similar workflow in parallel. 
-In applied settings this may be running the same analysis for multiple species.
-In methodological work, it is important to compare new methods or models to benchmarks.
-To divide a zoon workflow into multiple parallel analyses the `list` function is used.
-`list` can be used to split the analysis at any point: listing multiple occurrence modules together will run the workflow on different occurrence datasets while listing multiple model modules will fit different models to a shared, identical dataset but then apply output modules to each model separately.
-In cases where modules containing randomness need to be run many times - such as running a workflow with hundreds of simulated datasets - `replicate` can be used.
-This simply creates a `list` of the specified module, repeated a given number of times.
+### list
 
-The following workflow demonstrates use of the `Chain`s and `list`s to fit three different SDMs to a presence-background dataset for a UK mosquito species, after standardising the covariates.
+`list` splits a workflow, with each path using a different one of the listed modules, as illustrated in Figure 1D.
+For example listing multiple model modules would take the same occurrence and covariate data and fit three separate models, applying the same output modules to each model separately.
+Lists can also be used to run the same SDM procedure for multiple species, or for multiple sets of covariates, or to compare different process modules.
+Listing output modules has the same effect as chaining them, each module is applied separately.
+
+### Replicate
+
+Some steps in SDM analyses are stochastic and need to be run multiple times; `Replicate` enables this by generating a list with one module repeated a given number of times.
+`Replicate` could be used to run the same workflow for hundreds of simulated occurrence or covariate datasets, to generate multiple bootstraps for modelling, or to fit models with stochastic elements
+
+### Example workflow
+
+The following workflow demonstrates use of the chains and lists to fit three different SDMs to a presence-background dataset for a UK mosquito species, after standardising the covariates.
 The results of the three models are then returned in Shiny app allowing the user to interactively explore the occurrence and covariate data, model summaries and prediction maps.
@@
 
-### Inspecting, sharing and extending a zoon workflow
+## Inspecting, sharing and extending a zoon workflow
 
@@
 To do this, it first is necessary to provide additional metadata including the module tile and description, module type, author name and definitions for any non-mandatory arguments.
-zoon provides the function `BuildModule()` to facilitate entering this metadata and checking that it matches the module.
+zoon provides the function `BuildModule` to facilitate entering this metadata and checking that it matches the module.
 
@@
 To convert this code into a zoon module, we need to write it to a standalone file (named `spThin.R`) with the metadata required to build the module documentation.
-The zoon function `BuildModule()` helps with this step, and can also run checks to make sure we got everything right:
+The zoon function `BuildModule` helps with this step, and can also run checks to make sure we got everything right:
 

### 2016-12-21 ms.Rmd
first crack at example 3

@@
 
-#### Example 3. Evaluating MaxLike
+#### Example 3. What makes MaxEnt effective
 
-@Royle2012 proposed a model for fitting SDMs to presence-only data, which they suggest is able to estimate the absolute (rather than relative) probability of species presence in each grid cell.
-This model and the associated claims have been the subject of several research articles evaluating the model against both real-world and simulated datasets [@Fitzpatrick2013;@Merow2014;@Hastie2013;@Phillips2013].
+MaxEnt is one of the most popular models for presence-only SDM [@Ahmed2015], in part because it performed very well in the comprehensive comparison of @Elith2006.
+Unlike most other models use in SDM, MaxEnt considers occurrence records as arising from some probability distribution over space and attempts to model that distribution, subject to a 'maximum entropy' constraint [@Phillips2006].
+Subsequent research has shown that although it approaches the problem from a diffent perspective, the model MaxEnt fits is in fact equivalent to Poisson point process model [@Renner?].
+Moreover, so long as the number of background points large, this model is essentially the same as a logistic regression model [@?].
+Given MaxEnt's treatment of the data is not fundamentally different than other common models, @Renner2013 and others have suggested that MaxEnt's predictive power is instead due to the array of features it constructs, and the fact that it applies regularisation when fitting the model.
+Whilst the various features (candidate covariate transformations) increase the potential flexibility of the model, regularisation acts to prevent MaxEnt from overfitting to the data.
 
-In this example we illustrate how zoon can help contribute to discussions such as these by enabling rapid evaluation of a new method against a variety of datasets and test procedures.
+Here we illustrate how zoon can facilitate SDM methodological comparisons by fitting MaxEnt models with and without regularisation and feature construction.
+We fit MaxEnt with both the usual java implementation, and an implementation recently provided in the `maxnet` R package [@maxnet] which approximates the full MaxEnt model via a regularised logistic regression.
 
-For this example, it was necessary to create three new modules: two to load the presence/absence and presence-only Carolina Wren datasets from @Royle2012; and one to implement the MaxLike model.
+The following zoon workflow fits five different models to a presence-background dataset, plots a prediciton map generated by each and evaluates the predictive performance of each model against a presence-absence dataset.
+The presence-only and presence-absence data and covariates are for the Caroline wren in the USA, as analysed by @Royle2012 and @Merow2014, and distributed via the `maxlike` R package [@maxlike].
+The five models compared are: the java implementation of MaxEnt, the `maxnet` implementation (via the `MaxNet` zoon module), the `maxnet` version without regularisation, without features, and without regularisation or features.
+As in @Merow2014, each model is trained on the same randomly selected subset of 500 occurrence records, with 10,000 randomly selected background records.
+Statistics of the predictive performance of each fitted model are then calculated against the full presence-absence dataset, and prediction maps plotted in Figure 3.
 
-<!-- Possible experiments:
-* evaluating with different criteria (e.g. deviance or pseudo-r-squared which measure calibration capacity)
-* fitting the same models with GBIF data and evaluating against the BBS PA data,
-* fitting/evaluating on spatially-stratified holdout data
-* fitting the models with other species
-  (choose only one of these)
-  -->
-
-All of these have now been uploaded to the zoon modules repository under the names `CarolinaWrenPO`, `CarolinaWrenPA`, `CarolinaWrenRasters` and `MaxLike`.
-
-What the workflow looks like:
-
-```{r merowworkflowtemp, dpi = 600, eval = TRUE}
-Merow_Silander <- workflow(
+```{r maxnet_comparison, dpi = 600, fig.show = "hide", eval = TRUE}
+maxnet_comparison <- workflow(
   occurrence = CarolinaWrenPO,
   covariate  = CarolinaWrenRasters,
-    process = Background(n = 10000),
-    model = list(LogisticRegression, MaxEnt, MaxLike),
-    output = Chain(PerformanceMeasures, PrintMap(points = FALSE)))
+  process    = Chain(SubsampleOccurrence(500),
+                     Background(n = 10000),
+                     CarolinaWrenValidation),
+  model      = list(MaxEnt(args = 'threshold=false'),
+                    MaxNet,
+                    MaxNet(regmult = 0.005),
+                    MaxNet(features = 'l'),
+                    MaxNet(regmult = 0.005, features = 'l')),
+  output     = Chain(PerformanceMeasures,
+                     PrintMap(points = FALSE)))
 ```
 
-```{r merowworkflow,dpi = 600, eval = FALSE}
+```{r maxnet_plotting, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 9, fig.width = 1.5}
+# plot the figure to disk, then plot in a separate chunk with a legend
 
-Merow_Silander <- workflow(
-    occurrence = CarolinaWrenPO,
-    covariate = CarolinaWrenRasters,
-    process = Chain(Transform(trans = 'square',
-                                 replace = FALSE), 
-                       Background(n = 10000)),
-    model = list(LogisticRegression, MaxEnt, MaxLike),
-    output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap(points = FALSE), Coefficients))
+par(mfrow = c(5, 1))
+
 ```
 
-So we ran it again with disc-based spatial stratification:
+```{r maxnet_plots, cache = FALSE, echo = FALSE, out.width = "150px", fig.cap = ""}
+# knitr::include_graphics("figs/maxnet_plotting-1.png")
+```
 
-```{r merowChangeWorkflow, dpi = 600, eval = FALSE}
-Merow_Silander_Spatial <- ChangeWorkflow(
-    merow_and_silander,
-    process = Chain(..., PartitionDisc))
+```{r maxent_aucs, eval = TRUE, echo = FALSE}
+# AUCs, not quite sure what to do with these - embed them in the text somehow?
+perflist <- maxnet_comparison$report[seq(1, 10, by = 2)]
+AUCs <- sapply(perflist, function(x) x$auc)
+names(AUCs) <- c('MaxEnt',
+                 'MaxNet',
+                 'MaxNet (no regularization)',
+                 'MaxNet (no features)',
+                 'MaxNet (no regularization or features)')
+AUCs <- round(AUCs, digits = 4)
 ```
+
+The AUC statistics calculated against the presence-absence data were:
+```{r results = 'asis', echo = FALSE}
+cat(paste(names(AUCs), AUCs, collapse = '; '))
+cat('.')
+```
+Readers interested in other statistics, viewing response curves or adjusting the calidation method can easily reproduce and modify the analysis either by re-running the above code block or by downloading the workflow object from: <URL TBD>
 

### 2016-12-23 ms.Rmd
plotting for Example 3, and tidying up first section

--- 2016-12-20 ms.Rmd
@@
 
+The following code uses `workflow` to run simple a presence-background SDM for a mosquito species in the UK, fitting a MaxEnt model with default settings and 500 randomly placed background points.
+
+```{r plotworkflow, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 9, dpi = 300}
+mosquito1 <- workflow(occurrence = UKAnophelesPlumbeus,
+                      covariate  = UKBioclim,
+                      process    = Background(n = 500),
+                      model      = MaxEnt,
+                      output     = InteractiveMap)
+```
+
 ## Lists, Chains and Replicates
@@
 
-### Example workflow
+###
 
-The following workflow demonstrates use of the chains and lists to fit three different SDMs to a presence-background dataset for a UK mosquito species, after standardising the covariates.
-The results of the three models are then returned in Shiny app allowing the user to interactively explore the occurrence and covariate data, model summaries and prediction maps.
+The following workflow uses a chain to standardise the covariate rasters as well as generating background records, and uses a list to fit three different models: MaxEnt, Boosted Regression Trees (GBM) and RandomForest.
+The results of the three models are then returned in three Shiny apps allowing the user to interactively explore the occurrence and covariate data, model summaries and prediction maps.
 
-```{r plotworkflow, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 9, dpi = 300}
-work <- workflow(occurrence = UKAnophelesPlumbeus,
+```{r workflow2, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 9, dpi = 300}
+mosquito2 <- workflow(occurrence = UKAnophelesPlumbeus,
                       covariate  = UKBioclim,
@@
 
-We demonstrate the zoon R package by recreating two SDM analyses from published research articles as zoon workflows and showing how these workflow objects can easily be extended with new modules.
-Workflow objects created by these analyses can be accessed at [`figshare.com/articles/zoon_applications_paper_workflows`](http://figshare.com/articles/zoon_applications_paper_workflows).
+Next we demonstrate how zoon can be used in practice for running SDMs, creating and sharing new modules, and performing a methods comparison.
+The workflow objects created by these analyses can be accessed at [`figshare.com/articles/zoon_applications_paper_workflows`](http://figshare.com/articles/zoon_applications_paper_workflows).
 We encourage readers to download, interrogate and alter these workflows for themselves.
-Full code and metadata for all of the modules used in the examples below can be found at [`github.com/zoonproject/modules/R`](https://github.com/zoonproject/modules/R)
 
@@
 Next, we update the workflow to produce an interactive map enabling anyone to inspect the data and predictions on a zoomable map, and to inspect the response curves of the fitted model.
-These outputs are shown in Figure This map is shown in Figure 2, panels B and C.
+These outputs are shown in Figure 2, panels B and C.
 
@@
 ```{r feng_papes_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Outputs of the workflow objects `FengPapes' and `FengPapesUpdate'.
-A) Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap' module in the workflow `FengPapes'.
+A) Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap' module in the workflow `FengPapes' which encodes a whole published analysis.
 White areas are masked due to being in the sea or outside the MESS mask.
-B) A response curve produced by the `ResponseCurve' module for the first covariate, bio1 in the workflow `FengPapesUpdate'.
-C) A screenshot of the interactive map produced by the `InteractiveMap' modules in the workflow `FengPapesUpdate', displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results."}
+B) A response curve produced by the `ResponseCurve' module for the first covariate, bio1 in the workflow `FengPapesUpdate', which modifies the original analysis workflow.
+C) A screenshot of the interactive map produced by the `InteractiveMap' modules in the workflow `FengPapesUpdate', displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results.
+Any SDM analysis distributed as a zoon workflow can be easily be explored and scrutinized by modifying its output modules using the function `ChangeWorkflow'."}
 knitr::include_graphics("figs/combinefengupdate-1.png")
@@
 This module can now be shared so that others can use it in their zoon workflows.
-The best way of sharing it is via the zoon modules repository.
-Modules can uploaded there, via the module submission system at [URL TBD].
+Modules can be uploaded to the zoon modules repository via the online submission system at [URL TBD].
 
-#### Example 3. Evaluating MaxLike
+#### Example 3. What makes MaxEnt effective
 
-@Royle2012 proposed a model for fitting SDMs to presence-only data, which they suggest is able to estimate the absolute (rather than relative) probability of species presence in each grid cell.
-This model and the associated claims have been the subject of several research articles evaluating the model against both real-world and simulated datasets [@Fitzpatrick2013;@Merow2014;@Hastie2013;@Phillips2013].
+The popular MaxEnt SDM model has recently been shown to be equivalent to a Poisson point process model [@Renner2013], and to be closely approximated by a logistic regression model with weights applied to background datapoints [@Warton2010; @Fithian2013; @Renner2015].
+Given this close correspondence between MaxEnt and logistic regression, @Renner2013 and others have suggested that MaxEnt's superior predictive performance [@Elith2006] is most likely due to the array of features (candidate covariate transformations) it constructs and its use of regularisation to prevent overfitting.
 
-In this example we illustrate how zoon can help contribute to discussions such as these by enabling rapid evaluation of a new method against a variety of datasets and test procedures.
+zoon enables us to investigate this hypothesis by fitting MaxEnt models with and without regularisation and feature construction.
+We fit MaxEnt with both the usual java implementation as well as the downweighted logistic regression approximation provided in the `maxnet` R package [@maxnet].
+We also compare these models with a standard logistic regression model which does not apply the downweighting step.
+The following workflow fits these models to a previously published presence-background data on the Caroline wren in the USA [@Royle2012; @maxlike], maps their predictions (Figure 3) and evaluates the predictive performance of each model against the full presence-absence dataset by AUC.
 
-For this example, it was necessary to create three new modules: two to load the presence/absence and presence-only Carolina Wren datasets from @Royle2012; and one to implement the MaxLike model.
-
-<!-- Possible experiments:
-* evaluating with different criteria (e.g. deviance or pseudo-r-squared which measure calibration capacity)
-* fitting the same models with GBIF data and evaluating against the BBS PA data,
-* fitting/evaluating on spatially-stratified holdout data
-* fitting the models with other species
-  (choose only one of these)
-  -->
-
-All of these have now been uploaded to the zoon modules repository under the names `CarolinaWrenPO`, `CarolinaWrenPA`, `CarolinaWrenRasters` and `MaxLike`.
-
-What the workflow looks like:
-
-```{r merowworkflowtemp, dpi = 600, eval = TRUE}
-Merow_Silander <- workflow(
+```{r maxnet_comparison, dpi = 600, fig.show = "hide", eval = TRUE, fig.height = 7, fig.width = 9}
+maxnet_comparison <- workflow(
   occurrence = CarolinaWrenPO,
   covariate  = CarolinaWrenRasters,
-    process = Background(n = 10000),
-    model = list(LogisticRegression, MaxEnt, MaxLike),
-    output = Chain(PerformanceMeasures, PrintMap(points = FALSE)))
+  process    = Chain(SubsampleOccurrence(500),
+                     Background(n = 10000),
+                     CarolinaWrenValidation),
+  model      = list(MaxEnt(args = 'threshold=false'),
+                    MaxNet,
+                    MaxNet(regmult = 0.005),
+                    MaxNet(features = 'l'),
+                    MaxNet(regmult = 0.005, features = 'l'),
+                    LogisticRegression),
+  output     = Chain(PrintMap(points = FALSE),
+                     PerformanceMeasures))
 ```
 
-```{r merowworkflow,dpi = 600, eval = FALSE}
+```{r maxnet_plotting, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 6, fig.width = 9}
 
-Merow_Silander <- workflow(
-    occurrence = CarolinaWrenPO,
-    covariate = CarolinaWrenRasters,
-    process = Chain(Transform(trans = 'square',
-                                 replace = FALSE), 
-                       Background(n = 10000)),
-    model = list(LogisticRegression, MaxEnt, MaxLike),
-    output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap(points = FALSE), Coefficients))
+# plot the figure to disk, then plot in a separate chunk with a legend
+par(mfrow = c(3, 2),
+    mar = c(6, 5, 8, 5))
+
+# loop through plotting them
+for (i in 1:6) {
+  fpath <- sprintf('figs/maxnet_comparison-%i.png', i)
+  img <- brick(fpath)
+  plotRGB(img)
+  mtext(LETTERS[i],
+        side = 3,
+        line = -1.5,
+        adj = 0)
+}
+
 ```
 
-So we ran it again with disc-based spatial stratification:
+```{r maxnet_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Prediction of the distribution of the Carolina wren from 5 different models, produced by the workflow `maxnet\\_comparison'.
+A) MaxEnt without threshold features.
+B) MaxNet with default settings.
+C) MaxNet without regularisation.
+D) MaxNet with only linear features.
+E) MaxNet without regularisation and only linear features.
+F) Logistic regression.
+MaxNet with full features but no regularisation (C) gave the most local complexity, indicative of overfitting to the data.
+The models with only linear features (D-F) had a broader distribution, indicative of underfitting.
+Differences between MaxNet with linear features and no  regularisation (E) and logistic regression (F) are due to the simple downweighting applied to data in the former."}
+knitr::include_graphics("figs/maxnet_plotting-1.png")
+```
 
-```{r merowChangeWorkflow, dpi = 600, eval = FALSE}
-Merow_Silander_Spatial <- ChangeWorkflow(
-    merow_and_silander,
-    process = Chain(..., PartitionDisc))
+```{r maxent_aucs, eval = TRUE, echo = FALSE}
+# AUCs to be embedded in the text below
+perflist <- maxnet_comparison$report[seq(2, 12, by = 2)]
+AUCs <- sapply(perflist, function(x) x$auc)
+names(AUCs) <- c('MaxEnt',
+                 'MaxNet',
+                 'MaxNet (no regularization)',
+                 'MaxNet (linear features)',
+                 'MaxNet (linear features and no regularization)',
+                 'Logistic regression')
+AUCs <- round(AUCs, digits = 4)
 ```
+
+The AUC statistics calculated against the presence-absence data were:
+```{r results = 'asis', echo = FALSE}
+cat(paste(names(AUCs), AUCs, collapse = '; '))
+cat('.')
+```
+
+Readers interested in a more comprehensive comparison, or in exploring the response curves can easily reproduce and modify the analysis either by editing and re-running the above code block or by downloading the workflow object and using the `ChangeWorkflow` function.
 

### 2016-12-23 ms.Rmd
adding Dave Gav & closing the discussion on Ex 3

@@
 title: "The zoon R package for reproducible and shareable species distribution modelling"
-author: "Nick Golding, Tom August, Tim C.D. Lucas, Emiel Van Loon & Greg McInerny"
+author: "Nick Golding, Tom August, Tim C.D. Lucas, David K. Gavaghan, Emiel Van Loon & Greg McInerny"
 bibliography: zoon_app.bib
@@
 
-zoon enables us to investigate this hypothesis by fitting MaxEnt models with and without regularisation and feature construction.
+zoon enables us to investigate this hypothesis by easily comparing MaxEnt models fitted with and without regularisation and feature construction.
 We fit MaxEnt with both the usual java implementation as well as the downweighted logistic regression approximation provided in the `maxnet` R package [@maxnet].
@@
                  'MaxNet',
-                 'MaxNet (no regularization)',
-                 'MaxNet (linear features)',
-                 'MaxNet (linear features and no regularization)',
+                 'MaxNet with no regularisation',
+                 'MaxNet with only linear features',
+                 'MaxNet with only linear features and no regularisation',
                  'Logistic regression')
@@
 ```
+As expected, MaxEnt and MaxNet (the logistic regression approximation) generated similar predictions and had similar predictice performance.
+Likewise, once the regularisation and features of MaxNet were switched off, both the predictions and performance were very similar to Logistic regression.
 

### 2016-12-24 ms.Rmd
sorted out margins in maxnet plot

@@
 
-```{r plotworkflow, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 9, dpi = 300}
+```{r plotworkflow, eval = FALSE, echo = TRUE}
 mosquito1 <- workflow(occurrence = UKAnophelesPlumbeus,
@@
 
-@Aiello-Lammens2015 proposed an approach for dealing with spatial sampling bias in presence-only data, by 'thinning' the presence data, and provide an R package `spThin` to implement their procedure [@spThin].
+@Aiello-Lammens2015 proposed an approach for dealing with spatial sampling bias in presence-only data by `thinning' the presence data, and provide an R package `spThin` to implement their procedure [@spThin].
 We can incorporate this approach in a workflow, by defining a simple *process* module that adapts the zoon data into the spThin format, uses this package to apply the algorithm, and converts data back into zoon's expected format again:
@@
 The following workflow fits these models to a previously published presence-background data on the Caroline wren in the USA [@Royle2012; @maxlike], maps their predictions (Figure 3) and evaluates the predictive performance of each model against the full presence-absence dataset by AUC.
+Note that setting the regularisation constant in the `maxnet` package to 0 caused numerical errors, so we use a very small value instead.
 
-```{r maxnet_comparison, dpi = 600, fig.show = "hide", eval = TRUE, fig.height = 7, fig.width = 9}
+```{r maxnet_comparison, dpi = 600, fig.show = "hide", eval = TRUE, out.height = 600, out.width = 600}
 maxnet_comparison <- workflow(
@@
 ```
+```{r maxnet_buffer, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", out.height = 7, out.width = 7}
 
-```{r maxnet_plotting, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 6, fig.width = 9}
+# loop through adding a margin to each of the previous figures
+for (i in 1:6) {
+  fpath_in <- sprintf('figs/maxnet_comparison-%i.png', i)
+  fpath_out <- sprintf('figs/maxnet_buffer-%i.png', i)
+  img <- brick(fpath_in)
+  ext <- as.vector(extent(img))
+
+  png(fpath_out,
+      width = ncol(img),
+      height = nrow(img))
+
+  par(oma = rep(0, 4), mar = rep(0, 4))
+  plot.new()
+  plot.window(xlim = ext[1:2],
+            ylim = ext[3:4],
+            asp = 1)
+  plotRGB(img, maxpixels = Inf, add = TRUE)
+
+  dev.off()
+}
+```
+
+```{r maxnet_plotting, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 7, fig.width = 9}
 
 # plot the figure to disk, then plot in a separate chunk with a legend
-par(mfrow = c(3, 2),
-    mar = c(6, 5, 8, 5))
+par(mfrow = c(3, 2))
 
@@
 for (i in 1:6) {
-  fpath <- sprintf('figs/maxnet_comparison-%i.png', i)
+  fpath <- sprintf('figs/maxnet_buffer-%i.png', i)
   img <- brick(fpath)
-  plotRGB(img)
+  plotRGB(img, maxpixels = Inf)
   mtext(LETTERS[i],
@@
 At present, studies such as these might be considered as ‘closed knowledge’ even if they were reproducible from the publication.
-In order to facilitate a user –friendly approach to distributed working, future work will include the development of an online interface for accessing a repository of contributed modules and modifying workflows.
+In order to facilitate a user–friendly approach to distributed working, future work will include the development of an online interface for accessing a repository of contributed modules and modifying workflows.
 This interface might also host methodological discourse and openly evaluate proposed best practices in SDM.

### 2016-12-24 ms.Rmd
reworking general package info

@@
 
-## Lists, Chains and Replicates
+## Chains, Lists and Replicates
 
@@
 
-```{r workflow2, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 9, dpi = 300}
+```{r workflow2, eval = FALSE, echo = TRUE}
 mosquito2 <- workflow(occurrence = UKAnophelesPlumbeus,
@@
 
-## Inspecting, sharing and extending a zoon workflow
+### Sharing workflows
 
-A `zoonWorkflow` object
-Structure of workflow objects:
+The `zoonWorkflow` object returned by `workflow` contains all of the data, code and results used in each stage of the analysis.
+The object is therefore a self-contained representation of an entire analysis which can be saved as a binary `RData` object, archived, shared with colleagues or uploaded to the web.
+Anyone else may then load the object into their R session, load the zoon R package, and investigate the analysis.
 
-* code (call and modules used)
-* output of each module; data, results and intermediate steps
-* recording the session info and package and module versions
+zoon provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform figshare.
+`ZoonFigshare` takes a workflow object and some minimal metadata and uploads the workflow as an RData object, along with a metadata text file, to the user's figshare profile for others to download, inspect and modify.
 
-zoon workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.
-A zoon workflow returns an object of class `zoonWorkflow`.
-This object contains all the information needed to reproduce the analysis.
-This includes the original call, a record of the packages and modules used and their versions when the analysis was run as well as all the data and fitted models.
+### Exploring workflows
 
-A workflow is reproducible, reusable and alterable by way of the functions `RerunWorkflow` and `ChangeWorkflow`.
-Using `RerunWorkflow` a workflow object can be run from scratch or from part way though the analysis to avoid computationally slow steps.
-`ChangeWorkflow` takes as it's arguments a workflow object and new selections for any or all of the workflow steps.
-The workflow is then rerun from the first altered step.
+`zoonWorkflow` objects are simply R lists containing the outputs from each module in the workflow, as well as the modules used, their versions and arguments and information about the R session in which the workflow was run.
 
-Things you can do to workflows:
+The overall structure of a `zoonWorkflow` object can be inspected and visualised using the provided `print` and `plot` methods.
+The outputs and intermediate steps of the analysis can be extracted directly from the `zoonWorkflow` object, or by using the utility functions `Occurrence`, `Covariate`, `Process`, `Model` and `Output` to extract the outputs of each analytical stage.
+These functions return either the single R object outputted by each module, or list of objects when a list of modules was used in the workflow.
 
-* visualise the structure
-* execute whole thing from scratch (grabs new data from web)
-* execute from part way through
+### Reproducing and extending workflows
 
-zoon provides functionality to quickly update an existing workflow object (even one created by someone else) by switching out modules.
+Given a `zoonWorkflow` object, the analysis can be repeated in its entirety using the zoon function `RerunWorkflow`.
+This function can therefore be used to rapidly update an analysis whenever the underlying occurrence or covariate data aree updated.
+`RerunWorkflow` can also run from a specific stage of the workflow, for example using the previously downloaded data stored in the object but re-running an output module to generate a plot.
 
-Include a figure visualising the structure of the workflow object and how it can be adopted in the ChangeWorkflow function (corresponding to an example below?).
+Workflows can similarly be modified and re-run by replacing or adding modules in one or more of the five analytical steps using the function `ChangeWorkflow`.
+`ChangeWorkflow` only re-runs the channged modules and those downstream; the stored outputs of earlier modules are re-used rather than re-running them.
+
+This functionality therefore enables researchers to explore and alter existing workflows, such as from a published analysis, without having to re-run computationally expensive models or re-download datasets that may have changed in the interim.
+See example 2 below for a demonstration of this functionality.
+
+### Exploring available modules
+
+zoon modules are not distributed with the zoon R packages, but are instead downloaded on-the-fly for the online module repository.
+New modules uploaded to the repository therefore become instantly available to zoon users, without having to update the R package.
+zoon can query the currently available modules of each type via the function `GetModuleList`.
+
+All zoon modules on the repository are accompanied by documentation and metadata, similarly to R's helpfiles.
+The documentation for any module on the repository can be accessed from R using the function `ModuleHelp`.
+zoon also provides a function `ZoonCitation` to provide citation information; similarly to the `citation` function provided with R for citing R packages.
+A web service for interactive exploration of available modules and their documentation will also be made available soon.
 
@@
 
-#### Additional features
-
-##### Accessing module documentation
-
-<!-- to do -->
-
-All modules, created by the zoon authors or contributed by others, have documentation explaining their methods and how they are used.
-This documentation is easily accessed in R by using the function `ModuleHelp`.
-This function fetched the documentation from the module repository.
-
-##### Cross-validation
+### Cross-validation
 
@@
 After models are fitted to these subsets of the data, they are also fitted to the entire dataset as this will provide the best models for predictions from further new data.
-
-##### Sharing workflows
-
-zoon provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform figshare.
-`ZoonFigshare` takes a workflow object and some minimal metadata and uploads the workflow as an RData object, along with a metadata text file, to the user's figshare profile for others to download, inspect and modify.
 
@@
 
-#### Example 3. What makes MaxEnt effective
+#### Example 3. Unpacking MaxEnt
 
@@
 The models with only linear features (D-F) had a broader distribution, indicative of underfitting.
-Differences between MaxNet with linear features and no  regularisation (E) and logistic regression (F) are due to the simple downweighting applied to data in the former."}
+Differences between MaxNet with linear features and no  regularisation (E) and logistic regression (F) are due to the downweighting applied to background data in the former."}
 knitr::include_graphics("figs/maxnet_plotting-1.png")

### 2016-12-24 ms.Rmd
tweaks and cross-validation section

@@
 In order to overcome these problems, the data and methods underpinning SDM research need to be made more accessible, reproducible and modifiable by the whole research community.
-This can be achieved by faciliting access to methods[open science reference?], and making it easier to share and modify fully reproducible research objects [open science reference?].
+This can be achieved if technologies enable and encourage sharing of research as fully reproducible objects [@Peng2011], in ways that suit the diversity of users involved in SDM [@Ahmed2015].
 
@@
 We describe the modular structure of zoon workflows and how they can be constructed, shared, reproduced and modified.
-We then illustrate how these concepts enable better SDM research by reproducing a published SDM analysis; converting a recently proposed method into a zoon module; and performing a methodological comparison.
+We then illustrate how these concepts enable better SDM research by reproducing a published SDM analysis; converting a recently proposed method into a zoon module; and performing a reproducible methodological comparison.
 
@@
 
-#### Building modules
+### Cross-validation and external validation
 
-Modules are simply R function definitions, with a set of required inputs and outputs.
-
-As long as a valid module function is defined in an R session, it can be used in workflows in combination with modules downloaded from the zoon module repository.
-Whilst this makes it easy to develop modules locally, the strength of zoon comes from the ability to upload modules to the online repository so that others can access them.
-
-To do this, it first is necessary to provide additional metadata including the module tile and description, module type, author name and definitions for any non-mandatory arguments.
-zoon provides the function `BuildModule` to facilitate entering this metadata and checking that it matches the module.
-
-The zoon vignette *Building modules* provides a detailed tutorial for building modules of each type.
-
-### Cross-validation
-
-<!— needs work—>
-
-`workflow` sensibly handles cross-validation.
-Process modules are used to split the data into a training and test set or into multiple folds so that each fold in turn will be held back and used as the test for the predictive ability of any fitted models.
-The same test/train splits will be used for each analysis if the multiple models or methods are being used so that model performance measures are comparable.
-Furthermore, the test/train splits are saved along with the workflow so that other researchers adding to a workflow can make fair comparisons. 
-After models are fitted to these subsets of the data, they are also fitted to the entire dataset as this will provide the best models for predictions from further new data.
+`workflow` handles cross-validation and external validation internally, storing this information in the datframe of occurrence records.
+*Occurrence* or *process* modules can be used to assign records to one or more cross-validation folds (positive integers indicating different hold-out groups) or to an external validation dataset (indicated by a zero).
+At the *model* stage, `workflow` fits a separate model for each cross-validation fold then makes and stores predictions for the hold-out datapoints, as well as fitting a model using the whole dataset.
+In all model fitting, records flagged for external validation are omitted.
+These cross-validation predictions can then be subsequently analysed by *output* modules to estimate out-of sample predictive performance.
+Predictions can be also be made from the full model to an external validation dataset and evaluated.
 
@@
 
-```{r fengworkflow, eval = TRUE, dpi = 600, fig.show = "hide", results = "hide", message = FALSE, fig.height = 9, fig.width = 9}
+```{r fengworkflow, eval = TRUE, dpi = 600, fig.show = "hide", results = "hide", message = FALSE, fig.height = 7, fig.width = 9}
 FengPapes <- 
@@
         adj = 0)
+
 ```

### 2016-12-24 ms.Rmd
adding affiliations and rejiggling Fig 2 dimensions

@@
 title: "The zoon R package for reproducible and shareable species distribution modelling"
-author: "Nick Golding, Tom August, Tim C.D. Lucas, David K. Gavaghan, Emiel Van Loon & Greg McInerny"
+author: 
+  - Nick Golding$^1$, Tom August$^2$, Tim C.D. Lucas$^3$, David K. Gavaghan$^4$, Emiel Van Loon$^5$ & Greg McInerny$^6$
+  - $^1$ School of BioSciences, University of Melbourne, Parkville VIC, Australia
+  - $^2$ [please fill in]
+  - $^3$ [please fill in]
+  - $^4$ [please fill in]
+  - $^5$ [please fill in]
+  - $^6$ [please fill in]
 bibliography: zoon_app.bib
@@
 
-```{r fengChangeWorkflow, eval = TRUE, fig.show = "hide", fig.height = 9}
+```{r fengChangeWorkflow, eval = TRUE, fig.show = "hide", fig.height = 9, fig.width = 7.5}
 FengPapesUpdate <- 

### 2016-12-24 ms.Rmd
streamlining future work section

@@
 
-Reproducibility, open-data and open-science are topics that receive ever more attention all across science.
-Journals [@McNutt2016] and funders [@EC2016] are moving towards openly shared data and code resources as a requirement, as was set out in the executive order issued by US President Barack Obama in 2013 [@Obama2013].
-These trends strongly suggest researchers must at least find ways of documenting and disseminating code and data, if not ways that enable end users.
-Numerous platforms for publishing analytical outputs exist (e.g.
-GitHub, figshare, Dryad, CRAN), but end users will still face the demands of locating, deciphering and managing different resources across different platforms.
-Then, the real challenge of open and reproducible science, that has an increased capacity for innovation, is the challenge of increasing the usability of code and data.
+Reproducibility and open science are gaining attention [@Borregaard2016], and journals, funders and governments are increasingly making them a requirement [@McNutt2016; @EC2016; @Obama2013].
+Researchers must therefore find ways of at least disseminating code and data, if not making them usable.
+Despite numerous platforms for archiving and sharing code and data, users still face a challenge locating and using resources across different platforms.
+Increasing the usability of code and data is therefore the real challenge of open and reproducible science.
 
-However, “scientific software is developed either by domain scientists who have little training in user interface (UI) or experience (UX) design, or by software engineers who build broad systems that do not suit the niche needs or practices of domain scientists” [@Thomer2016].
-Then, there is a need for environments that can address niche needs (e.g. of a modelling discipline such as SDM) and that address usability issues for diverse outputs produced by diverse scientists.
-The SDM community is large enough to warrant a discipline specific platform such as zoon.
-The Zoon Project offers a new opportunity for the SDM community to develop community modelling benchmarks, and resolve scientific and methodological questions in ways that the current culture of publishing cannot achieve.
-It may be that zoon can also be used for other modelling activities that have similar workflows, or use similar data and processing steps.
+The size and fractured nature of the SDM research community therefore warrants a discipline-specific open research platform such as zoon to resolve these problems.
+By providing a common interface for SDM and the capacity to create and share methodological experiments zoon offers a new opportunity for the SDM community to develop community modelling benchmarks, and resolve scientific and methodological questions in ways that the current culture of publishing cannot achieve.
 
-During the development of zoon we have improvised a user-centred, participatory design process that builds on our prior experiences with software development and soliciting user feedback at 3 stages.
-An ideation workshop set the initial requirements before any code was written, and two further workshops have allowed users input on an initial prototype distributed via GitHub and then a stable R package.
-This is by no means a full UX approach [@Pavelin2012], but it has directly informed the design. 
-Continued user engagement will be vital to developing a critical mass of zoon modules that can incentivise modellers to contribute to community resources that enable evidence-based decisions about best practice.
-An important step would be the publication of a large model inter-comparison within the zoon environment, such as that produced by @Elith2006 or more recently in @Beaumont2016.
-At present, studies such as these might be considered as ‘closed knowledge’ even if they were reproducible from the publication.
-In order to facilitate a user–friendly approach to distributed working, future work will include the development of an online interface for accessing a repository of contributed modules and modifying workflows.
-This interface might also host methodological discourse and openly evaluate proposed best practices in SDM.
-In sum, we hope that the zoon R package represents a step towards a more reproducible ecosystem of SDM software.
-
-To assist users a number of tutorials are provided on the GitHub repository and as vignettes distributed with zoon – including  guides on creating workflows and modules, as well as full technical details for module developers.
+zoon was both conceived and developed based on input from users.
+Continuing and expanding user engagement will be vital to forming a critical mass of ecologists creating and using zoon modules in their research.
+To assist users a number of tutorials on using zoon and developing modules are provided on the zoonproject GitHub repository and as vignettes distributed with zoon.
+We intend to supplement these with a gallery of research examples and forum whereby the community can discuss and evaluate best SDM practice in a eproducible way.
 

### 2016-12-24 ms.Rmd
remove duplicate Royle citation & tidy up EC citation

@@
 We also compare these models with a standard logistic regression model which does not apply the downweighting step.
-The following workflow fits these models to a previously published presence-background data on the Caroline wren in the USA [@Royle2012; @maxlike], maps their predictions (Figure 3) and evaluates the predictive performance of each model against the full presence-absence dataset by AUC.
-Note that setting the regularisation constant in the `maxnet` package to 0 caused numerical errors, so we use a very small value instead.
+The following workflow fits these models to a previously published presence-background data on the Caroline wren in the USA [ @maxlike], maps their predictions (Figure 3) and evaluates the predictive performance of each model against the full presence-absence dataset by AUC.
+Note that setting the regularisation constant in the `maxnet` package to zero caused numerical errors, so we use a very small value instead.
 

### 2017-01-04 ms.Rmd
Add my affiliation
Sorry, promise I'll have a better look asap. Ddon't let me hold back submission etc.

@@
   - $^2$ [please fill in]
-  - $^3$ [please fill in]
+  - $^3$ Oxford Big Data Institute, Li Ka Shing Centre for Health Information and Discovery Nuffield Department of Medicine, University of Oxford, Oxford, U.K.
   - $^4$ [please fill in]

### 2017-01-16 ms.Rmd
Language edits and typos.
* Some typos

 * version 0.5-3 to version 0.5.3 (think this is correct).

*SDM to SDMs in certain cases. I'm not sure if this is correct, but I find it offputting when Models is abbreviated just to M. I have left cases where the word is "Modelling" to be SDM.

* Typo in Fig 1A that I can't fix. "usingh local resources or API". Also "Model(s)" is the only case with bracketed 's'. "processes", "Predictor variables", "Ouputs" are equally singular or plural. Think "Model(s)" should be changed to "Models".

* Attempt to clarify RerunWorkflow.


* "White areas are masked du

@@
 
-5. We explain how zoon works and demonstrate how it can be used to construct a complete reproducible SDM analyses, create and share a new module, and perform a methodological comparison study.
+5. We explain how zoon works and demonstrate how it can be used to construct a completely reproducible SDM analyses, create and share a new module, and perform a methodological comparison study.
 
@@
 
-For example, a fundamental dispute over the ability of SDM to detect environmental associations [@Beale2008, @Araujo2009, @Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].
+For example, a fundamental dispute over the ability of SDMs to detect environmental associations [@Beale2008, @Araujo2009, @Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].
 Similarly, in the past decade the SDM community has failed to carry out any large scale comparisons of SDM methods, still relying on the work of @Elith2006 to make decisions, despite the development of hundreds of new methods as well as improved evaluation procedures [@Roberts2016].
@@
 
-The zoon R package has been developed specifically to improve reproducibility and comparability of SDM in R by allowing users to encode entire SDM analyses as reproducible workflows of independently executable, community-contributed modules.
+The zoon R package has been developed specifically to improve reproducibility and comparability of SDMs in R by allowing users to encode entire SDM analyses as reproducible workflows of independently executable, community-contributed modules.
 The module-workflow structure enables scientists to more easily create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure 1).
@@
 
-This paper introduces version 0.5-3 of the zoon R package.
+This paper introduces version 0.5.3 of the zoon R package.
 We describe the modular structure of zoon workflows and how they can be constructed, shared, reproduced and modified.
@@
 Chains can also be applied to occurrence or covariate modules, to combine multiple datasets (e.g. occurrences from different databases) into one.
-Chaining output modules simply runs each module separately, allowing the user to create multiple maps and summary figures, calcualte performance metrics and create other model outputs in one workflow.
+Chaining output modules simply runs each module separately, allowing the user to create multiple maps and summary figures, calculate performance metrics and create other model outputs in one workflow.
 Model modules are the only module type that may not be chained, since their outputs and inputs are always different.
@@
 Some steps in SDM analyses are stochastic and need to be run multiple times; `Replicate` enables this by generating a list with one module repeated a given number of times.
-`Replicate` could be used to run the same workflow for hundreds of simulated occurrence or covariate datasets, to generate multiple bootstraps for modelling, or to fit models with stochastic elements
+`Replicate` could be used to run the same workflow for hundreds of simulated occurrence or covariate datasets, to generate multiple bootstraps for modelling, or to fit models with stochastic elements.
 
@@
 Given a `zoonWorkflow` object, the analysis can be repeated in its entirety using the zoon function `RerunWorkflow`.
-This function can therefore be used to rapidly update an analysis whenever the underlying occurrence or covariate data aree updated.
-`RerunWorkflow` can also run from a specific stage of the workflow, for example using the previously downloaded data stored in the object but re-running an output module to generate a plot.
+This function can therefore be used to rapidly update an analysis whenever the underlying occurrence or covariate data (from an online repository for example) are updated.
+Used in this way the function is equivalent to copying the authors original source code.
+`RerunWorkflow` can also run from a specific stage of the workflow, for example using the previously downloaded data stored in the object but re-running an output module to recreate a plot.
 
 Workflows can similarly be modified and re-run by replacing or adding modules in one or more of the five analytical steps using the function `ChangeWorkflow`.
-`ChangeWorkflow` only re-runs the channged modules and those downstream; the stored outputs of earlier modules are re-used rather than re-running them.
+`ChangeWorkflow` only re-runs the changed modules and those downstream; the stored outputs of earlier modules are re-used rather than re-running them.
 
 This functionality therefore enables researchers to explore and alter existing workflows, such as from a published analysis, without having to re-run computationally expensive models or re-download datasets that may have changed in the interim.
-See example 2 below for a demonstration of this functionality.
+See Example 2 below for a demonstration of this functionality.
 
@@
 
-zoon modules are not distributed with the zoon R packages, but are instead downloaded on-the-fly for the online module repository.
+zoon modules are not distributed with the zoon R packages, but are instead downloaded on-the-fly from the online module repository.
 New modules uploaded to the repository therefore become instantly available to zoon users, without having to update the R package.
@@
 The documentation for any module on the repository can be accessed from R using the function `ModuleHelp`.
-zoon also provides a function `ZoonCitation` to provide citation information; similarly to the `citation` function provided with R for citing R packages.
+zoon also provides a function `ZoonCitation` to provide citation information, similarly to the `citation` function provided with R for citing R packages.
 A web service for interactive exploration of available modules and their documentation will also be made available soon.
@@
 
-@Aiello-Lammens2015 proposed an approach for dealing with spatial sampling bias in presence-only data by `thinning' the presence data, and provide an R package `spThin` to implement their procedure [@spThin].
+@Aiello-Lammens2015 proposed an approach for dealing with spatial sampling bias in presence-only data by 'thinning' the presence data, and provide an R package `spThin` to implement their procedure [@spThin].
 We can incorporate this approach in a workflow, by defining a simple *process* module that adapts the zoon data into the spThin format, uses this package to apply the algorithm, and converts data back into zoon's expected format again:
@@
 To assist users a number of tutorials on using zoon and developing modules are provided on the zoonproject GitHub repository and as vignettes distributed with zoon.
-We intend to supplement these with a gallery of research examples and forum whereby the community can discuss and evaluate best SDM practice in a eproducible way.
+We intend to supplement these with a gallery of research examples and a forum where the community can discuss and evaluate best SDM practice in a reproducible way.
 

### 2017-01-17 ms.Rmd
Change text and legends to more accurately reflect feng analysis. * No raw data in PrintMap (for clarity I assume). *MESS masking only in interactive map. * We do not only us USA data. * Clarify what is thresholded.

@@
 
-This workflow plots a static map of the predicted distribution, and the underlying data.
+This workflow plots a static map of the predicted distribution.
 This map is shown in Figure 2A and corresponds to figure 3 in @Feng2015.
@@
 ```{r feng_papes_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Outputs of the workflow objects `FengPapes' and `FengPapesUpdate'.
-A) Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap' module in the workflow `FengPapes' which encodes a whole published analysis.
-White areas are masked due to being in the sea or outside the MESS mask.
+A) Map of the MaxEnt predicted distribution thresholded at 5% by the `PrintMap' module in the workflow `FengPapes' which encodes a whole published analysis.
 B) A response curve produced by the `ResponseCurve' module for the first covariate, bio1 in the workflow `FengPapesUpdate', which modifies the original analysis workflow.
 C) A screenshot of the interactive map produced by the `InteractiveMap' modules in the workflow `FengPapesUpdate', displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results.
+White areas are masked due to being in the MESS mask.
 Any SDM analysis distributed as a zoon workflow can be easily be explored and scrutinized by modifying its output modules using the function `ChangeWorkflow'."}

### 2017-01-17 ms.Rmd
Remove more references to USA in armadillo analysis.

@@
 
-#### Example 1. Modelling the potential distribution of nine-banded armadillo in the USA
+#### Example 1. Modelling the potential distribution of nine-banded armadillo
 
-@Feng2015 constructed a MaxEnt species distribution model for nine-banded armadillo in the USA using presence-only data on the species' current distribution, and the bioclim [@Hijmans2005] set of environmental correlates.
+@Feng2015 constructed a MaxEnt species distribution model for nine-banded armadillo using presence-only data on the species' current distribution, and the bioclim [@Hijmans2005] set of environmental correlates.
 This model was then used to predict areas in the Americas which may be suitable for the species to become established.

### 2017-01-17 ms.Rmd
Temper saying "whole published analysis". Feng and papes do more than this one workflow...

@@
 ```{r feng_papes_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Outputs of the workflow objects `FengPapes' and `FengPapesUpdate'.
-A) Map of the MaxEnt predicted distribution thresholded at 5% by the `PrintMap' module in the workflow `FengPapes' which encodes a whole published analysis.
+A) Map of the MaxEnt predicted distribution thresholded at 5% by the `PrintMap' module in the workflow `FengPapes' which encodes the core of a published analysis.
 B) A response curve produced by the `ResponseCurve' module for the first covariate, bio1 in the workflow `FengPapesUpdate', which modifies the original analysis workflow.

### 2017-01-19 ms.Rmd

--- 2017-01-17 ms.Rmd
@@
                              threshold = 0.05,
+                             thresholdmethod = 'falsenegative',
                              xlim = c(-130, -70),
@@
 ```{r feng_papes_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Outputs of the workflow objects `FengPapes' and `FengPapesUpdate'.
-A) Map of the MaxEnt predicted distribution thresholded at 5% by the `PrintMap' module in the workflow `FengPapes' which encodes a whole published analysis.
+A) Map of the MaxEnt predicted distribution, with a 5\% omissions rate threshold, by the `PrintMap' module in the workflow `FengPapes' which encodes the core of a published analysis.
 B) A response curve produced by the `ResponseCurve' module for the first covariate, bio1 in the workflow `FengPapesUpdate', which modifies the original analysis workflow.

### 2017-01-23 ms.Rmd
minor fix to figure legend & rerun figure

@@
 ```{r feng_papes_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Outputs of the workflow objects `FengPapes' and `FengPapesUpdate'.
-A) Map of the MaxEnt predicted distribution, with a 5\% omissions rate threshold, by the `PrintMap' module in the workflow `FengPapes' which encodes the core of a published analysis.
+A) Map of the MaxEnt predicted distribution, with a 5\\% omission rate threshold, by the `PrintMap' module in the workflow `FengPapes' which encodes the core of a published analysis.
 B) A response curve produced by the `ResponseCurve' module for the first covariate, bio1 in the workflow `FengPapesUpdate', which modifies the original analysis workflow.

### 2017-01-23 ms.Rmd
making the last of Tom's text edits

@@
 author: 
-  - Nick Golding$^1$, Tom August$^2$, Tim C.D. Lucas$^3$, David K. Gavaghan$^4$, Emiel Van Loon$^5$ & Greg McInerny$^6$
+  - Nick Golding$^1$, Tom A. August$^2$, Tim C.D. Lucas$^3$, David K. Gavaghan$^4$, Emiel Van Loon$^5$ & Greg McInerny$^6$
   - $^1$ School of BioSciences, University of Melbourne, Parkville VIC, Australia
-  - $^2$ [please fill in]
+  - $^2$ NERC Centre for Ecology & Hydrology, Crowmarsh Gifford, Maclean Building, Wallingford, OX10 8BB, UK
   - $^3$ Oxford Big Data Institute, Li Ka Shing Centre for Health Information and Discovery Nuffield Department of Medicine, University of Oxford, Oxford, U.K.
@@
 As a result, the SDM community is divided into research silos, with groups of researchers collaborating primarily with others who use the same software, even within specific R packages [@Ahmed2015], presenting a barrier to disseminating new SDM findings and methods.
-This is compounded by the almost all available SDM software being focussed on analytical tasks such as constructing models, rather than enabling scientists to produce analyses that others can reproduce and modify.
-The inability of SDM researchers to reproduce, scrutinise, and build on others' research prevents rigorous peer review and synthesis of research findings across studies, as well as its capacity to be a self-correcting process [@Boulton2012].
+This is compounded since almost all available SDM software are focussed on analytical tasks such as constructing models, rather than enabling scientists to produce analyses that others can reproduce and modify.
+The inability of SDM researchers to reproduce, scrutinise, and build on others' research prevents rigorous peer review, synthesis of research findings across studies, and reduces its capacity to be a self-correcting process [@Boulton2012].
 
-For example, a fundamental dispute over the ability of SDMs to detect environmental associations [@Beale2008, @Araujo2009, @Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].
+For example, a fundamental dispute over the ability of SDMs to detect environmental associations [@Beale2008; @Araujo2009; @Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].
 Similarly, in the past decade the SDM community has failed to carry out any large scale comparisons of SDM methods, still relying on the work of @Elith2006 to make decisions, despite the development of hundreds of new methods as well as improved evaluation procedures [@Roberts2016].

### 2017-01-23 ms.Rmd
adding Dave Gavaghan's text edits

@@
 author: 
-  - Nick Golding$^1$, Tom A. August$^2$, Tim C.D. Lucas$^3$, David K. Gavaghan$^4$, Emiel Van Loon$^5$ & Greg McInerny$^6$
+  - Nick Golding$^1$, Tom A. August$^2$, Tim C.D. Lucas$^3$, David J. Gavaghan$^4$, Emiel Van Loon$^5$ & Greg McInerny$^6$
   - $^1$ School of BioSciences, University of Melbourne, Parkville VIC, Australia
@@
   - $^3$ Oxford Big Data Institute, Li Ka Shing Centre for Health Information and Discovery Nuffield Department of Medicine, University of Oxford, Oxford, U.K.
-  - $^4$ [please fill in]
+  - $^4$ Department of Computer Science, University of Oxford, Oxford, UK
   - $^5$ [please fill in]
@@
 For example, a fundamental dispute over the ability of SDMs to detect environmental associations [@Beale2008; @Araujo2009; @Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].
-Similarly, in the past decade the SDM community has failed to carry out any large scale comparisons of SDM methods, still relying on the work of @Elith2006 to make decisions, despite the development of hundreds of new methods as well as improved evaluation procedures [@Roberts2016].
+Similarly, in the past decade the SDM community has failed to carry out any large scale comparisons of SDM methods, still relying on the work of @Elith2006 to make decisions, despite the development of hundreds of new methods as well as improved evaluation procedures since the publication of that study [@Roberts2016].
 The community's inability to repeat this comprehensive analysis is due to both lack of access to comparable data and the difficulty in learning and applying the many different pieces of software for model fitting.

### 2017-01-24 ms.Rmd
incorporating Tom's more major comments, including updating Figure 1

@@
 It may also have contributed to widespread misunderstandings about how apply and interpret different modelling approaches [@Yackulic2013].
-In order to overcome these problems, the data and methods underpinning SDM research need to be made more accessible, reproducible and modifiable by the whole research community.
+In order to overcome these problems, the data and code underpinning SDM research need to be made more accessible, reproducible and modifiable by the whole research community.
 This can be achieved if technologies enable and encourage sharing of research as fully reproducible objects [@Peng2011], in ways that suit the diversity of users involved in SDM [@Ahmed2015].
@@
 
-This paper introduces version 0.5.3 of the zoon R package.
+This paper introduces version 0.6 of the zoon R package.
 We describe the modular structure of zoon workflows and how they can be constructed, shared, reproduced and modified.
@@
 A) Description of the five module types.
-B) Flow diagram illustrating how objects are passed between different module types: `data frame' - an occurrence dataframe; `ras' - a RasterStack object; `mod' - a ZoonModel object; `?' - any user-defined inputs or outputs.
+B) Flow diagram illustrating how objects are passed between different module types: `data frame' - an dataframe of occurrence records; `raster' - a RasterStack object of the covariates; `model' - a ZoonModel object, generating standardised predictions from a given model.
 C) The flow diagram implied by chaining two `process' modules.
@@
 zoon enables us to investigate this hypothesis by easily comparing MaxEnt models fitted with and without regularisation and feature construction.
-We fit MaxEnt with both the usual java implementation as well as the downweighted logistic regression approximation provided in the `maxnet` R package [@maxnet].
+We fit MaxEnt with both the widely used java implementation (omitting threshold features) as well as the downweighted logistic regression approximation provided in the `maxnet` R package [@maxnet].
 We also compare these models with a standard logistic regression model which does not apply the downweighting step.
-The following workflow fits these models to a previously published presence-background data on the Caroline wren in the USA [ @maxlike], maps their predictions (Figure 3) and evaluates the predictive performance of each model against the full presence-absence dataset by AUC.
-Note that setting the regularisation constant in the `maxnet` package to zero caused numerical errors, so we use a very small value instead.
+The following workflow fits these models to a previously published presence-background data on the Caroline wren in the USA [@maxlike], maps their predictions (Figure 3) and evaluates the predictive performance of each model against the full presence-absence dataset by AUC.
+Note that setting the regularisation constant in the `maxnet` package to zero caused numerical errors, so we use a small value instead.
 
@@
 
+### Supporting Information
+
+Version 0.6 of the zoon R package is archived at https://doi.org/10.5281/zenodo.240926.
+The code required to reproduce this manuscript can be downloaded at https://github.com/zoonproject/zoon_app_paper (to be replaced with zenodo archive link after proofing stage).
+
 ### References

### 2017-01-24 ms.Rmd

@@
       in_header: latex_preamble.tex
+    highlight: null
 ---

### 2017-01-27 ms.Rmd

@@
                              ylim = c(20, 50)))
-
 ```

### 2017-01-31 ms.Rmd
enlarging effect plot labels and making landscape

@@
 This workflow plots a static map of the predicted distribution.
-This map is shown in Figure 2A and corresponds to figure 3 in @Feng2015.
+This map is shown in Figure 2A and corresponds to Figure 3 in @Feng2015.
 The resulting workflow contains all the code required to re-run the workflow, the input data and the results of executing each module.
@@
 
-```{r fengChangeWorkflow, eval = TRUE, fig.show = "hide", fig.height = 9, fig.width = 7.5}
+```{r fengChangeWorkflow, eval = TRUE, fig.show = "hide", fig.height = 9, fig.width = 7.5, dev.args = list(pointsize = 16)}
 FengPapesUpdate <- 
@@
 
+\newpage
+\blandscape
 ```{r feng_papes_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Outputs of the workflow objects `FengPapes' and `FengPapesUpdate'.
@@
 ```
+\elandscape
+\newpage
 

### 2017-01-31 ms.Rmd
Greg & Emiel's minor edits

@@
 author: 
-  - Nick Golding$^1$, Tom A. August$^2$, Tim C.D. Lucas$^3$, David J. Gavaghan$^4$, Emiel Van Loon$^5$ & Greg McInerny$^6$
+  - Nick Golding$^1$, Tom A. August$^2$, Tim C.D. Lucas$^3$, David J. Gavaghan$^4$, E. Emiel van Loon$^5$ & Greg McInerny$^6$
   - $^1$ School of BioSciences, University of Melbourne, Parkville VIC, Australia
   - $^2$ NERC Centre for Ecology & Hydrology, Crowmarsh Gifford, Maclean Building, Wallingford, OX10 8BB, UK
-  - $^3$ Oxford Big Data Institute, Li Ka Shing Centre for Health Information and Discovery Nuffield Department of Medicine, University of Oxford, Oxford, U.K.
+  - $^3$ Oxford Big Data Institute, Li Ka Shing Centre for Health Information and Discovery Nuffield Department of Medicine, University of Oxford, Oxford, UK
   - $^4$ Department of Computer Science, University of Oxford, Oxford, UK
-  - $^5$ [please fill in]
-  - $^6$ [please fill in]
+  - $^5$ Institute for Biodiversity and Ecosystem Dynamics, University of Amsterdam, Amsterdam, NL
+  - $^6$ Centre for Interdisciplinary Methodologies, Social Sciences, University of Warwick, Coventry, CV4 7AL, UK
 bibliography: zoon_app.bib
@@
 The SDM software market is similarly large and diverse [@Ahmed2015].
-Whilst most SDM users rely either on the MaxEnt standalone application [@Phillips2006] or the R programming language [@R] as the first choice software for their analyses [@Ahmed2015], there are many competing SDM-specific R packages, each implementing different approaches to fitting SDMs [@Thuiller2009; @dismo; @Naimi2016], as well as a range of alternative standalone applications.
+Whilst most SDM users rely either on the MaxEnt standalone application [@Phillips2006] or the R programming language [@R] as the first choice software for their analyses [@Ahmed2015] a variety of SDM-specific R packages are available, each implementing different approaches to fitting SDMs [@Thuiller2009; @dismo; @Naimi2016], as well as a range of alternative standalone applications.
 
-As a result, the SDM community is divided into research silos, with groups of researchers collaborating primarily with others who use the same software, even within specific R packages [@Ahmed2015], presenting a barrier to disseminating new SDM findings and methods.
-This is compounded since almost all available SDM software are focussed on analytical tasks such as constructing models, rather than enabling scientists to produce analyses that others can reproduce and modify.
-The inability of SDM researchers to reproduce, scrutinise, and build on others' research prevents rigorous peer review, synthesis of research findings across studies, and reduces its capacity to be a self-correcting process [@Boulton2012].
+As a result, the SDM community can become siloed, with groups of researchers collaborating primarily with others who use the same software, even within specific R packages [@Ahmed2015], presenting a barrier to disseminating new SDM findings and methods.
+This is compounded since almost all available SDM software is focussed on analytical tasks such as constructing models, rather than enabling scientists to produce analyses in a format that others can reproduce and modify.
+The inability of SDM researchers to reproduce, scrutinise, and build on others' research prevents rigorous peer review, synthesis of research findings across studies, and reduces capacity for the science to be a self-correcting process [@Boulton2012].
 
 For example, a fundamental dispute over the ability of SDMs to detect environmental associations [@Beale2008; @Araujo2009; @Beale2009] was left unresolved as only the original publication shared their analytical code [@Beale2008].
-Similarly, in the past decade the SDM community has failed to carry out any large scale comparisons of SDM methods, still relying on the work of @Elith2006 to make decisions, despite the development of hundreds of new methods as well as improved evaluation procedures since the publication of that study [@Roberts2016].
+Similarly, in the past decade the SDM community has failed to carry out any large scale comparisons of SDM methods, still relying on the work of @Elith2006 to select models and software [@Joppa2014], despite the development of many new methods as well as improved evaluation procedures since the publication of that study [@Roberts2016].
 The community's inability to repeat this comprehensive analysis is due to both lack of access to comparable data and the difficulty in learning and applying the many different pieces of software for model fitting.
 
-The SDM community's inability to compare models and repeat analyses has left the field unable to decisively answer even basic questions about how best to do SDM.
+Given current software and patterns of collaboration, the SDM community is unlikely to repeat or modify analyses.
+Nor is it likely to produce the large model comparisons that would answer even basic questions about how best to do SDM.
 It may also have contributed to widespread misunderstandings about how apply and interpret different modelling approaches [@Yackulic2013].
@@
 Whilst zoon's modular nature is similar to other SDM R packages such as BioMod2 [@Thuiller2009] and sdm [@Naimi2016], zoon instead pulls each module from an open repository that any SDM user can contribute to, and makes it easy for non-developers to contribute modules.
-zoon's focus on reproducibile and modifiable workflows is inspired by workflow systems such as Taverna [@Wolstencroft2013a] and the BIOVEL system [@DeGiovanni2015; @DeGiovanni2016], but embeds an SDM-specific workflow system within R, making it much more accessible to the SDM community.
+zoon's focus on reproducibile and modifiable workflows is inspired by repositories such as the Cardiac Web Lab [@Cooper2016] and workflow systems such as Taverna [@Wolstencroft2013a] and the BIOVEL system [@DeGiovanni2015; @DeGiovanni2016], but embeds an SDM-specific workflow system within R, making it much more accessible to the SDM community.
 
@@
 Lists can also be used to run the same SDM procedure for multiple species, or for multiple sets of covariates, or to compare different process modules.
-Listing output modules has the same effect as chaining them, each module is applied separately.
+Listing output modules causes each module to be applied separately.
 
@@
 ```
+
+### Cross-validation and external validation
+
+`workflow` handles cross-validation and external validation internally, storing this information in the datframe of occurrence records.
+*Occurrence* or *process* modules can be used to assign records to one or more cross-validation folds (positive integers indicating different hold-out groups) or to an external validation dataset (indicated by a zero).
+At the *model* stage, `workflow` fits a separate model for each cross-validation fold then makes and stores predictions for the hold-out datapoints, as well as fitting a model using the whole dataset.
+In all model fitting, records flagged for external validation are omitted.
+These cross-validation predictions can then be subsequently analysed by *output* modules to estimate out-of sample predictive performance.
+Predictions can be also be made from the full model to an external validation dataset and evaluated.
 
@@
 
-zoon modules are not distributed with the zoon R packages, but are instead downloaded on-the-fly from the online module repository.
+zoon modules are not distributed with the zoon R package, but are instead downloaded on-the-fly from the online module repository.
 New modules uploaded to the repository therefore become instantly available to zoon users, without having to update the R package.
@@
 A web service for interactive exploration of available modules and their documentation will also be made available soon.
-
-### Cross-validation and external validation
-
-`workflow` handles cross-validation and external validation internally, storing this information in the datframe of occurrence records.
-*Occurrence* or *process* modules can be used to assign records to one or more cross-validation folds (positive integers indicating different hold-out groups) or to an external validation dataset (indicated by a zero).
-At the *model* stage, `workflow` fits a separate model for each cross-validation fold then makes and stores predictions for the hold-out datapoints, as well as fitting a model using the whole dataset.
-In all model fitting, records flagged for external validation are omitted.
-These cross-validation predictions can then be subsequently analysed by *output* modules to estimate out-of sample predictive performance.
-Predictions can be also be made from the full model to an external validation dataset and evaluated.
 
@@
 
-```{r maxnet_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Prediction of the distribution of the Carolina wren from 5 different models, produced by the workflow `maxnet\\_comparison'.
+```{r maxnet_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Prediction of the distribution of the Carolina wren from 6 different models, produced by the workflow `maxnet\\_comparison'.
 A) MaxEnt without threshold features.
@@
 C) MaxNet without regularisation.
-D) MaxNet with only linear features.
+D) MaxNet with regularisation but only linear features.
 E) MaxNet without regularisation and only linear features.
@@
 
-Reproducibility and open science are gaining attention [@Borregaard2016], and journals, funders and governments are increasingly making them a requirement [@McNutt2016; @EC2016; @Obama2013].
+Reproducibility and open science are gaining significant attention [@Borregaard2016], and journals, funders and governments are increasingly making them a requirement [@McNutt2016; @EC2016; @Obama2013].
 Researchers must therefore find ways of at least disseminating code and data, if not making them usable.
@@
 
+### Acknowledgements
+
+The 2020 Science programme is funded through the EPSRC Cross-Disciplinary Interface Programme (grant number EP/I017909/1).
+
 ### References

### 2017-01-31 ms.Rmd
rename maxent chunks and files; save workflows

@@
 
-```{r maxnet_comparison, dpi = 600, fig.show = "hide", eval = TRUE, out.height = 600, out.width = 600}
-maxnet_comparison <- workflow(
+```{r maxent_comparison, dpi = 600, fig.show = "hide", eval = TRUE, out.height = 600, out.width = 600}
+MaxEntComparison <- workflow(
   occurrence = CarolinaWrenPO,
@@
 ```
-```{r maxnet_buffer, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", out.height = 7, out.width = 7}
+```{r maxent_buffer, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", out.height = 7, out.width = 7}
 
@@
 for (i in 1:6) {
-  fpath_in <- sprintf('figs/maxnet_comparison-%i.png', i)
-  fpath_out <- sprintf('figs/maxnet_buffer-%i.png', i)
+  fpath_in <- sprintf('figs/maxent_comparison-%i.png', i)
+  fpath_out <- sprintf('figs/maxent_buffer-%i.png', i)
   img <- brick(fpath_in)
@@
 
-```{r maxnet_plotting, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 7, fig.width = 9}
+```{r save_workflows, echo = FALSE, eval = TRUE}
+# save the three workflow objects
+save(FengPapes, FengPapesUpdate, MaxEntComparison,
+     file = 'zoon_applications_paper_workflows.RData')
+```
+
+```{r maxent_plotting, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 7, fig.width = 9}
 
@@
 for (i in 1:6) {
-  fpath <- sprintf('figs/maxnet_buffer-%i.png', i)
+  fpath <- sprintf('figs/maxent_buffer-%i.png', i)
   img <- brick(fpath)
@@
 
-```{r maxnet_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Prediction of the distribution of the Carolina wren from 6 different models, produced by the workflow `maxnet\\_comparison'.
+```{r maxent_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Prediction of the distribution of the Carolina wren from 6 different models, produced by the workflow `MaxEntComparison'.
 A) MaxEnt without threshold features.
@@
 Differences between MaxNet with linear features and no  regularisation (E) and logistic regression (F) are due to the downweighting applied to background data in the former."}
-knitr::include_graphics("figs/maxnet_plotting-1.png")
+knitr::include_graphics("figs/maxent_plotting-1.png")
 ```
@@
 # AUCs to be embedded in the text below
-perflist <- maxnet_comparison$report[seq(2, 12, by = 2)]
+perflist <- MaxEntComparison$report[seq(2, 12, by = 2)]
 AUCs <- sapply(perflist, function(x) x$auc)

### 2017-01-31 ms.Rmd
unpack reproducibility

@@
 
-The zoon R package has been developed specifically to improve reproducibility and comparability of SDMs in R by allowing users to encode entire SDM analyses as reproducible workflows of independently executable, community-contributed modules.
+The zoon R package has been developed specifically to improve reproducibility and comparability of SDMs in R by allowing users to encode entire SDM analyses as repeatable and extensible workflows consisting of independently executable, community-contributed modules.
 The module-workflow structure enables scientists to more easily create and share components of their analysis; and then access, modify, reuse and combine the components of others (see below and Figure 1).

### 2017-01-31 zoon_app_cover.Rmd
add cover letter

Dear Professor Freckleton,

I am pleased to submit the article: "The zoon R package for reproducible and shareable species distribution modelling" for consideration as an Application paper.

The Zoon Project was set up to improve reproducibility in the ecological sub discipline of species distribution modeling (SDM). This paper describes the core functionality of the project: the zoon R package, which lets ecologists construct and modify fully-reproducible SDM workflows consisting of community-contributed software modules held in an online repository.

Like the many other SDM R packages, zoon enables users to fit of SDMs with a variety of methods. However the aim and novel contribution of zoon is quite different - to facilitate the SDM community in sharing methods (modules) and experiments (workflows) as standardised, repeatable and reusable artefacts. This aligns with the journal's recent efforts to improve archiving of analytical code and has the same end goal - improving the ecological community's ability to build on each others' work. We therefore believe that our manuscript would be well suited for publication as an Application paper.

Note that two of the authors are MEE associate editors and that both Bob O'Hara and Rich FitzJohn have been involved in the development of Zoon, so one of the three remaining Applications AEs would be best placed to handle the manuscript.

Yours Sincerely,
Nick Golding, on behalf of the authors

### 2017-01-31 ms.Rmd

@@
 Next we demonstrate how zoon can be used in practice for running SDMs, creating and sharing new modules, and performing a methods comparison.
-The workflow objects created by these analyses can be accessed at [`figshare.com/articles/zoon_applications_paper_workflows`](http://figshare.com/articles/zoon_applications_paper_workflows).
+The workflow objects created by these analyses can be accessed at [`figshare.com/articles/zoon_applications_paper_workflows/4597792`](http://figshare.com/articles/zoon_applications_paper_workflows/4597792).
 We encourage readers to download, interrogate and alter these workflows for themselves.

### 2017-03-24 ms.Rmd
makes minor text edits and typos from reviews

@@
 Nor is it likely to produce the large model comparisons that would answer even basic questions about how best to do SDM.
-It may also have contributed to widespread misunderstandings about how apply and interpret different modelling approaches [@Yackulic2013].
+It may also have contributed to widespread misunderstandings about how to apply and interpret different modelling approaches [@Yackulic2013].
 In order to overcome these problems, the data and code underpinning SDM research need to be made more accessible, reproducible and modifiable by the whole research community.
@@
 Whilst zoon's modular nature is similar to other SDM R packages such as BioMod2 [@Thuiller2009] and sdm [@Naimi2016], zoon instead pulls each module from an open repository that any SDM user can contribute to, and makes it easy for non-developers to contribute modules.
-zoon's focus on reproducibile and modifiable workflows is inspired by repositories such as the Cardiac Web Lab [@Cooper2016] and workflow systems such as Taverna [@Wolstencroft2013a] and the BIOVEL system [@DeGiovanni2015; @DeGiovanni2016], but embeds an SDM-specific workflow system within R, making it much more accessible to the SDM community.
+zoon's focus on reproducible and modifiable workflows is inspired by repositories such as the Cardiac Web Lab [@Cooper2016] and workflow systems such as Taverna [@Wolstencroft2013a] and the BIOVEL system [@DeGiovanni2015; @DeGiovanni2016], but embeds an SDM-specific workflow system within R, making it much more accessible to the SDM community.
 
@@
 
-The following workflow uses a chain to standardise the covariate rasters as well as generating background records, and uses a list to fit three different models: MaxEnt, Boosted Regression Trees (GBM) and RandomForest.
-The results of the three models are then returned in three Shiny apps allowing the user to interactively explore the occurrence and covariate data, model summaries and prediction maps.
+The following workflow uses a chain to standardise the covariate rasters as well as generating background records, and uses a list to fit three different models: MaxEnt, Boosted Regression Trees (also known as Generalized Boosted Regression Models; GBM) and RandomForest.
+The results of the three models are then returned in three Shiny apps (created by the `Appify` module) allowing the user to interactively explore the occurrence and covariate data, model summaries and prediction maps.
 
@@
 
-\newpage
 \blandscape
@@
 \elandscape
-\newpage
 
@@
 We also compare these models with a standard logistic regression model which does not apply the downweighting step.
-The following workflow fits these models to a previously published presence-background data on the Caroline wren in the USA [@maxlike], maps their predictions (Figure 3) and evaluates the predictive performance of each model against the full presence-absence dataset by AUC.
+The following workflow fits these models to a previously published presence-background data on the Carolina wren in the USA [@maxlike], maps their predictions (Figure 3) and evaluates the predictive performance of each model against the full presence-absence dataset by AUC.
 Note that setting the regularisation constant in the `maxnet` package to zero caused numerical errors, so we use a small value instead.
@@
 ```
-As expected, MaxEnt and MaxNet (the logistic regression approximation) generated similar predictions and had similar predictice performance.
-Likewise, once the regularisation and features of MaxNet were switched off, both the predictions and performance were very similar to Logistic regression.
+As expected, MaxEnt and MaxNet (the logistic regression approximation) generated similar predictions and had similar predictive performance.
+Likewise, once the regularisation and features of MaxNet were switched off, both the predictions and performance were very similar to logistic regression.
 
-Readers interested in a more comprehensive comparison, or in exploring the response curves can easily reproduce and modify the analysis either by editing and re-running the above code block or by downloading the workflow object and using the `ChangeWorkflow` function.
+Readers interested in a more comprehensive comparison, or in exploring the response curves, can easily reproduce and modify the analysis either by editing and re-running the above code block or by downloading the workflow object and using the `ChangeWorkflow` function.
 

### 2017-04-10 response_to_reviewers.md
make edits in response to reviewers' comments

### Response to reviewers

Thank you very much for the rapid and detailed assessment of our manuscript. We'd like to thanks the AE and reviewers for their positive comments, enthusiasm and very helpful suggestions, all of which have been taken on board in this revision. I have detailed our responses to these suggestions below, with text additions in boldface. I have also corrected all of the typos flagged during the review process (those corrections are not detailed here).

Nick Golding, on behalf of the authors.

#### Reviewer 1 comments in email:

> "For example: add to the manuscript an attachment with the methods to build "modules"."

The zoon R package provides three vignettes, including a 4000-word vignette on how to build modules. We would rather users access this vignette via the package or the package's CRAN page than include it as a supplement, since that version can be updated along with future versions of the package (the vignette as at this release is also archived there). We have added the following sentence to the section *Example 2. Building a spatial thinning module* to refer explicitly to this vignette:

"**The zoon package vignette *Building modules* provides full details and examples of how to build modules of each type.**"

> "Finally, I want to make a mention to the platform to share data. The rights acquired by journals to publications that are not open access. This could generate some kind of legal conflict. It would be appropriate to mention something about the manuscript."

Like the reviewer, we envision that workflows representing whole analyses could be shared as supplementary material to journal articles. Even for paywalled publications, these supplementary materials are typically hosted by journals outside the paywall, so can be made freely available. Moreover, paywalled journals may hold the copyright for a manuscript, but not for the analytical results themselves, so authors are free to share analytical code and data, for example on online platforms such as figshare. We feel that discussing this second point would be out of place in this article, but we have amended the manuscript to clarify the point about supplementary information:

"**Workflow objects could therefore be provided as supplementary information to journal articles, or hosted online to meet journal requirements for public archiving of data and software.**"

##### Comments on pdf:

> P5: "I wonder, at the time of reading this sentence, can anyone contribute? Any type of filter or selection? Otherwise (eg, data cleaning) still have the same limitations."
>
> P11: "Is the data filtered in any way? If so, it should be mentioned here. Criteria for uploading the data. To homogenize it."

These are important points, also noted by Reviewer 2. The section *Exploring available modules* has now been slightly expanded to address these related issues, as follows:

"**The only pre-requisite for uploading a module to the online repository is that it passes a series of automated unit and integration tests to ensure interoperability with other modules. Similarly to the CRAN archive of R packages, zoon leaves the SDM community to determine which methods and software should be used, rather than acting as a gatekeeper of SDM methods.**

A web service for interactive exploration of available modules and their documentation **is currently being developed. This platform may include systems by which users can discuss and rate different modules, making it easier for the community to promote the best modules, flag methodological issues, and collaborate on module development.**"

> P9: "This sentence is the Appify definition, right ?. It is not clear."

The text has been edited to read:

"... in three Shiny apps **(created by the `Appify` module)** allowing the user ..."

> P10: "This function is interesting and valuable for the scientific community. However, how do you handle the rights acquired by the Journal on the results they would be sharing on this platform?"

Resolved as per second comment above.

> P10: "I just checked on the R Page, and it would be interesting to include as an attachment the basic steps to create the different modules that appear on that website."

Resolved as per first comment above.

> P12: "Is there a function to export maps in ASCII formats?"

We have added a sentence as follows:

"**Various other output modules could be used at this point, for example to validate models, project distributions to new regions, or output predictions in a variety of formats.**"

#### Reviewer 2 comments:

>  "First, there needs to be a clear definition of what a module is. For example, it wasn’t necessarily obvious that the modelling modules are not drawing directly from modelling functions already available (i.e. ‘randomForest’ vs ‘RandomForest’). Particularly, as people upload more modules with their own naming schemes, it may not be clear for less experienced users what the difference are and what the consequences of these differences could be for their analyses."

Thank you for highlighting this gap in the write up. We have added the following to the *Building a workflow* section, which introduces modules.

"Each of these steps is encoded as one or more software modules **(snippets of R code) to complete one of these tasks. zoon modules can use methods from any R package, providing a common interface between the many SDM packages already available.**"

The potential issue of multiple similarly-named modules is closely related to reviewer 1's comment about filtering and exploring modules. In our edits in response to that comment, we discuss plans for a web platform where module developers and discuss, rate, and collaborate on modules. Ultimately, we expect the SDM community will be able to promote the best modules and ensure they are used, as happens already with R packages.

>  "Secondly, there is no obvious quality control of the modules that people might draw on. Are there going to be mechanisms by which errors in code are reported? Is this going to be managed through self-policing? I think a discussion of this is required."

This has been addressed by adding text to the section *Exploring available modules*, as detailed in response to Reviewer 1. The new text explains the existing testing of code, the philosophy of the modules repository, and plans for a web platform to make community discussion of modules easier.

> "Finally, I think the authors may have missed an opportunity. An additional step in the workflow for testing if models are fit-for-purpose would be extremely beneficial i.e. testing model residuals forspatial autocorrelation, testing variables for excessive correlation, are the sample sizes large enough given the number of covariates, and are all the models in an ensemble useful. These are important considerations when fitting SDMs, and are far too often overlooked, particularly in packages such as ‘BioMod2’ and ‘sdm’. If we want to widen up the accessibility of SDMs beyond core modellers and improve their application by the less experienced, packages that aim to improve ease of use need to also offer opportunities for testing if a model has been fitted correctly. I understand that this will be a significant undertaking, but should be something that the authors consider, discuss, and potentially implement in a future version of the package."

We agree that automated checking and validation of analyses is an important and under-appreciated aspect of SDM modelling, and something that zoon could help to improve. Rather than adding an additional module type to zoon, such checks would be best handled via output modules. We as the developers of zoon do not want to be seen as dictating which methods are 'correct', but would prefer this be determined by consensus of the SDM community. We would therefore welcome the development by other parties of output modules that check whether workflows are fit for purpose. We have highlighted this opportunity in the Discussion, below an existing related paragraph:

"By providing a common interface for SDM and the capacity to create and share methodological experiments zoon offers a new opportunity for the SDM community to develop community modelling benchmarks, and resolve scientific and methodological questions in ways that the current culture of publishing cannot achieve.

**zoon also has the potential to improve methodological standards in SDM, a research area in which there is widespread misunderstanding and misuse of techniques (Yackulic *et al.* 2013). SDM methodologists could encode model validation procedures and best-practice guidance as zoon output modules, enabling end users to rapidly ensure their modelling procedures are fit-for-purpose (Guillera-Arroita *et al.* 2015).**"

> "Figure 1: What’s the difference between the grey and the black lines? Why in Figure 1C does the line come out of the bottom of the second process icon? Am I missing something?"

The colour and orientation of the arrows are stylistic and simply intended to focus the reader on the relevant part of each diagram. Black lines are used to highloght the inputs and outputs of the module type which lists and chains are being applied. In Figure 1C the line goes out of the bottom of the box for symmetry with the black line going into the box. We believe these choice help make this an engaging figure, and need not be detailed in the figure caption.

### 2017-04-10 ms.Rmd
make edits in response to reviewers' comments

@@
 zoon encodes SDM analyses as a simple workflow of five key steps: obtaining occurrence data; obtaining covariate data; applying some processes to these data; fitting one or more models; and generating any outputs (Figure 1A).
-Each of these steps is encoded as one or more software modules of a given type.
+Each of these steps is encoded as one or more software modules (snippets of R code) to complete one of these tasks.
+zoon modules can use methods from any R package, providing a common interface between the many SDM packages already available.
 Users combine these modules via a call to the `workflow` function, which executes each module in turn, before returning a `zoonWorkflow` object - a shareable, extensible and fully reproducible documentation of the SDM analysis.
@@
 
+Workflow objects could therefore be provided as supplementary information to journal articles, or hosted online to meet journal requirements for public archiving of data and software. 
 zoon provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform figshare.
@@
 zoon also provides a function `ZoonCitation` to provide citation information, similarly to the `citation` function provided with R for citing R packages.
-A web service for interactive exploration of available modules and their documentation will also be made available soon.
+
+The only pre-requisite for uploading a module to the online repository is that it passes a series of automated unit and integration tests to ensure interoperability with other modules.
+Similarly to the CRAN archive of R packages, zoon leaves the SDM community to determine which methods and software should be used, rather than acting as a gatekeeper of SDM methods.
+
+A web service for interactive exploration of available modules and their documentation is currently being developed.
+This platform may include systems by which users can rate different modules, making it easier for the community to promote the best modules, flag methodological issues, and collaborate on module development.
 
@@
 These outputs are shown in Figure 2, panels B and C.
+Various other output modules could be used here, e.g. to validate models, project distributions to new regions, or save prediction maps in a variety of formats.
 
@@
 Modules can be uploaded to the zoon modules repository via the online submission system at [URL TBD].
+The zoon package vignette *Building modules* provides full details and examples of how to build modules of each type.
 
@@
 
+zoon also has the potential to improve methodological standards in SDM, a research area in which there is widespread misunderstanding and misuse of techniques [@Yackulic2013].
+SDM methodologists could encode model validation procedures and best-practice guidance as zoon output modules, enabling end users to rapidly ensure their modelling procedures are fit for purpose [@Guillera-Arroita2015].
+
 zoon was both conceived and developed based on input from users.

### 2017-04-10 response_to_reviewers.md
fix typo in response

@@
 Thank you very much for the rapid and detailed assessment of our manuscript.
-We'd like to thanks the AE and reviewers for their positive comments, enthusiasm and very helpful suggestions, all of which have been taken on board in this revision.
+We'd like to thank the AE and reviewers for their positive comments, enthusiasm and very helpful suggestions, all of which have been taken on board in this revision.
 I have detailed our responses to these suggestions below, with text additions in boldface.
@@
 
-#### Reviewer 1 comments in email:
+### #
 
-> "For example: add to the manuscript an attachment with the methods to build "modules"."
+---
+
+### #
+
+#### Reviewer 1 comments (email & pdf):
+
+####
+
+> "For example: add to the manuscript an attachment with the methods to build 'modules'."
+>
+> P10: "I just checked on the R Page, and it would be interesting to include as an attachment the basic steps to create the different modules that appear on that website."
 
@@
 
+####
+
 > "Finally, I want to make a mention to the platform to share data.
@@
 It would be appropriate to mention something about the manuscript."
+>
+> P10: "This function is interesting and valuable for the scientific community. However, how do you handle the rights acquired by the Journal on the results they would be sharing on this platform?"
 
@@
 
-##### Comments on pdf:
+### #
 
@@
 
+### #
+
 > P9: "This sentence is the Appify definition, right ?. It is not clear."
@@
 
-> P10: "This function is interesting and valuable for the scientific community. However, how do you handle the rights acquired by the Journal on the results they would be sharing on this platform?"
-
-Resolved as per second comment above.
-
-> P10: "I just checked on the R Page, and it would be interesting to include as an attachment the basic steps to create the different modules that appear on that website."
-
-Resolved as per first comment above.
+### #
 
@@
 
+### #
+
+---
+
+### #
+
 #### Reviewer 2 comments:
+
+### #
 
@@
 
-Thank you for highlighting this gap in the write up. We have added the following to the *Building a workflow* section, which introduces modules.
+We have added the following clarifying text to the *Building a workflow* section, which introduces modules.
 
@@
 
-The potential issue of multiple similarly-named modules is closely related to reviewer 1's comment about filtering and exploring modules.
+The potential issue of multiple similarly-named modules is closely related to Reviewer 1's comments above about filtering and exploring modules.
 In our edits in response to that comment, we discuss plans for a web platform where module developers and discuss, rate, and collaborate on modules.
 Ultimately, we expect the SDM community will be able to promote the best modules and ensure they are used, as happens already with R packages.
+
+### #
 
@@
 The new text explains the existing testing of code, the philosophy of the modules repository, and plans for a web platform to make community discussion of modules easier.
+
+### #
 
@@
 **zoon also has the potential to improve methodological standards in SDM, a research area in which there is widespread misunderstanding and misuse of techniques (Yackulic *et al.* 2013).
-SDM methodologists could encode model validation procedures and best-practice guidance as zoon output modules, enabling end users to rapidly ensure their modelling procedures are fit-for-purpose (Guillera-Arroita *et al.* 2015).**"
+SDM methodologists could encode model validation procedures and best-practice guidance as zoon output modules, enabling end users to rapidly ensure their modelling procedures are fit for purpose (Guillera-Arroita *et al.* 2015).**"
+
+### #
 
@@
 
-The colour and orientation of the arrows are stylistic and simply intended to focus the reader on the relevant part of each diagram.
-Black lines are used to highloght the inputs and outputs of the module type which lists and chains are being applied.
-In Figure 1C the line goes out of the bottom of the box for symmetry with the black line going into the box.
-We believe these choice help make this an engaging figure, and need not be detailed in the figure caption.
+The colour and orientation of the arrows are stylistic and intended to focus the reader's eye on the relevant part of each diagram.
+Black lines are used to highlight the inputs and outputs of the module type which lists and chains are being applied.
+In Figure 1C the black line goes out of the bottom of the box for symmetry with the black line going into the box.
+We believe these choices help make this an engaging figure, and need not be detailed in the figure caption.

### 2017-04-10 compile.sh
add response compilation script & header

@@ -1,0 +1,3 @@
+#!/bin/bash
+pandoc -s -S -V geometry:margin=1in -H response_header.tex response_to_reviewers.md -o response_to_reviewers.pdf
+

### 2017-04-13 ms.Rmd
add author contributions section

@@
 
-### Supporting Information
+### Supporting information
 
@@
 The code required to reproduce this manuscript can be downloaded at https://github.com/zoonproject/zoon_app_paper (to be replaced with zenodo archive link after proofing stage).
+
+### Author's contributions
+All authors contributed to the design of the zoon R package.
+TAA, TCDL and NG implemented and programmed the R package with input from the other authors.
+NG wrote the first draft of the manuscript and all authors contributed to subsequent revisions and gave final approval for publication.
 

### 2017-04-13 ms.Rmd
tidying up references

@@
 This is compounded since almost all available SDM software is focussed on analytical tasks such as constructing models, rather than enabling scientists to produce analyses in a format that others can reproduce and modify.
-The inability of SDM researchers to reproduce, scrutinise, and build on others' research prevents rigorous peer review, synthesis of research findings across studies, and reduces capacity for the science to be a self-correcting process [@Boulton2012].
+The inability of SDM researchers to reproduce, scrutinise, and build on others' research prevents rigorous peer review, synthesis of research findings across studies, and reduces capacity for the science to be a self-correcting process [@Roysoc2012].
 
@@
 Whilst zoon's modular nature is similar to other SDM R packages such as BioMod2 [@Thuiller2009] and sdm [@Naimi2016], zoon instead pulls each module from an open repository that any SDM user can contribute to, and makes it easy for non-developers to contribute modules.
-zoon's focus on reproducible and modifiable workflows is inspired by repositories such as the Cardiac Web Lab [@Cooper2016] and workflow systems such as Taverna [@Wolstencroft2013a] and the BIOVEL system [@DeGiovanni2015; @DeGiovanni2016], but embeds an SDM-specific workflow system within R, making it much more accessible to the SDM community.
+zoon's focus on reproducible and modifiable workflows is inspired by repositories such as the Cardiac Web Lab [@Cooper2016] and workflow systems such as Taverna [@Wolstencroft2013a] and the BIOVEL system [@DeGiovanni2016], but embeds an SDM-specific workflow system within R, making it much more accessible to the SDM community.
 

### 2017-05-16 ms.Rmd
add data accessibility section
switch to word doc

@@
 output:
-  pdf_document:
-    fig_caption: yes
-    includes:
-      in_header: latex_preamble.tex
+  word_document:
+    fig_caption: true
     highlight: null
@@
 Figure 1B illustrates how data and outputs are passed between modules of the five different types in a a zoon workflow.
-
-```{r flow_diagram, echo = FALSE, out.width = "375px", fig.cap = "The modular SDM structure encoded by a zoon workflow.
-A) Description of the five module types.
-B) Flow diagram illustrating how objects are passed between different module types: `data frame' - an dataframe of occurrence records; `raster' - a RasterStack object of the covariates; `model' - a ZoonModel object, generating standardised predictions from a given model.
-C) The flow diagram implied by chaining two `process' modules.
-D) The flow diagram implied by listing three `model' modules.
-Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the zoon vignette `Building a module'."}
-knitr::include_graphics("figs/diagrams.png")
-```
 
@@
 
-\blandscape
-```{r feng_papes_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Outputs of the workflow objects `FengPapes' and `FengPapesUpdate'.
-A) Map of the MaxEnt predicted distribution, with a 5\\% omission rate threshold, by the `PrintMap' module in the workflow `FengPapes' which encodes the core of a published analysis.
-B) A response curve produced by the `ResponseCurve' module for the first covariate, bio1 in the workflow `FengPapesUpdate', which modifies the original analysis workflow.
-C) A screenshot of the interactive map produced by the `InteractiveMap' modules in the workflow `FengPapesUpdate', displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results.
-White areas are masked due to being in the MESS mask.
-Any SDM analysis distributed as a zoon workflow can be easily be explored and scrutinized by modifying its output modules using the function `ChangeWorkflow'."}
-knitr::include_graphics("figs/combinefengupdate-1.png")
-```
-\elandscape
-
 #### Example 2. Building a spatial thinning module
@@
 
-```{r maxent_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Prediction of the distribution of the Carolina wren from 6 different models, produced by the workflow `MaxEntComparison'.
-A) MaxEnt without threshold features.
-B) MaxNet with default settings.
-C) MaxNet without regularisation.
-D) MaxNet with regularisation but only linear features.
-E) MaxNet without regularisation and only linear features.
-F) Logistic regression.
-MaxNet with full features but no regularisation (C) gave the most local complexity, indicative of overfitting to the data.
-The models with only linear features (D-F) had a broader distribution, indicative of underfitting.
-Differences between MaxNet with linear features and no  regularisation (E) and logistic regression (F) are due to the downweighting applied to background data in the former."}
-knitr::include_graphics("figs/maxent_plotting-1.png")
-```
-
 ```{r maxent_aucs, eval = TRUE, echo = FALSE}
@@
 
+### Data accessibility
+
+Nine-banded armadillo data were downloaded from GBIF, Carolina Wren data from @maxlike were accessed via the `maxlike` R package. Both datasets can be accessed using the code presented here.
+
+### Figures
+
+```{r flow_diagram, echo = FALSE, fig.cap = "Figure 1.
+The modular SDM structure encoded by a zoon workflow.
+A) Description of the five module types.
+B) Flow diagram illustrating how objects are passed between different module types: 'data frame' - an dataframe of occurrence records; 'raster' - a RasterStack object of the covariates; 'model' - a ZoonModel object, generating standardised predictions from a given model.
+C) The flow diagram implied by chaining two `process' modules.
+D) The flow diagram implied by listing three 'model' modules.
+Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the zoon vignette 'Building a module'."}
+knitr::include_graphics("figs/diagrams.png")
+```
+
+```{r feng_papes_plots, cache = FALSE, echo = FALSE, fig.cap = "Figure 2.
+Outputs of the workflow objects 'FengPapes' and 'FengPapesUpdate'.
+A) Map of the MaxEnt predicted distribution, with a 5\\% omission rate threshold, by the 'PrintMap' module in the workflow 'FengPapes' which encodes the core of a published analysis.
+B) A response curve produced by the 'ResponseCurve' module for the first covariate, bio1 in the workflow 'FengPapesUpdate', which modifies the original analysis workflow.
+C) A screenshot of the interactive map produced by the 'InteractiveMap' modules in the workflow 'FengPapesUpdate', displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results.
+White areas are masked due to being in the MESS mask.
+Any SDM analysis distributed as a zoon workflow can be easily be explored and scrutinized by modifying its output modules using the function 'ChangeWorkflow'."}
+knitr::include_graphics("figs/combinefengupdate-1.png")
+```
+
+```{r maxent_plots, cache = FALSE, echo = FALSE, fig.cap = "Figure 3.
+Prediction of the distribution of the Carolina wren from 6 different models, produced by the workflow 'MaxEntComparison'.
+A) MaxEnt without threshold features.
+B) MaxNet with default settings.
+C) MaxNet without regularisation.
+D) MaxNet with regularisation but only linear features.
+E) MaxNet without regularisation and only linear features.
+F) Logistic regression.
+MaxNet with full features but no regularisation (C) gave the most local complexity, indicative of overfitting to the data.
+The models with only linear features (D-F) had a broader distribution, indicative of underfitting.
+Differences between MaxNet with linear features and no  regularisation (E) and logistic regression (F) are due to the downweighting applied to background data in the former."}
+knitr::include_graphics("figs/maxent_plotting-1.png")
+```
+
 ### References

### 2017-07-26 ms.Rmd
add link to website
add zenodo link to ms and readme

@@
 This module can now be shared so that others can use it in their zoon workflows.
-Modules can be uploaded to the zoon modules repository via the online submission system at [URL TBD].
+Modules can be uploaded to the zoon modules repository via the online submission system at [zoonproject.org](http://zoonproject.org).
 The zoon package vignette *Building modules* provides full details and examples of how to build modules of each type.
@@
 
-Version 0.6 of the zoon R package is archived at https://doi.org/10.5281/zenodo.240926.
-The code required to reproduce this manuscript can be downloaded at https://github.com/zoonproject/zoon_app_paper (to be replaced with zenodo archive link after proofing stage).
+Version 0.6 of the zoon R package is archived at [doi.org/10.5281/zenodo.240926](https://doi.org/10.5281/zenodo.240926).
+The code required to reproduce this manuscript is archived at [doi.org/10.5281/zenodo.834870](https://doi.org/10.5281/zenodo.834870).
 
